---
title: "COVID-19 scRNA-seq & Bulk RNA-seq Analysis - All Phases Combined"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

# Combined Analysis Pipeline

This document contains **all 13 phases** of the COVID-19 PBMC analysis,
combined into a single executable notebook.

---

## Master Setup

```{r setup, message=FALSE, warning=FALSE}
library(Seurat)
library(Matrix)
library(tidyverse)
library(patchwork)
library(SingleR)
library(celldex)
library(dittoSeq)
library(pheatmap)
library(EnhancedVolcano)
library(ggrepel)
library(RColorBrewer)
library(DESeq2)
library(GEOquery)
library(org.Hs.eg.db)
library(apeglm)
library(factoextra)
library(clusterProfiler)
library(enrichplot)
library(msigdbr)
library(GSVA)

knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 12,
  fig.height = 7
)
```


---

# ====================================================================
# PHASE1 - SCRNA QC PREPROCESSING
# ====================================================================


## Goal

This document implements **Phase 1** of the project workflow using the same logic as **Tutorial 10**:

- load 10X scRNA-seq data from `/Data/`
- inspect matrix dimensions and sparsity
- create a Seurat object
- calculate mitochondrial percentage
- visualize QC metrics
- choose filtering thresholds
- filter low-quality cells

> This file is written so you can use it directly in your project and later add your own explanations/decision checkpoint text.

---

## 1. Setup

```{r p1-p1-init, message=FALSE, warning=FALSE}
PLOTS_DIR <- "plots/phase1"
dir.create(PLOTS_DIR, showWarnings = FALSE, recursive = TRUE)
```

---

## 2. User-defined parameters

Edit these values after inspecting the QC plots.

```{r p1-parameters}
# Root folder containing the dataset
DATA_ROOT <- "Data/"

# Seurat project name
PROJECT_NAME <- "COVID_PBMC"

# QC thresholds (initial defaults; adjust after inspection)
MIN_FEATURES   <- 500
MAX_FEATURES   <- 6000
MIN_COUNTS     <- 1000   # remove cells with too few UMIs (likely empty droplets)
MAX_COUNTS     <- 40000  # remove cells with too many UMIs (likely doublets)
MAX_PERCENT_MT <- 20

# Optional: minimum number of cells expressing a gene to keep the gene at object creation
MIN_CELLS_PER_GENE <- 3

# (PLOTS_DIR already defined in setup chunk above)
```

---

## 3. Locate and load the 10X data

The 10X directory for GSE149689 is set explicitly below.
A valid 10X directory must contain:

- `matrix.mtx` or `matrix.mtx.gz`
- `barcodes.tsv` or `barcodes.tsv.gz`
- `features.tsv` / `genes.tsv` (or gzipped versions)

```{r p1-detect-10x-dir}
TENX_DIR <- "Data/GSE149689"
cat("Using 10X directory:\n", TENX_DIR, "\n")
```

```{r p1-load-data}
counts <- Read10X(data.dir = TENX_DIR)

cat("Preview of the count matrix:\n")
counts[1:min(10, nrow(counts)), 1:min(3, ncol(counts))]
```

---

## 4. Inspect matrix dimensions and sparsity

```{r p1-matrix-summary}
total_entries <- as.numeric(nrow(counts)) * as.numeric(ncol(counts))
nonzero_entries <- length(counts@x)
sparsity_pct <- 100 * (1 - nonzero_entries / total_entries)
avg_genes_per_cell <- mean(Matrix::colSums(counts > 0))

cat("Matrix dimensions:\n")
cat("  Genes (rows): ", nrow(counts), "\n", sep = "")
cat("  Cells (columns): ", ncol(counts), "\n", sep = "")
cat("  Total entries: ", format(total_entries, big.mark = ","), "\n", sep = "")
cat("  Non-zero entries: ", format(nonzero_entries, big.mark = ","), "\n", sep = "")
cat("  Sparsity: ", round(sparsity_pct, 2), "%\n", sep = "")
cat("  Average genes per cell: ", round(avg_genes_per_cell, 1), "\n", sep = "")
```

---

## 5. Genes detected per cell

```{r p1-genes-per-cell}
genes_per_cell <- Matrix::colSums(counts > 0)
summary(genes_per_cell)
```

```{r p1-genes-per-cell-plot, fig.cap="Genes detected per cell (sorted, log scale)"}
plot(
  sort(genes_per_cell),
  xlab = "Cells (sorted)",
  ylab = "Genes detected",
  log = "y",
  main = "Genes per Cell Distribution",
  pch = 16,
  cex = 0.35
)

abline(h = MIN_FEATURES, col = "red", lwd = 2, lty = 2)
abline(h = MAX_FEATURES, col = "blue", lwd = 2, lty = 2)
legend(
  "topleft",
  legend = c(
    paste("MIN_FEATURES =", MIN_FEATURES),
    paste("MAX_FEATURES =", MAX_FEATURES)
  ),
  col = c("red", "blue"),
  lty = 2,
  lwd = 2,
  bty = "n"
)

# Save plot
p_genes <- recordPlot()
png(file.path(PLOTS_DIR, "phase1_step05_genes_per_cell.png"), width = 900, height = 550)
replayPlot(p_genes)
dev.off()
```

---

## 6. Create Seurat object

```{r p1-create-seurat}
seurat <- CreateSeuratObject(
  counts = counts,
  project = PROJECT_NAME,
  min.cells = MIN_CELLS_PER_GENE
)

cat("Seurat object created.\n")
cat("  Cells:", ncol(seurat), "\n")
cat("  Genes:", nrow(seurat), "\n")
```

---

## 7. Add mitochondrial QC metric

```{r p1-percent-mt}
seurat[["percent.mt"]] <- PercentageFeatureSet(seurat, pattern = "^MT-")

summary(seurat$percent.mt)
```

```{r p1-mito-distribution, fig.cap="Mitochondrial percentage per cell (sorted)"}
plot(
  sort(seurat$percent.mt),
  xlab = "Cells (sorted)",
  ylab = "% mitochondrial reads",
  main = "Mitochondrial Content Distribution",
  pch = 16,
  cex = 0.35
)

abline(h = MAX_PERCENT_MT, col = "darkgreen", lwd = 2, lty = 2)
legend(
  "topleft",
  legend = paste("MAX_PERCENT_MT =", MAX_PERCENT_MT),
  col = "darkgreen",
  lty = 2,
  lwd = 2,
  bty = "n"
)

# Save plot
p_mito <- recordPlot()
png(file.path(PLOTS_DIR, "phase1_step07_mito_distribution.png"), width = 900, height = 550)
replayPlot(p_mito)
dev.off()
```

---

## 8. Visualize QC metrics

```{r p1-violin-qc, fig.width=11, fig.height=4}
p_vln_qc <- VlnPlot(
  seurat,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
  ncol = 3,
  pt.size = 0.1
) + NoLegend()
p_vln_qc
ggsave(file.path(PLOTS_DIR, "phase1_step08_violin_qc_metrics.png"),
       plot = p_vln_qc, width = 11, height = 4, dpi = 150)
```

```{r p1-scatter-qc, fig.width=10, fig.height=4}
p1 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "percent.mt") +
  ggtitle("nCount_RNA vs percent.mt")

p2 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") +
  ggtitle("nCount_RNA vs nFeature_RNA")

p_scatter_qc <- p1 + p2
p_scatter_qc
ggsave(file.path(PLOTS_DIR, "phase1_step08_scatter_qc_metrics.png"),
       plot = p_scatter_qc, width = 10, height = 4, dpi = 150)
```

---

## 9. Summarize filtering rules before applying them

```{r p1-qc-summary-before}
qc_summary_before <- tibble(
  metric = c("cells_total",
             paste0("nFeature_RNA <= ", MIN_FEATURES),
             paste0("nFeature_RNA >= ", MAX_FEATURES),
             paste0("nCount_RNA <= ",   MIN_COUNTS),
             paste0("nCount_RNA >= ",   MAX_COUNTS),
             paste0("percent.mt >= ",   MAX_PERCENT_MT)),
  value = c(
    ncol(seurat),
    sum(seurat$nFeature_RNA <= MIN_FEATURES),
    sum(seurat$nFeature_RNA >= MAX_FEATURES),
    sum(seurat$nCount_RNA   <= MIN_COUNTS),
    sum(seurat$nCount_RNA   >= MAX_COUNTS),
    sum(seurat$percent.mt   >= MAX_PERCENT_MT)
  )
)

qc_summary_before
```

---

## 10. Filter cells

```{r p1-filter-cells}
seurat_filtered <- subset(
  seurat,
  subset = nFeature_RNA > MIN_FEATURES &
           nFeature_RNA < MAX_FEATURES &
           nCount_RNA   > MIN_COUNTS   &
           nCount_RNA   < MAX_COUNTS   &
           percent.mt   < MAX_PERCENT_MT
)

cat("Cells before filtering:", ncol(seurat), "\n")
cat("Cells after filtering:", ncol(seurat_filtered), "\n")
cat("Cells removed:", ncol(seurat) - ncol(seurat_filtered), "\n")
```

---

## 11. Compare QC metrics before vs after filtering

```{r p1-compare-violin, fig.width=11, fig.height=8}
p_before <- VlnPlot(
  seurat,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
  ncol = 3,
  pt.size = 0.1
) + plot_annotation(title = "Before filtering")

p_after <- VlnPlot(
  seurat_filtered,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
  ncol = 3,
  pt.size = 0.1
) + plot_annotation(title = "After filtering")

p_compare <- p_before / p_after
p_compare
ggsave(file.path(PLOTS_DIR, "phase1_step11_before_after_filtering.png"),
       plot = p_compare, width = 11, height = 8, dpi = 150)
```

```{r p1-compare-summary-after}
qc_summary_after <- tibble(
  metric = c("cells_remaining",
             "median_nFeature_RNA",
             "median_nCount_RNA",
             "median_percent.mt"),
  value = c(
    ncol(seurat_filtered),
    median(seurat_filtered$nFeature_RNA),
    median(seurat_filtered$nCount_RNA),
    median(seurat_filtered$percent.mt)
  )
)

qc_summary_after
```

---

## 12. Save the filtered object

```{r p1-save-object}
saveRDS(seurat_filtered, file = file.path(DATA_ROOT, "phase1_seurat_filtered.rds"))
cat("Saved filtered Seurat object to:\n", file.path(DATA_ROOT, "phase1_seurat_filtered.rds"), "\n")
```

---

## 13. Decision checkpoint 1

**Decision:** I filtered cells using thresholds on nFeature_RNA, nCount_RNA, and percent.mt.

**What I chose:**

| Threshold | Value | Reason |
|---|---|---|
| MIN_FEATURES | 500 | Sits just above the empty-droplet cliff in the genes-per-cell plot |
| MAX_FEATURES | 6,000 | Catches the small doublet spike at the far right of the distribution |
| MIN_COUNTS | 1,000 | Removes weak-capture droplets independently of gene count |
| MAX_COUNTS | 40,000 | Removes high-UMI doublets |
| MAX_PERCENT_MT | 20% | Preserves the large flat healthy region, removes the exponential damage tail |

**Outcome:**

- Cells before: 85,144
- Cells after: **55,548** (29,596 removed, 34.8%)
- Post-filtering medians: 1,872 genes/cell, 5,845 UMIs/cell, 7.3% mito

**Why these thresholds:**  
All five thresholds were chosen by visually inspecting the QC plots (steps 5, 7, 8).
The genes-per-cell distribution clearly showed a dropout cliff below 500 and a doublet spike above 6,000.
The mitochondrial distribution showed a flat healthy region up to ~15% followed by a sharp exponential rise — 20% sits at the start of that rise and is a standard cutoff in the literature.
The nCount thresholds add an independent check: 1,000 removes very weak droplets that passed the gene filter, and 40,000 removes high-UMI doublets.

**Largest sources of removed cells:**
- 22,774 cells (26.7%) — too few genes (empty droplets / ambient RNA)
- 18,282 cells (21.5%) — too high mitochondrial content (damaged cells, consistent with severe COVID samples)
- 13,959 cells (16.4%) — too few UMIs (weak capture)
- 238 cells (0.3%) — doublet candidates (very few, consistent with a well-run experiment)

**Alternatives considered:**  
Adaptive thresholds based on median absolute deviation (MAD) could be used instead of fixed values.
Dedicated doublet-detection tools (e.g. DoubletFinder) could replace the upper gene/count cutoffs.
For a final version, these would be worth applying, especially for the doublet filter.

---

## 14. Next step

After Phase 1, continue with:

- `SCTransform()`
- `RunPCA()`
- `ElbowPlot()`
- `RunUMAP()` / `RunTSNE()`
- `FindNeighbors()` and `FindClusters()`

That corresponds to the next part of Tutorial 10 and your project roadmap.

---

# ====================================================================
# PHASE2 - NORMALIZATION DIMREDUCTION
# ====================================================================


## Goal

This document implements **Phase 2** of the project workflow:

- normalize with SCTransform (regressing out mitochondrial percentage)
- run PCA and inspect loadings / heatmaps
- choose the number of PCs via elbow plot
- run UMAP and t-SNE embeddings

---

## 1. Setup

```{r p2-p2-init, message=FALSE, warning=FALSE}
dir.create(PLOTS_DIR, showWarnings = FALSE, recursive = TRUE)
```

---

## 2. Load the filtered Seurat object from Phase 1

```{r p2-load-object}
seurat <- readRDS("Data/phase1_seurat_filtered.rds")

cat("Loaded filtered Seurat object.\n")
cat("  Cells:", ncol(seurat), "\n")
cat("  Genes:", nrow(seurat), "\n")
```

---

## 3. User-defined parameters

```{r p2-parameters}
# Number of PCs to use for UMAP / t-SNE (adjust after inspecting elbow plot)
N_PCS <- 20

# Folder where all Phase 2 plots will be saved
PLOTS_DIR <- "plots/phase2"
```

---

## 4. Normalization

We use the standard LogNormalize workflow (NormalizeData → FindVariableFeatures →
ScaleData) because **both SCTransform v1 and v2** fail with an out-of-memory error
at the `sctransform::correct()` step, which allocates a dense cells × genes matrix
(~9.1 GB) — this is inherent to SCTransform's architecture and not resolved by
the `glmGamPoi` backend or `vst.flavor = "v2"`.

The LogNormalize workflow is well-validated, widely used in published scRNA-seq
studies, and produces comparable downstream results for clustering and
differential expression.

```{r p2-normalize}
seurat <- NormalizeData(seurat, normalization.method = "LogNormalize", scale.factor = 10000)
seurat <- FindVariableFeatures(seurat, selection.method = "vst", nfeatures = 3000)
seurat <- ScaleData(seurat, vars.to.regress = "percent.mt")
```

---

## 5. Run PCA

```{r p2-run-pca}
seurat <- RunPCA(seurat, verbose = FALSE)
```

### 5a. Inspect PC loadings

```{r p2-pc-loadings, fig.width=10, fig.height=8}
p_loadings <- VizDimLoadings(seurat, dims = 1:3, reduction = "pca")
p_loadings
ggsave(file.path(PLOTS_DIR, "phase2_step05a_pc_loadings.png"),
       plot = p_loadings, width = 10, height = 8, dpi = 150)
```

### 5b. PC heatmaps

```{r p2-pc-heatmaps, fig.width=10, fig.height=16}
png(file.path(PLOTS_DIR, "phase2_step05b_pc_heatmaps.png"), width = 1000, height = 1600)
DimHeatmap(seurat, dims = 1:20, cells = 500, balanced = TRUE)
dev.off()
DimHeatmap(seurat, dims = 1:20, cells = 500, balanced = TRUE)
```

---

## 6. Choose number of PCs — Elbow plot

```{r p2-elbow-plot, fig.cap="Elbow plot — choose PCs where the curve flattens"}
p_elbow <- ElbowPlot(seurat, ndims = 30) +
  geom_vline(xintercept = N_PCS, linetype = "dashed", color = "red") +
  annotate("text", x = N_PCS + 1, y = max(Stdev(seurat[["pca"]])) * 0.9,
           label = paste("N_PCS =", N_PCS), hjust = 0, color = "red")
p_elbow
ggsave(file.path(PLOTS_DIR, "phase2_step06_elbow_plot.png"),
       plot = p_elbow, width = 8, height = 5, dpi = 150)
```

---

## 7. Run UMAP

```{r p2-run-umap}
seurat <- RunUMAP(seurat, dims = 1:N_PCS, verbose = FALSE)
```

```{r p2-plot-umap, fig.cap="UMAP embedding"}
p_umap <- DimPlot(seurat, reduction = "umap") + ggtitle("UMAP")
p_umap
ggsave(file.path(PLOTS_DIR, "phase2_step07_umap.png"),
       plot = p_umap, width = 8, height = 6, dpi = 150)
```

---

## 8. Run t-SNE

```{r p2-run-tsne}
seurat <- RunTSNE(seurat, dims = 1:N_PCS, verbose = FALSE)
```

```{r p2-plot-tsne, fig.cap="t-SNE embedding"}
p_tsne <- DimPlot(seurat, reduction = "tsne") + ggtitle("t-SNE")
p_tsne
ggsave(file.path(PLOTS_DIR, "phase2_step08_tsne.png"),
       plot = p_tsne, width = 8, height = 6, dpi = 150)
```

---

## 9. Side-by-side comparison

```{r p2-umap-vs-tsne, fig.width=14, fig.height=5}
p1 <- DimPlot(seurat, reduction = "umap") + ggtitle("UMAP")
p2 <- DimPlot(seurat, reduction = "tsne") + ggtitle("t-SNE")
p_compare <- p1 + p2
p_compare
ggsave(file.path(PLOTS_DIR, "phase2_step09_umap_vs_tsne.png"),
       plot = p_compare, width = 14, height = 5, dpi = 150)
```

---

## 10. Save the processed object

```{r p2-save-object}
saveRDS(seurat, file = "Data/phase2_seurat_processed.rds")
cat("Saved processed Seurat object to: Data/phase2_seurat_processed.rds\n")
```

---

## 11. Decision Checkpoint 2

**Decision 1 — Normalization method:** LogNormalize

- **What I chose:** NormalizeData (LogNormalize, scale.factor = 10,000) → FindVariableFeatures (VST, 3,000 genes) → ScaleData (regress `percent.mt`)
- **Why not SCTransform:** Both SCTransform v1 and v2 (`vst.flavor = "v2"` with `glmGamPoi`) crash with an OOM error at `sctransform::correct()`, which attempts to allocate a ~9.1 GB dense matrix. This is an architectural limitation for datasets of this size. LogNormalize is well-established in the literature and produces concordant results for clustering and differential expression.
- **TODO:** Run SCTransform on an SSH server with more RAM for the final version.

---

**Decision 2 — Number of PCs:** `r N_PCS`

- **What I chose:** 20 PCs
- **Justification from the elbow plot:** PC1 dominates with std ≈ 14 (the monocyte signal). The curve declines gradually through PC2–10 (std 8 → 2.5), then slows to a nearly flat plateau from PC10–20. After PC20 the curve is essentially flat at std ≈ 1–1.5 — the noise floor. There is no single sharp elbow; the transition is gradual, which is typical for complex multi-celltype datasets. PC20 is the start of the plateau.
- **Justification from PC heatmaps:** PC1–3 show extremely clean two-block patterns. PC4–15 retain visible structure. PC16–20 are patchier but not pure noise. Going beyond 20 would add PCs that are predominantly noise.
- **Justification from PC loadings:** PC1 = monocyte markers (FCN1, LYZ, S100A9), PC2 = platelet markers (CLEC1B, TUBB1), PC3 = B cells vs NK/T cells (CD79A vs NKG7, CCL5). All three PCs are biologically interpretable, confirming PCA captured real structure.
- **Alternatives considered:** 15 PCs would likely lose sub-population structure visible in PC16–20; 25 PCs would add noise. N_PCS = 20 is the best-supported cutoff given the evidence from all three diagnostic plots.

---

## 12. Next step

Continue with Phase 3:

- `FindNeighbors()` and `FindClusters()`
- Cell-type annotation with manual markers and SingleR
- `FindAllMarkers()` and marker heatmap

---

# ====================================================================
# PHASE3 - CLUSTERING ANNOTATION
# ====================================================================


## Goal

This document implements **Phase 3** of the project workflow:
  
- Graph-based clustering at multiple resolutions
- Manual marker gene inspection
- Automated annotation with SingleR
- Differential marker discovery with `FindAllMarkers()`

---

## 1. Setup

```{r p3-p3-init, message=FALSE, warning=FALSE}
PLOTS_DIR <- "plots/phase3"
dir.create(PLOTS_DIR, showWarnings = FALSE, recursive = TRUE)
```

---

## 2. Load the processed Seurat object from Phase 2

```{r p3-load-object}
seurat <- readRDS("Data/phase2_seurat_processed.rds")

cat("Loaded Phase 2 Seurat object.\n")
cat("  Cells:", ncol(seurat), "\n")
cat("  Genes:", nrow(seurat), "\n")
cat("  Default assay:", DefaultAssay(seurat), "\n")
```

---

## 3. User-defined parameters

```{r p3-parameters}
# Number of PCs (must match Phase 2)
N_PCS <- 20

# Clustering resolutions to compare
RESOLUTIONS <- c(0.1, 0.3, 0.5, 0.8, 1.0)

# Final resolution to use downstream (adjust after inspection)
FINAL_RES <- 0.5
```

---

## 4. Find Neighbors

Build the shared-nearest-neighbor (SNN) graph on the PCA embedding.

```{r p3-find-neighbors}
seurat <- FindNeighbors(seurat, dims = 1:N_PCS, verbose = FALSE)
```

---

## 5. Find Clusters at multiple resolutions

```{r p3-find-clusters}
for (res in RESOLUTIONS) {
  seurat <- FindClusters(seurat, resolution = res, verbose = FALSE)
}

# Show cluster counts per resolution
res_summary <- data.frame(
  Resolution = RESOLUTIONS,
  Clusters   = sapply(RESOLUTIONS, function(r) {
    length(unique(seurat@meta.data[[paste0("RNA_snn_res.", r)]]))
  })
)
knitr::kable(res_summary, caption = "Number of clusters at each resolution")
```

---

## 6. Compare resolutions on UMAP

```{r p3-compare-resolutions, fig.width=16, fig.height=12}
plots <- lapply(RESOLUTIONS, function(r) {
  col <- paste0("RNA_snn_res.", r)
  DimPlot(seurat, group.by = col, reduction = "umap", label = TRUE, label.size = 3) +
    ggtitle(paste0("res = ", r)) +
    NoLegend()
})
p_compare_res <- wrap_plots(plots, ncol = 3)
p_compare_res
ggsave(file.path(PLOTS_DIR, "phase3_step06_resolutions_comparison.png"),
       plot = p_compare_res, width = 16, height = 12, dpi = 150)
```

---

## 7. Set final clustering resolution

```{r p3-set-final-resolution}
Idents(seurat) <- paste0("RNA_snn_res.", FINAL_RES)
cat("Using resolution:", FINAL_RES, "\n")
cat("Number of clusters:", length(levels(Idents(seurat))), "\n")
```

### 7a. UMAP by cluster (final)

```{r p3-umap-clusters, fig.width=10, fig.height=7}
p_umap_clusters <- DimPlot(seurat, reduction = "umap", label = TRUE, pt.size = 0.3) +
  ggtitle(paste0("Clusters (res = ", FINAL_RES, ")"))
p_umap_clusters
ggsave(file.path(PLOTS_DIR, "phase3_step07a_umap_clusters_final.png"),
       plot = p_umap_clusters, width = 10, height = 7, dpi = 150)
```

### 7b. Compare t-SNE vs UMAP

```{r p3-tsne-vs-umap, fig.width=16, fig.height=6}
p1 <- DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle("UMAP") + NoLegend()
p2 <- DimPlot(seurat, reduction = "tsne", label = TRUE) + ggtitle("t-SNE") + NoLegend()
p_compare_embed <- p1 + p2
p_compare_embed
ggsave(file.path(PLOTS_DIR, "phase3_step07b_umap_vs_tsne.png"),
       plot = p_compare_embed, width = 16, height = 6, dpi = 150)
```

---

## 8. Manual marker gene inspection

### 8a. Monocyte markers

```{r p3-markers-monocytes, fig.width=12, fig.height=4}
p_markers_mono <- FeaturePlot(seurat, features = c("CD14", "LYZ", "FCGR3A"),
            reduction = "umap", ncol = 3)
p_markers_mono
ggsave(file.path(PLOTS_DIR, "phase3_step08a_markers_monocytes.png"),
       plot = p_markers_mono, width = 12, height = 4, dpi = 150)
```

### 8b. T-cell markers

```{r p3-markers-tcells, fig.width=16, fig.height=4}
p_markers_t <- FeaturePlot(seurat, features = c("CD3D", "CD3E", "CD4", "CD8A"),
            reduction = "umap", ncol = 4)
p_markers_t
ggsave(file.path(PLOTS_DIR, "phase3_step08b_markers_tcells.png"),
       plot = p_markers_t, width = 16, height = 4, dpi = 150)
```
```

### 8c. B-cell markers

```{r p3-markers-bcells, fig.width=8, fig.height=4}
p_markers_b <- FeaturePlot(seurat, features = c("MS4A1", "CD79A"),
            reduction = "umap", ncol = 2)
p_markers_b
ggsave(file.path(PLOTS_DIR, "phase3_step08c_markers_bcells.png"),
       plot = p_markers_b, width = 8, height = 4, dpi = 150)
```

### 8d. NK-cell markers

```{r p3-markers-nk, fig.width=8, fig.height=4}
p_markers_nk <- FeaturePlot(seurat, features = c("NKG7", "GNLY"),
            reduction = "umap", ncol = 2)
p_markers_nk
ggsave(file.path(PLOTS_DIR, "phase3_step08d_markers_nk.png"),
       plot = p_markers_nk, width = 8, height = 4, dpi = 150)
```

### 8e. Dot plot of canonical markers

```{r p3-dotplot-markers, fig.width=12, fig.height=6}
canonical_markers <- c(
  "CD14", "LYZ", "FCGR3A",       # Monocytes
  "CD3D", "CD3E", "CD4", "CD8A", # T cells
  "MS4A1", "CD79A",               # B cells
  "NKG7", "GNLY",                 # NK cells
  "FCER1A", "CST3",               # Dendritic cells
  "PPBP"                           # Platelets / Megakaryocytes
)
p_dotplot <- DotPlot(seurat, features = canonical_markers) +
  RotatedAxis() +
  ggtitle("Canonical marker expression by cluster")
p_dotplot
ggsave(file.path(PLOTS_DIR, "phase3_step08e_dotplot_markers.png"),
       plot = p_dotplot, width = 12, height = 6, dpi = 150)
```

---

## 9. Automated annotation with SingleR

```{r p3-singler-annotation}
# Use BlueprintEncodeData as reference
singler_ref <- BlueprintEncodeData()

# Get normalized expression matrix
# (SingleR expects log-normalized counts)
expr_matrix <- GetAssayData(seurat, layer = "data")

# Run SingleR (main labels)
singler_results <- SingleR(
  test   = expr_matrix,
  ref    = singler_ref,
  labels = singler_ref$label.main
)

cat("SingleR annotation complete.\n")
cat("Label distribution:\n")
print(sort(table(singler_results$labels), decreasing = TRUE))
```

### 9a. SingleR score heatmap

```{r p3-singler-heatmap, fig.width=12, fig.height=8}
png(file.path(PLOTS_DIR, "phase3_step09a_singler_heatmap.png"), width = 1200, height = 800)
plotScoreHeatmap(singler_results)
dev.off()
plotScoreHeatmap(singler_results)
```

### 9b. Add SingleR labels to Seurat object

```{r p3-add-singler-labels}
seurat$singler_main <- singler_results$labels

# Cross-tabulate clusters vs SingleR labels
ct <- table(Cluster = Idents(seurat), CellType = seurat$singler_main)
knitr::kable(ct, caption = "Cluster × SingleR cell-type cross-tabulation")
```

### 9c. UMAP coloured by SingleR annotation

```{r p3-umap-singler, fig.width=12, fig.height=7}
p_umap_singler <- DimPlot(seurat, group.by = "singler_main", reduction = "umap",
        label = TRUE, label.size = 3, repel = TRUE) +
  ggtitle("SingleR cell-type annotation")
p_umap_singler
ggsave(file.path(PLOTS_DIR, "phase3_step09c_umap_singler.png"),
       plot = p_umap_singler, width = 12, height = 7, dpi = 150)
```

### 9d. dittoSeq UMAP

```{r p3-ditto-umap, fig.width=12, fig.height=7}
p_ditto <- dittoDimPlot(seurat, var = "singler_main", reduction.use = "umap")
ggsave(file.path(PLOTS_DIR, "phase3_step09d_ditto_umap.png"),
       plot = p_ditto, width = 12, height = 7, dpi = 150)
p_ditto
```

---

## 10. Find all markers (cluster-defining genes)

```{r p3-find-all-markers}
# FindAllMarkers on 55K cells x 23K genes exceeds the 16 GB vector memory limit
# because Seurat 5 materializes the full matrix for LFC calculation before any
# subsampling. Fix: downsample the object itself to 300 cells per cluster so
# both the test and background sides are small (300 x 23 = ~6900 total cells).
# 300 cells per cluster is sufficient to detect robust cluster-defining markers.
set.seed(42)
seurat_sub <- subset(seurat, cells = WhichCells(seurat, downsample = 300))
cat("Downsampled object:", ncol(seurat_sub), "cells\n")

all_markers <- FindAllMarkers(
  seurat_sub,
  only.pos        = TRUE,
  min.pct         = 0.25,
  logfc.threshold = 0.5,
  features        = VariableFeatures(seurat)  # limit to 3000 variable genes
)

cat("Total markers found:", nrow(all_markers), "\n")

# Show top 5 per cluster
top5 <- all_markers %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 5)

knitr::kable(top5, caption = "Top 5 markers per cluster (by log2 FC)")
```

### 10a. Marker Heatmap

```{r p3-marker-heatmap, fig.width=14, fig.height=10}
top5_genes <- top5 %>% pull(gene) %>% unique()
# Use seurat_sub (already downsampled to ~6500 cells) — DoHeatmap on the full
# 55K-cell object exceeds the 16 GB limit. The downsampled object shows the
# same expression patterns since it was drawn proportionally from all clusters.
p_heatmap <- DoHeatmap(seurat_sub, features = top5_genes, size = 3) +
  ggtitle("Top 5 markers per cluster")
png(file.path(PLOTS_DIR, "phase3_step10a_marker_heatmap.png"), width = 1400, height = 1000)
print(p_heatmap)
dev.off()
p_heatmap
```

---

## 11. Save the annotated object

```{r p3-save-object}
saveRDS(seurat, file = "Data/phase3_seurat_annotated.rds")

# Also save the markers table for later use
write.csv(all_markers, file = "Data/phase3_all_markers.csv", row.names = FALSE)

cat("Saved annotated Seurat object to: Data/phase3_seurat_annotated.rds\n")
cat("Saved marker table to: Data/phase3_all_markers.csv\n")
```

---

## 12. Decision Checkpoint 3

**Decision 1 — Clustering resolution: `r FINAL_RES`**

Five resolutions were tested: 0.1 (12 clusters), 0.3 (19), 0.5 (23), 0.8 (35), 1.0 (34).

Resolution 0.5 was chosen for three reasons:
1. **Stability test:** Cluster count *dropped* from 35 to 34 going from 0.8 to 1.0 — the algorithm is re-merging its own artificial splits. Anything ≥ 0.8 is over-split.
2. **UMAP inspection:** At res = 0.5 every visually distinct region of the UMAP has its own cluster and no continuous regions are fragmented. At res = 0.8 the T/NK right continent breaks into confetti.
3. **Marker validation:** `FindAllMarkers` returned biologically identifiable marker genes for all 23 clusters, confirming each cluster represents a real population.

**Decision 2 — FindAllMarkers on a downsampled object (300 cells/cluster)**

The full 55,548-cell object hits the 16 GB memory limit during DE testing. The object was downsampled to 300 cells per cluster (6,513 total) before running `FindAllMarkers`. This is a valid approach: 300 cells per cluster gives ample statistical power to detect highly expressed, cluster-specific markers, and the same 3,000 variable features from the full object were used.

**Population summary (from manual markers + SingleR cross-tabulation):**

- **Myeloid (right UMAP continent):** Clusters 2, 6, 7, 8, 10, 11, 12, 16, 18, 19 → monocyte sub-states (classical, non-classical, inflammasome-active, ISG-high, chemokine-secreting, proliferating)
- **Lymphoid (left UMAP continent):** Clusters 0, 1, 3, 4, 13, 20, 21 → CD8⁺ T, CD4⁺ T, NK, γδ T, Tregs
- **B cells (bottom island):** Clusters 5, 9 → naive and memory B cells
- **Dendritic cells:** Cluster 17 (FCER1A⁺)
- **Contamination:** Cluster 14 (platelets, PPBP/PTCRA), cluster 22 (erythrocytes, GYPA/GYPB)
- **Proliferating cells:** Cluster 15 (TYMS, CDC45 — S-phase genes)

**Discrepancies between manual markers and SingleR:**

- SingleR labels the entire left continent as NK cells + CD8⁺/CD4⁺ T cells, consistent with manual markers.
- Cluster 14 is labelled "Erythrocytes/HSC/Eosinophils" by SingleR — this matches the mixed/contamination interpretation from the dotplot (PPBP = platelets; GYPA-expressing cells in cluster 22 are the true erythrocytes).
- No major contradictions between the two approaches.

---

## 13. Next step

Continue with Phase 4 (CD14+ Monocyte Deep-Dive):

- Subset CD14+ monocytes
- ISG and TNF/IL-1 module scoring
- Co-activation analysis by disease severity
- Donor-level statistical tests

---

# ====================================================================
# PHASE4 - MONOCYTE DEEPDIVE
# ====================================================================


## Goal

This document implements **Phase 4** of the project workflow:

- Annotate samples with severity metadata from GEO/paper
- Subset CD14+ monocytes
- ISG and TNF/IL-1β module scoring
- Co-activation analysis (scatter, thresholds)
- Cell-level and donor-level severity comparisons
- Differential expression in monocytes + volcano plot

---

## 1. Setup

```{r p4-p4-init, message=FALSE, warning=FALSE}
dir.create("plots/phase4", showWarnings = FALSE, recursive = TRUE)
```

---

## 2. Load the annotated Seurat object from Phase 3

```{r p4-load-object}
seurat <- readRDS("Data/phase3_seurat_annotated.rds")

cat("Loaded Phase 3 Seurat object.\n")
cat("  Cells:", ncol(seurat), "\n")
cat("  Genes:", nrow(seurat), "\n")
```

---

## 3. Add sample / severity metadata

Barcode suffixes in GSE149689 encode sample identity (from GEO: Lee et al.,
Sci Immunol 2020). Severity assignment (mild vs severe COVID) is based on the
paper's Table S1.

```{r p4-add-metadata}
# Extract barcode suffix → sample mapping from GEO
suffix_to_sample <- c(
  "1"  = "nCoV_1",  "2"  = "nCoV_2",  "3"  = "Flu_1",
  "4"  = "Flu_2",   "5"  = "Normal_1","6"  = "Flu_3",
  "7"  = "Flu_4",   "8"  = "Flu_5",   "9"  = "nCoV_3",
  "10" = "nCoV_4",  "11" = "nCoV_5",  "12" = "nCoV_6",
  "13" = "Normal_2","14" = "Normal_3","15" = "nCoV_7",
  "16" = "nCoV_8",  "17" = "nCoV_9",  "18" = "nCoV_10",
  "19" = "Normal_4","20" = "nCoV_11"
)

# Condition (broad)
suffix_to_condition <- c(
  "1"  = "COVID", "2"  = "COVID", "3"  = "Flu",
  "4"  = "Flu",   "5"  = "Healthy","6"  = "Flu",
  "7"  = "Flu",   "8"  = "Flu",   "9"  = "COVID",
  "10" = "COVID", "11" = "COVID", "12" = "COVID",
  "13" = "Healthy","14"= "Healthy","15"= "COVID",
  "16" = "COVID", "17" = "COVID", "18" = "COVID",
  "19" = "Healthy","20"= "COVID"
)

# Severity (from Lee et al. 2020, Table S1)
# nCoV 1-7 = severe COVID; nCoV 8-11 = mild COVID
suffix_to_severity <- c(
  "1"  = "Severe",  "2"  = "Severe",  "3"  = "Flu",
  "4"  = "Flu",     "5"  = "Healthy", "6"  = "Flu",
  "7"  = "Flu",     "8"  = "Flu",     "9"  = "Severe",
  "10" = "Severe",  "11" = "Severe",  "12" = "Severe",
  "13" = "Healthy", "14" = "Healthy", "15" = "Severe",
  "16" = "Mild",    "17" = "Mild",    "18" = "Mild",
  "19" = "Healthy", "20" = "Mild"
)

# Assign to Seurat object
barcodes <- colnames(seurat)
suffixes <- sub(".*-", "", barcodes)

seurat$sample    <- unname(suffix_to_sample[suffixes])
seurat$condition <- unname(suffix_to_condition[suffixes])
seurat$severity  <- unname(suffix_to_severity[suffixes])
seurat$donor     <- unname(suffix_to_sample[suffixes])   # donor = sample in this dataset

# Set severity as ordered factor
seurat$severity <- factor(seurat$severity,
                          levels = c("Healthy", "Mild", "Severe", "Flu"))

cat("Metadata summary:\n")
print(table(seurat$severity))
cat("\nDonors per severity group:\n")
print(tapply(seurat$donor, seurat$severity, function(x) length(unique(x))))
```

### 3a. UMAP split by severity

```{r p4-umap-severity, fig.width=16, fig.height=5}
p <- DimPlot(seurat, group.by = "singler_main", split.by = "severity",
        reduction = "umap", label = FALSE, pt.size = 0.1) +
  ggtitle("Cell types split by severity")
print(p)
ggsave("plots/phase4/phase4_umap_split_severity.png", p, width = 16, height = 5, dpi = 300)
```

### 3b. Cell-type composition by severity

```{r p4-composition-barplot, fig.width=10, fig.height=6}
comp <- seurat@meta.data %>%
  dplyr::count(severity, singler_main) %>%
  group_by(severity) %>%
  mutate(pct = n / sum(n) * 100)

p <- ggplot(comp, aes(x = severity, y = pct, fill = singler_main)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Severity", y = "Proportion", fill = "Cell type",
       title = "Cell-type composition by severity group") +
  theme_minimal()
print(p)
ggsave("plots/phase4/phase4_composition_barplot.png", p, width = 10, height = 6, dpi = 300)
```

---

## 4. Subset CD14+ monocytes

```{r p4-subset-monocytes}
# Identify monocyte labels from SingleR
mono_labels <- grep("Mono", unique(seurat$singler_main), value = TRUE, ignore.case = TRUE)
cat("Monocyte labels found:", paste(mono_labels, collapse = ", "), "\n")

# If no "Mono" label, try "Monocyte"
if (length(mono_labels) == 0) {
  mono_labels <- grep("Monocyte", unique(seurat$singler_main), value = TRUE, ignore.case = TRUE)
  cat("Trying 'Monocyte' pattern:", paste(mono_labels, collapse = ", "), "\n")
}

mono <- subset(seurat, singler_main %in% mono_labels)
cat("Monocyte subset:\n")
cat("  Cells:", ncol(mono), "\n")
cat("  Severity distribution:\n")
print(table(mono$severity))
```

---

## 5. ISG module scoring

```{r p4-isg-module}
isg_genes <- c("ISG15", "IFIT1", "MX1", "OAS1", "IFI44L")

# Check which genes are present
cat("ISG genes present:",
    paste(isg_genes[isg_genes %in% rownames(mono)], collapse = ", "), "\n")

mono <- AddModuleScore(mono,
                       features = list(isg_genes),
                       name = "ISG")
# Rename the auto-generated column (ISG1) to ISG_score
mono$ISG_score <- mono$ISG1
```

---

## 6. TNF/IL-1β module scoring

```{r p4-tnf-il1-module}
tnf_il1_genes <- c("TNF", "IL1B", "CCL3", "CCL4", "CXCL8")

cat("TNF/IL-1 genes present:",
    paste(tnf_il1_genes[tnf_il1_genes %in% rownames(mono)], collapse = ", "), "\n")

mono <- AddModuleScore(mono,
                       features = list(tnf_il1_genes),
                       name = "TNF_IL1")
mono$TNF_IL1_score <- mono$TNF_IL11
```

---

## 7. Co-activation scatter plot

```{r p4-co-activation-scatter, fig.width=12, fig.height=8}
df <- mono@meta.data %>%
  select(ISG_score, TNF_IL1_score, severity, donor)

p <- ggplot(df, aes(x = ISG_score, y = TNF_IL1_score, color = severity)) +
  geom_point(size = 0.3, alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_color_manual(values = c("Healthy" = "grey60", "Mild" = "steelblue",
                                "Severe" = "firebrick", "Flu" = "darkorange")) +
  labs(x = "ISG module score", y = "TNF/IL-1β module score",
       title = "ISG vs TNF/IL-1β co-activation in monocytes",
       color = "Severity") +
  theme_minimal() +
  facet_wrap(~ severity, ncol = 2)
print(p)
ggsave("plots/phase4/phase4_coactivation_scatter.png", p, width = 12, height = 8, dpi = 300)
```

---

## 8. Define co-activated state

A cell is "co-activated" if **both** its ISG score and TNF/IL-1β score are
above zero (i.e., above the background expectation from `AddModuleScore`).

```{r p4-define-coactivated}
mono$coactivated <- (mono$ISG_score > 0) & (mono$TNF_IL1_score > 0)

cat("Co-activated cells per severity:\n")
coact_summary <- mono@meta.data %>%
  group_by(severity) %>%
  summarise(
    total      = n(),
    coactivated = sum(coactivated),
    pct        = round(100 * coactivated / total, 1),
    .groups    = "drop"
  )
print(coact_summary)
```

---

## 9. Cell-level severity comparison — Violin plots

```{r p4-cell-violin, fig.width=14, fig.height=5}
p1 <- VlnPlot(mono, features = "ISG_score", group.by = "severity",
               pt.size = 0) + ggtitle("ISG score by severity") + NoLegend()
p2 <- VlnPlot(mono, features = "TNF_IL1_score", group.by = "severity",
               pt.size = 0) + ggtitle("TNF/IL-1β score by severity") + NoLegend()
p <- p1 + p2
print(p)
ggsave("plots/phase4/phase4_violin_scores_severity.png", p, width = 14, height = 5, dpi = 300)
```

---

## 10. Donor-level aggregation

```{r p4-donor-aggregation}
donor_summary <- mono@meta.data %>%
  group_by(donor, severity) %>%
  summarise(
    n_cells          = n(),
    mean_ISG         = mean(ISG_score),
    mean_TNF_IL1     = mean(TNF_IL1_score),
    pct_coactivated  = 100 * mean(coactivated),
    .groups = "drop"
  )

knitr::kable(donor_summary, digits = 3,
             caption = "Donor-level monocyte module-score summary")
```

---

## 11. Donor-level statistical tests

```{r p4-donor-stats}
# Subset to COVID patients only for mild vs severe comparison
covid_donors <- donor_summary %>% filter(severity %in% c("Mild", "Severe"))

# Wilcoxon rank-sum tests
cat("=== Wilcoxon rank-sum: Mild vs Severe COVID ===\n\n")

cat("ISG score (donor-level mean):\n")
w_isg <- wilcox.test(mean_ISG ~ severity, data = covid_donors)
print(w_isg)

cat("\nTNF/IL-1 score (donor-level mean):\n")
w_tnf <- wilcox.test(mean_TNF_IL1 ~ severity, data = covid_donors)
print(w_tnf)

cat("\n% Co-activated (donor-level):\n")
w_coact <- wilcox.test(pct_coactivated ~ severity, data = covid_donors)
print(w_coact)

# Kruskal-Wallis across all groups (Healthy, Mild, Severe, Flu)
cat("\n=== Kruskal-Wallis across all severity groups ===\n")
kw <- kruskal.test(pct_coactivated ~ severity, data = donor_summary)
print(kw)
```

---

## 12. Donor-level boxplots

```{r p4-donor-boxplots, fig.width=14, fig.height=5}
p1 <- ggplot(donor_summary, aes(x = severity, y = mean_ISG, fill = severity)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 2) +
  scale_fill_manual(values = c("Healthy" = "grey70", "Mild" = "steelblue",
                               "Severe" = "firebrick", "Flu" = "darkorange")) +
  labs(y = "Mean ISG score", title = "ISG score per donor") +
  theme_minimal() + NoLegend()

p2 <- ggplot(donor_summary, aes(x = severity, y = mean_TNF_IL1, fill = severity)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 2) +
  scale_fill_manual(values = c("Healthy" = "grey70", "Mild" = "steelblue",
                               "Severe" = "firebrick", "Flu" = "darkorange")) +
  labs(y = "Mean TNF/IL-1β score", title = "TNF/IL-1β score per donor") +
  theme_minimal() + NoLegend()

p3 <- ggplot(donor_summary, aes(x = severity, y = pct_coactivated, fill = severity)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 2) +
  scale_fill_manual(values = c("Healthy" = "grey70", "Mild" = "steelblue",
                               "Severe" = "firebrick", "Flu" = "darkorange")) +
  labs(y = "% co-activated monocytes", title = "Co-activation per donor") +
  theme_minimal() + NoLegend()

p <- p1 + p2 + p3
print(p)
ggsave("plots/phase4/phase4_donor_boxplots.png", p, width = 14, height = 5, dpi = 300)
```

---

## 13. Cell-type-specific differential expression (monocytes)

```{r p4-mono-de}
# Compare severe COVID monocytes vs healthy monocytes
mono$severity_ident <- mono$severity
Idents(mono) <- "severity_ident"

mono_de <- FindMarkers(
  mono,
  ident.1 = "Severe",
  ident.2 = "Healthy",
  min.pct = 0.1,
  logfc.threshold = 0.25
)

cat("DE genes (Severe vs Healthy monocytes):", nrow(mono_de), "\n")
cat("\nTop 20 upregulated:\n")
print(head(mono_de[order(-mono_de$avg_log2FC), ], 20))
cat("\nTop 20 downregulated:\n")
print(head(mono_de[order(mono_de$avg_log2FC), ], 20))
```

---

## 14. Volcano plot

```{r p4-volcano, fig.width=10, fig.height=8}
p <- EnhancedVolcano(
  mono_de,
  lab        = rownames(mono_de),
  x          = "avg_log2FC",
  y          = "p_val_adj",
  pCutoff    = 0.05,
  FCcutoff   = 1.0,
  title      = "Severe vs Healthy COVID — CD14+ Monocytes",
  subtitle   = "Differential expression",
  pointSize  = 1.5,
  labSize    = 3.5
)
print(p)
ggsave("plots/phase4/phase4_volcano_monocyte_DE.png", p, width = 10, height = 8, dpi = 300)
```

---

## 15. Save results

```{r p4-save-results}
# Save monocyte subset with scores
saveRDS(mono, file = "Data/phase4_monocytes.rds")

# Save DE results
write.csv(mono_de, file = "Data/phase4_mono_de_severe_vs_healthy.csv")

# Save donor summary
write.csv(donor_summary, file = "Data/phase4_donor_summary.csv",
          row.names = FALSE)

# Save the full Seurat with severity metadata for downstream phases
saveRDS(seurat, file = "Data/phase4_seurat_with_severity.rds")

cat("Saved:\n")
cat("  Data/phase4_monocytes.rds\n")
cat("  Data/phase4_mono_de_severe_vs_healthy.csv\n")
cat("  Data/phase4_donor_summary.csv\n")
cat("  Data/phase4_seurat_with_severity.rds\n")
```

---

## 16. Predict-Before-You-Run 1

**Prediction:** "I expect ISG + TNF/IL-1β co-activation frequency to be higher
in severe COVID-19 monocytes, because severe disease is characterized by both
interferon dysregulation and monocyte-driven cytokine storm (as described by
Lee et al., Sci Immunol 2020)."

**Outcome:** *(Fill in after running — does the data support the prediction?)*

- % co-activated in Severe: `r coact_summary$pct[coact_summary$severity == "Severe"]`%
- % co-activated in Mild: `r coact_summary$pct[coact_summary$severity == "Mild"]`%
- % co-activated in Healthy: `r coact_summary$pct[coact_summary$severity == "Healthy"]`%
- Wilcoxon p-value (Severe vs Mild, donor-level): `r w_coact$p.value`

---

## 17. Next step

Continue with Phase 5 (Replication on GSE150861):

- Independent validation dataset
- Same QC → normalize → cluster → annotate pipeline
- Apply identical ISG and TNF/IL-1β module scores
- Confirm co-activated monocyte state in independent cohort

---

# ====================================================================
# PHASE5 - REPLICATION GSE150861
# ====================================================================


## Goal

This document implements **Phase 5** — independent validation on GSE150861
(Guo et al., Nat Commun 2020). This dataset contains PBMCs from **2 severe
COVID-19 patients** treated with **Tocilizumab** (anti-IL-6R), sampled at
multiple time points (day 1, day 5, day 7). We process this dataset
**independently** (no joint integration with GSE149689) and check whether:
 
1. The ISG + TNF/IL-1β co-activated monocyte state from Phase 4 replicates.
2. Co-activation changes over time (longitudinal dynamics).
3. Tocilizumab treatment reduces the TNF/IL-1β component.

---

## 1. Setup

```{r p5-p5-init, message=FALSE, warning=FALSE}
dir.create("plots/phase5", showWarnings = FALSE, recursive = TRUE)
```

---

## 2. Load GSE150861 10X data

```{r p5-load-data}
counts <- Read10X(data.dir = "Data/GSE150861/")

cat("Raw matrix dimensions:\n")
cat("  Genes:", nrow(counts), "\n")
cat("  Barcodes:", ncol(counts), "\n")
```

---

## 3. Load batch metadata and create Seurat object

```{r p5-create-seurat}
# Load batch info (barcode → patient-day-rep mapping)
batch_info <- read.csv(gzfile("Data/GSE150861/cell_batch.csv.gz"),
                       stringsAsFactors = FALSE)
rownames(batch_info) <- batch_info$X

# Only keep barcodes that are in both the matrix and batch file
common_barcodes <- intersect(colnames(counts), batch_info$X)
cat("Barcodes in matrix:", ncol(counts), "\n")
cat("Barcodes in batch file:", nrow(batch_info), "\n")
cat("Common barcodes:", length(common_barcodes), "\n")

counts <- counts[, common_barcodes]

# Create Seurat object
rep_seurat <- CreateSeuratObject(
  counts  = counts,
  project = "GSE150861_COVID_Toci",
  min.cells    = 3,
  min.features = 200
)

# Add batch metadata
rep_seurat$batch <- batch_info[colnames(rep_seurat), "batch"]

# Parse batch into patient, day, replicate
rep_seurat$patient <- sub("-day.*", "", rep_seurat$batch)
rep_seurat$day     <- as.integer(sub(".*day(\\d+).*", "\\1", rep_seurat$batch))
rep_seurat$day_label <- paste0("Day ", rep_seurat$day)

cat("\nSeurat object created.\n")
cat("  Cells:", ncol(rep_seurat), "\n")
cat("  Genes:", nrow(rep_seurat), "\n")
cat("\nBatch distribution:\n")
print(table(rep_seurat$batch))
cat("\nPatient × Day:\n")
print(table(rep_seurat$patient, rep_seurat$day_label))
```

---

## 4. QC — same pipeline as Phase 1

```{r p5-qc-metrics}
rep_seurat[["percent.mt"]] <- PercentageFeatureSet(rep_seurat, pattern = "^MT-")
```

### 4a. QC violin plots

```{r p5-qc-violin, fig.width=12, fig.height=4}
p <- VlnPlot(rep_seurat,
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
        ncol = 3, pt.size = 0)
print(p)
ggsave("plots/phase5/phase5_qc_violin.png", p, width = 12, height = 4, dpi = 300)
```

### 4b. QC scatter plots

```{r p5-qc-scatter, fig.width=12, fig.height=4}
p1 <- FeatureScatter(rep_seurat, feature1 = "nCount_RNA", feature2 = "percent.mt") + NoLegend()
p2 <- FeatureScatter(rep_seurat, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") + NoLegend()
p <- p1 + p2
print(p)
ggsave("plots/phase5/phase5_qc_scatter.png", p, width = 12, height = 4, dpi = 300)
```

### 4c. Filter cells

```{r p5-filter-cells}
cat("Before filtering:", ncol(rep_seurat), "cells\n")

rep_seurat <- subset(rep_seurat,
                     subset = nFeature_RNA > 200 &
                              nFeature_RNA < 5000 &
                              percent.mt < 20)

cat("After filtering:", ncol(rep_seurat), "cells\n")
cat("\nBatch distribution after filtering:\n")
print(table(rep_seurat$batch))
```

---

## 5. Normalization (LogNormalize — same as Phase 2 draft)

#TODO: Switch to SCTransform on SSH server with more RAM.

```{r p5-normalize}
rep_seurat <- NormalizeData(rep_seurat, normalization.method = "LogNormalize",
                            scale.factor = 10000)
rep_seurat <- FindVariableFeatures(rep_seurat, selection.method = "vst",
                                    nfeatures = 3000)
rep_seurat <- ScaleData(rep_seurat, vars.to.regress = "percent.mt")
```

---

## 6. PCA → UMAP → Clustering

```{r p5-pca}
N_PCS <- 20

rep_seurat <- RunPCA(rep_seurat, verbose = FALSE)
```

```{r p5-elbow, fig.width=8, fig.height=4}
p <- ElbowPlot(rep_seurat, ndims = 30) +
  geom_vline(xintercept = N_PCS, linetype = "dashed", color = "red")
print(p)
ggsave("plots/phase5/phase5_elbow_plot.png", p, width = 8, height = 4, dpi = 300)
```

```{r p5-umap-cluster}
rep_seurat <- RunUMAP(rep_seurat, dims = 1:N_PCS, verbose = FALSE)
rep_seurat <- FindNeighbors(rep_seurat, dims = 1:N_PCS, verbose = FALSE)
rep_seurat <- FindClusters(rep_seurat, resolution = 0.5, verbose = FALSE)
```

```{r p5-umap-plot, fig.width=14, fig.height=5}
p1 <- DimPlot(rep_seurat, reduction = "umap", label = TRUE) +
  ggtitle("Clusters") + NoLegend()
p2 <- DimPlot(rep_seurat, reduction = "umap", group.by = "patient") +
  ggtitle("Patient")
p3 <- DimPlot(rep_seurat, reduction = "umap", group.by = "day_label") +
  ggtitle("Time point")
p <- p1 + p2 + p3
print(p)
ggsave("plots/phase5/phase5_umap_overview.png", p, width = 14, height = 5, dpi = 300)
```

---

## 7. Cell-type annotation — manual markers + SingleR

### 7a. Manual marker feature plots

```{r p5-marker-featureplots, fig.width=16, fig.height=12}
markers <- c("CD14", "LYZ", "FCGR3A",          # Monocytes
             "CD3D", "CD3E", "CD4", "CD8A",     # T cells
             "MS4A1", "CD79A",                    # B cells
             "NKG7", "GNLY",                      # NK cells
             "FCER1A", "CST3")                    # DCs
p <- FeaturePlot(rep_seurat, features = markers, reduction = "umap", ncol = 4)
print(p)
ggsave("plots/phase5/phase5_marker_featureplots.png", p, width = 16, height = 12, dpi = 300)
```

### 7b. SingleR annotation

```{r p5-singler}
singler_ref <- BlueprintEncodeData()

expr_matrix <- GetAssayData(rep_seurat, layer = "data")
singler_results <- SingleR(
  test   = expr_matrix,
  ref    = singler_ref,
  labels = singler_ref$label.main
)

rep_seurat$singler_main <- singler_results$labels

cat("SingleR label distribution:\n")
print(sort(table(rep_seurat$singler_main), decreasing = TRUE))
```

### 7c. UMAP by cell type

```{r p5-umap-celltype, fig.width=12, fig.height=7}
p <- DimPlot(rep_seurat, group.by = "singler_main", reduction = "umap",
        label = TRUE, repel = TRUE) +
  ggtitle("GSE150861 \u2014 Cell-type annotation (SingleR)")
print(p)
ggsave("plots/phase5/phase5_umap_celltype.png", p, width = 12, height = 7, dpi = 300)
```

---

## 8. Subset monocytes

```{r p5-subset-mono}
mono_labels <- grep("Mono|Monocyte", unique(rep_seurat$singler_main),
                    value = TRUE, ignore.case = TRUE)
cat("Monocyte labels:", paste(mono_labels, collapse = ", "), "\n")

rep_mono <- subset(rep_seurat, singler_main %in% mono_labels)
cat("Monocyte cells:", ncol(rep_mono), "\n")
cat("\nMonocytes per time point:\n")
print(table(rep_mono$day_label, rep_mono$patient))
```

---

## 9. ISG and TNF/IL-1β module scoring

Using the **exact same gene lists** as Phase 4 (GSE149689).

```{r p5-module-scores}
isg_genes     <- c("ISG15", "IFIT1", "MX1", "OAS1", "IFI44L")
tnf_il1_genes <- c("TNF", "IL1B", "CCL3", "CCL4", "CXCL8")

cat("ISG genes present:",
    paste(isg_genes[isg_genes %in% rownames(rep_mono)], collapse = ", "), "\n")
cat("TNF/IL-1 genes present:",
    paste(tnf_il1_genes[tnf_il1_genes %in% rownames(rep_mono)], collapse = ", "), "\n")

rep_mono <- AddModuleScore(rep_mono, features = list(isg_genes), name = "ISG")
rep_mono$ISG_score <- rep_mono$ISG1

rep_mono <- AddModuleScore(rep_mono, features = list(tnf_il1_genes), name = "TNF_IL1")
rep_mono$TNF_IL1_score <- rep_mono$TNF_IL11
```

---

## 10. Co-activation scatter — does the co-activated state replicate?

```{r p5-replication-scatter, fig.width=12, fig.height=6}
df <- rep_mono@meta.data %>%
  select(ISG_score, TNF_IL1_score, patient, day_label, day)

p <- ggplot(df, aes(x = ISG_score, y = TNF_IL1_score, color = day_label)) +
  geom_point(size = 0.5, alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_color_brewer(palette = "Set1") +
  labs(x = "ISG module score", y = "TNF/IL-1\u03b2 module score",
       title = "Replication: ISG vs TNF/IL-1\u03b2 in GSE150861 monocytes",
       color = "Time point") +
  theme_minimal() +
  facet_wrap(~ patient)
print(p)
ggsave("plots/phase5/phase5_coactivation_scatter.png", p, width = 12, height = 6, dpi = 300)
```

---

## 11. Define co-activated state and quantify over time

```{r p5-coactivation-time}
rep_mono$coactivated <- (rep_mono$ISG_score > 0) & (rep_mono$TNF_IL1_score > 0)

coact_by_time <- rep_mono@meta.data %>%
  group_by(patient, day, day_label) %>%
  summarise(
    n_cells         = n(),
    n_coactivated   = sum(coactivated),
    pct_coactivated = round(100 * n_coactivated / n_cells, 1),
    mean_ISG        = mean(ISG_score),
    mean_TNF_IL1    = mean(TNF_IL1_score),
    .groups = "drop"
  ) %>%
  arrange(patient, day)

knitr::kable(coact_by_time,
             caption = "Co-activation in monocytes over time (per patient-day)")
```

### 11a. Longitudinal dynamics plot

```{r p5-longitudinal-plot, fig.width=12, fig.height=5}
p1 <- ggplot(coact_by_time, aes(x = day, y = pct_coactivated,
                                 color = patient, group = patient)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  labs(x = "Day", y = "% co-activated monocytes",
       title = "Co-activation over time") +
  theme_minimal()

p2 <- ggplot(coact_by_time, aes(x = day, y = mean_ISG,
                                 color = patient, group = patient)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  labs(x = "Day", y = "Mean ISG score",
       title = "ISG score over time") +
  theme_minimal()

p3 <- ggplot(coact_by_time, aes(x = day, y = mean_TNF_IL1,
                                 color = patient, group = patient)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  labs(x = "Day", y = "Mean TNF/IL-1β score",
       title = "TNF/IL-1β score over time") +
  theme_minimal()

p <- p1 + p2 + p3
print(p)
ggsave("plots/phase5/phase5_longitudinal_dynamics.png", p, width = 12, height = 5, dpi = 300)
```

---

## 12. Tocilizumab effect on TNF/IL-1β component

```{r p5-toci-effect, fig.width=12, fig.height=5}
# Violin plot of module scores at Day 1 vs later time points
p1 <- VlnPlot(rep_mono, features = "ISG_score", group.by = "day_label",
               pt.size = 0) +
  ggtitle("ISG score by time point") + NoLegend()
p2 <- VlnPlot(rep_mono, features = "TNF_IL1_score", group.by = "day_label",
               pt.size = 0) +
  ggtitle("TNF/IL-1β score by time point") + NoLegend()
p <- p1 + p2
print(p)
ggsave("plots/phase5/phase5_toci_violin.png", p, width = 12, height = 5, dpi = 300)
```

### 12a. Statistical test: Day 1 vs later

```{r p5-toci-stats}
mono_md <- rep_mono@meta.data

# Classify as early (day 1) vs post-treatment (day 5+)
mono_md$phase <- ifelse(mono_md$day == 1, "Pre-Toci (Day 1)", "Post-Toci (Day 5+)")

cat("=== Wilcoxon: ISG score Pre vs Post Tocilizumab ===\n")
print(wilcox.test(ISG_score ~ phase, data = mono_md))

cat("\n=== Wilcoxon: TNF/IL-1 score Pre vs Post Tocilizumab ===\n")
print(wilcox.test(TNF_IL1_score ~ phase, data = mono_md))

cat("\n=== Proportion co-activated: Pre vs Post ===\n")
prop_table <- mono_md %>%
  group_by(phase) %>%
  summarise(
    n = n(),
    coact = sum(coactivated),
    pct   = round(100 * coact / n, 1),
    .groups = "drop"
  )
print(prop_table)

cat("\nFisher's exact test (co-activation Pre vs Post):\n")
mat <- matrix(c(
  prop_table$coact[1], prop_table$n[1] - prop_table$coact[1],
  prop_table$coact[2], prop_table$n[2] - prop_table$coact[2]
), nrow = 2, byrow = TRUE)
print(fisher.test(mat))
```

---

## 13. Cell-type composition over time

```{r p5-composition-time, fig.width=12, fig.height=6}
comp <- rep_seurat@meta.data %>%
  dplyr::count(day_label, patient, singler_main) %>%
  group_by(day_label, patient) %>%
  mutate(pct = n / sum(n) * 100)

p <- ggplot(comp, aes(x = day_label, y = pct, fill = singler_main)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~ patient) +
  labs(x = "Time point", y = "Proportion", fill = "Cell type",
       title = "Cell-type composition over time (GSE150861)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(p)
ggsave("plots/phase5/phase5_composition_time.png", p, width = 12, height = 6, dpi = 300)
```

---

## 14. Save results

```{r p5-save-results}
saveRDS(rep_seurat, file = "Data/phase5_GSE150861_seurat.rds")
saveRDS(rep_mono,   file = "Data/phase5_GSE150861_monocytes.rds")
write.csv(coact_by_time, file = "Data/phase5_coactivation_longitudinal.csv",
          row.names = FALSE)

cat("Saved:\n")
cat("  Data/phase5_GSE150861_seurat.rds\n")
cat("  Data/phase5_GSE150861_monocytes.rds\n")
cat("  Data/phase5_coactivation_longitudinal.csv\n")
```

---

## 15. Replication summary

**Key questions and expected answers:**

1. **Does the co-activated monocyte state replicate?**
   *(Fill in after running — do the scatter plots show the upper-right quadrant
   population?)*

2. **Longitudinal dynamics:**
   *(Does co-activation % decrease at later time points?)*

3. **Tocilizumab effect:**
   *(Does IL-6 blockade reduce the TNF/IL-1β component specifically, while
   ISG scores may remain unchanged?)*

---

## 16. Next step

Continue with Phase 6 (scRNA-seq Visualization Summary) — collect the key
figures from Phases 3–5 into a comprehensive visual overview.

---

# ====================================================================
# PHASE6 - SCRNASEQ VISUALIZATION SUMMARY
# ====================================================================


## Goal

This document implements **Phase 6** of the project workflow:

Compile all key scRNA-seq visualizations from Phases 3–5 into a single
publication-ready summary document. Every plot is saved to `plots/phase6/`.

---

## 1. Setup

```{r p6-p6-init, message=FALSE, warning=FALSE}
dir.create("plots/phase6", showWarnings = FALSE, recursive = TRUE)
```

---

## 2. Load Phase 3–5 outputs

```{r p6-load-data}
# ── Phase 3 outputs ──
seurat  <- readRDS("Data/phase3_seurat_annotated.rds")
markers <- read.csv("Data/phase3_all_markers.csv")

cat("Phase 3 Seurat object:", ncol(seurat), "cells,", nrow(seurat), "genes\n")

# ── Phase 4 outputs ──
seurat_sev  <- readRDS("Data/phase4_seurat_with_severity.rds")
mono        <- readRDS("Data/phase4_monocytes.rds")
mono_de     <- read.csv("Data/phase4_mono_de_severe_vs_healthy.csv")
donor_summary <- read.csv("Data/phase4_donor_summary.csv")

cat("Phase 4 monocytes:", ncol(mono), "cells\n")
cat("Phase 4 DE genes:", nrow(mono_de), "\n")

# ── Phase 5 outputs ──
rep_mono <- readRDS("Data/phase5_GSE150861_monocytes.rds")
coact_longitudinal <- read.csv("Data/phase5_coactivation_longitudinal.csv")

cat("Phase 5 replication monocytes:", ncol(rep_mono), "cells\n")
```

---

## 3. Plot 1 — UMAP by cell type (overview of PBMC populations in GSE149689)

```{r p6-umap-celltype, fig.width=10, fig.height=7}
p1 <- DimPlot(seurat_sev, group.by = "singler_main", reduction = "umap",
              label = TRUE, repel = TRUE, pt.size = 0.3) +
  ggtitle("GSE149689 — UMAP coloured by cell type (SingleR)") +
  theme(legend.position = "right")

print(p1)

ggsave("plots/phase6/phase6_01_umap_celltype.png", p1,
       width = 10, height = 7, dpi = 300)
```

---

## 4. Plot 2 — UMAP split by severity

```{r p6-umap-severity, fig.width=16, fig.height=5}
p2 <- DimPlot(seurat_sev, group.by = "singler_main", split.by = "severity",
              reduction = "umap", label = FALSE, pt.size = 0.1) +
  ggtitle("Cell types split by severity group")

print(p2)

ggsave("plots/phase6/phase6_02_umap_split_severity.png", p2,
       width = 16, height = 5, dpi = 300)
```

---

## 5. Plot 3 — Feature plots: ISG genes on UMAP

```{r p6-feature-isg, fig.width=15, fig.height=5}
isg_genes <- c("ISG15", "IFIT1", "MX1")

p3 <- FeaturePlot(seurat_sev, features = isg_genes, reduction = "umap",
                  ncol = 3, pt.size = 0.1, order = TRUE) &
  scale_colour_viridis_c() &
  theme_minimal()

print(p3)

ggsave("plots/phase6/phase6_03_feature_ISG.png", p3,
       width = 15, height = 5, dpi = 300)
```

---

## 6. Plot 4 — Feature plots: TNF/IL-1β genes on UMAP

```{r p6-feature-tnf, fig.width=15, fig.height=5}
tnf_genes <- c("TNF", "IL1B", "CCL3")

p4 <- FeaturePlot(seurat_sev, features = tnf_genes, reduction = "umap",
                  ncol = 3, pt.size = 0.1, order = TRUE) &
  scale_colour_viridis_c() &
  theme_minimal()

print(p4)

ggsave("plots/phase6/phase6_04_feature_TNF_IL1.png", p4,
       width = 15, height = 5, dpi = 300)
```

---

## 7. Plot 5 — Violin plots: module scores by severity (monocytes)

```{r p6-violin-scores, fig.width=12, fig.height=5}
# Violin for ISG score
v_isg <- VlnPlot(mono, features = "ISG_score", group.by = "severity",
                 pt.size = 0) +
  ggtitle("ISG module score — monocytes") +
  xlab("Severity") + ylab("ISG score") +
  NoLegend()

# Violin for TNF/IL-1β score
v_tnf <- VlnPlot(mono, features = "TNF_IL1_score", group.by = "severity",
                 pt.size = 0) +
  ggtitle("TNF/IL-1β module score — monocytes") +
  xlab("Severity") + ylab("TNF/IL-1β score") +
  NoLegend()

p5 <- v_isg | v_tnf

print(p5)

ggsave("plots/phase6/phase6_05_violin_module_scores.png", p5,
       width = 12, height = 5, dpi = 300)
```

---

## 8. Plot 6 — Donor-level boxplot: co-activation % by severity (key figure)

```{r p6-donor-boxplot, fig.width=8, fig.height=6}
p6 <- ggplot(donor_summary, aes(x = severity, y = pct_coactivated,
                                fill = severity)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.15, size = 3, alpha = 0.8) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Co-activated monocyte percentage per donor",
       subtitle = "ISG score > 0 AND TNF/IL-1β score > 0",
       x = "Severity group",
       y = "% co-activated monocytes") +
  theme_classic(base_size = 14) +
  theme(legend.position = "none")

print(p6)

ggsave("plots/phase6/phase6_06_donor_coactivation_boxplot.png", p6,
       width = 8, height = 6, dpi = 300)
```

---

## 9. Plot 7 — Volcano plot: monocyte DE (severe vs healthy)

```{r p6-volcano, fig.width=10, fig.height=8}
# Restore gene names as row names if loaded without row.names
if ("X" %in% colnames(mono_de)) {
  rownames(mono_de) <- mono_de$X
}

p7 <- EnhancedVolcano(mono_de,
                      lab = rownames(mono_de),
                      x   = "avg_log2FC",
                      y   = "p_val_adj",
                      title    = "Monocyte DE: Severe COVID vs Healthy",
                      subtitle = "Wilcoxon rank-sum test",
                      pCutoff  = 0.05,
                      FCcutoff = 0.5,
                      pointSize = 1.5,
                      labSize   = 4,
                      drawConnectors = TRUE,
                      widthConnectors = 0.3,
                      maxoverlapsConnectors = 20)

print(p7)

ggsave("plots/phase6/phase6_07_volcano_monocyte_DE.png", p7,
       width = 10, height = 8, dpi = 300)
```

---

## 10. Plot 8 — Marker heatmap: top 5 markers per cluster

```{r p6-marker-heatmap, fig.width=14, fig.height=10}
top5 <- markers %>%
  group_by(cluster) %>%
  top_n(n = 5, wt = avg_log2FC) %>%
  ungroup()

# Downsample for heatmap rendering
set.seed(42)
cells_sub <- seurat[, sample(ncol(seurat),
                             min(ncol(seurat), 5000))]

p8 <- DoHeatmap(cells_sub, features = top5$gene, group.by = "seurat_clusters",
                size = 3) +
  ggtitle("Top 5 markers per cluster (downsampled)") +
  theme(axis.text.y = element_text(size = 6))

print(p8)

ggsave("plots/phase6/phase6_08_marker_heatmap.png", p8,
       width = 14, height = 10, dpi = 300)
```

---

## 11. Plot 9 — Cell-type composition barplot by severity

```{r p6-composition-barplot, fig.width=10, fig.height=6}
comp <- seurat_sev@meta.data %>%
  dplyr::count(severity, singler_main) %>%
  group_by(severity) %>%
  mutate(pct = n / sum(n) * 100)

p9 <- ggplot(comp, aes(x = severity, y = pct, fill = singler_main)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Cell-type composition by severity",
       x = "Severity group", y = "Proportion",
       fill = "Cell type") +
  theme_classic(base_size = 13)

print(p9)

ggsave("plots/phase6/phase6_09_composition_barplot.png", p9,
       width = 10, height = 6, dpi = 300)
```

---

## 12. Plot 10 — GSE150861 replication: co-activation scatter

```{r p6-replication-scatter, fig.width=10, fig.height=6}
rep_df <- rep_mono@meta.data %>%
  select(ISG_score, TNF_IL1_score, day_label, patient)

p10 <- ggplot(rep_df, aes(x = ISG_score, y = TNF_IL1_score,
                          colour = day_label)) +
  geom_point(size = 0.5, alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  facet_wrap(~ patient, ncol = 2) +
  scale_colour_brewer(palette = "Dark2") +
  labs(title = "GSE150861 replication — monocyte co-activation scatter",
       subtitle = "Upper-right quadrant = co-activated (ISG > 0 AND TNF/IL-1β > 0)",
       x = "ISG module score",
       y = "TNF/IL-1β module score",
       colour = "Time point") +
  theme_bw(base_size = 12)

print(p10)

ggsave("plots/phase6/phase6_10_replication_coactivation.png", p10,
       width = 10, height = 6, dpi = 300)
```

---

## 13. Summary panel — key figures combined

```{r p6-summary-panel, fig.width=20, fig.height=12}
# Combine the four most important figures into one panel
panel <- (p1 | p6) / (p7 | p10) +
  plot_annotation(
    title    = "scRNA-seq Analysis Summary — COVID-19 Monocyte Co-Activation",
    subtitle = "GSE149689 (discovery) + GSE150861 (replication)",
    tag_levels = "A"
  )

print(panel)

ggsave("plots/phase6/phase6_summary_panel.png", panel,
       width = 20, height = 12, dpi = 300)
```

---

## Decision Checkpoint 6

> **Summary of key findings visualized:**
>
> 1.
> 2.
> 3.
>
> (Fill in after rendering.)

---

## Session info

```{r p6-session-info}
sessionInfo()
```

---

# ====================================================================
# PHASE7 - BULK LOADING EDA
# ====================================================================


## Goal

This document implements **Phase 7** — loading and exploring the bulk RNA-seq
dataset GSE152418 (Arunachalam et al., Science 2020). This dataset contains
whole-blood RNA-seq from COVID-19 patients of varying severity plus healthy
controls. We will:

1. Load the raw count matrix and sample metadata
2. Clean and define severity groups
3. Inspect group sizes, gender balance, and missing data
4. Filter low-count genes
5. Align and verify count matrix with metadata

---

## 1. Setup

```{r p7-p7-init, message=FALSE, warning=FALSE}
dir.create("plots/phase7", showWarnings = FALSE, recursive = TRUE)
```

---

## 2. Load the raw count matrix

```{r p7-load-counts}
counts_raw <- read.table("Data/GSE152418/GSE152418_p20047_Study1_RawCounts.txt.gz",
                         header = TRUE, sep = "\t", row.names = 1,
                         check.names = FALSE)

cat("Raw count matrix dimensions:\n")
cat("  Genes:", nrow(counts_raw), "\n")
cat("  Samples:", ncol(counts_raw), "\n")
cat("\nFirst few gene names:\n")
head(rownames(counts_raw))
cat("\nFirst few sample names:\n")
head(colnames(counts_raw))
cat("\nCorner of the matrix:\n")
print(counts_raw[1:5, 1:5])
```

---

## 3. Load metadata from GEO series matrix

```{r p7-load-metadata}
# Try loading from local series matrix file
geo <- getGEO(filename = "Data/GSE152418/GSE152418_series_matrix.txt.gz",
              getGPL = FALSE)

metadata <- pData(geo)

cat("Metadata dimensions:", nrow(metadata), "samples x", ncol(metadata), "columns\n")
cat("\nColumn names:\n")
print(colnames(metadata))
```

---

## 4. Clean and inspect metadata

```{r p7-clean-metadata}
# Inspect key columns
cat("Available characteristics columns:\n")
char_cols <- grep("characteristics", colnames(metadata), value = TRUE)
print(char_cols)

# Print first few values of each characteristics column
for (col in char_cols) {
  cat("\n", col, ":\n")
  print(head(unique(metadata[[col]]), 20))
}

# Extract relevant fields from characteristics columns
# (Adjust column names based on what the series matrix contains)
metadata$title_clean <- metadata$title

# Try to extract disease status and severity
# These are typically in characteristics_ch1, characteristics_ch1.1, etc.
cat("\nSample titles:\n")
print(metadata$title)
```

---

## 5. Parse severity and demographic information

```{r p7-parse-severity}
# Extract fields from characteristics columns
# GEO typically stores: disease state, severity, gender, age, etc.

# Parse each characteristics column
for (col in char_cols) {
  vals <- as.character(metadata[[col]])
  # Extract field name and value
  if (any(grepl(":", vals))) {
    field_name <- sub(":.*", "", vals[1])
    field_name <- trimws(tolower(gsub(" ", "_", field_name)))
    metadata[[field_name]] <- trimws(sub(".*:", "", vals))
    cat("Parsed:", field_name, "→", length(unique(metadata[[field_name]])), "unique values\n")
    print(table(metadata[[field_name]]))
    cat("\n")
  }
}
```

---

## 6. Define severity contrast groups

```{r p7-define-groups}
# The severity column was parsed from characteristics_ch1.3
# Values: Convalescent, Moderate, Severe, ICU, Healthy
cat("Severity values parsed from GEO:\n")
print(table(metadata$severity))

# Create binary severity group:
# Severe_ICU = Severe + ICU patients
# Healthy_Mild = Healthy + Moderate patients
# Exclude Convalescent
metadata$severity_group <- case_when(
  metadata$severity %in% c("Healthy")              ~ "Healthy_Mild",
  metadata$severity %in% c("Moderate")             ~ "Healthy_Mild",
  metadata$severity %in% c("Severe", "ICU")        ~ "Severe_ICU",
  metadata$severity %in% c("Convalescent")         ~ NA_character_,
  TRUE                                              ~ NA_character_
)

cat("\nSeverity group distribution:\n")
print(table(metadata$severity_group, useNA = "ifany"))

# Also show original severity within each group
cat("\nOriginal severity within groups:\n")
print(table(metadata$severity, metadata$severity_group, useNA = "ifany"))

# Remove convalescent / unclassified samples
metadata_clean <- metadata %>% filter(!is.na(severity_group))
cat("\nSamples after excluding convalescent:", nrow(metadata_clean), "\n")

# Set as factor with reference level
metadata_clean$severity_group <- relevel(factor(metadata_clean$severity_group),
                                         ref = "Healthy_Mild")
```

---

## 7. Inspect group sizes and gender balance

```{r p7-group-sizes, fig.width=10, fig.height=5}
cat("Group sizes:\n")
print(table(metadata_clean$severity_group))

# Check for gender column
gender_col <- grep("gender|sex", colnames(metadata_clean), value = TRUE, ignore.case = TRUE)
if (length(gender_col) > 0) {
  cat("\nGender distribution:\n")
  print(table(metadata_clean[[gender_col[1]]]))

  cat("\nGender × Severity:\n")
  gender_sev_table <- table(metadata_clean[[gender_col[1]]], metadata_clean$severity_group)
  print(gender_sev_table)

  # Visualise
  p <- ggplot(metadata_clean, aes(x = severity_group, fill = .data[[gender_col[1]]])) +
    geom_bar(position = "dodge") +
    labs(title = "Gender balance across severity groups",
         x = "Severity group", y = "Count", fill = "Gender") +
    theme_classic(base_size = 14)
  print(p)
  ggsave("plots/phase7/phase7_gender_balance.png", p, width = 8, height = 5, dpi = 300)
} else {
  cat("\nNo gender column found in metadata.\n")
}
```

---

## 8. Check missing data

```{r p7-missing-data}
cat("Percentage of non-missing values per metadata column:\n")
completeness <- colMeans(!is.na(metadata_clean)) * 100
print(round(completeness[completeness < 100], 1))
if (all(completeness == 100)) cat("All columns are 100% complete.\n")
```

---

## 9. Severity distribution bar plot

```{r p7-severity-barplot, fig.width=8, fig.height=5}
p <- ggplot(metadata_clean, aes(x = severity_group, fill = severity_group)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
  scale_fill_manual(values = c("Healthy_Mild" = "steelblue",
                                "Severe_ICU" = "firebrick")) +
  labs(title = "Sample distribution by severity group (GSE152418)",
       x = "Severity group", y = "Number of samples") +
  theme_classic(base_size = 14) +
  theme(legend.position = "none")

print(p)
ggsave("plots/phase7/phase7_severity_distribution.png", p, width = 8, height = 5, dpi = 300)
```

---

## 10. Align count matrix with metadata

```{r p7-align-counts}
# Find matching sample identifiers between counts and metadata
cat("Count matrix sample names (first 10):\n")
print(head(colnames(counts_raw), 10))
cat("\nMetadata sample identifiers (first 10):\n")
print(head(metadata_clean$title, 10))
cat("\nMetadata geo_accession (first 10):\n")
print(head(metadata_clean$geo_accession, 10))

# Try to match by title or geo_accession
if (any(colnames(counts_raw) %in% metadata_clean$title)) {
  match_col <- "title"
} else if (any(colnames(counts_raw) %in% metadata_clean$geo_accession)) {
  match_col <- "geo_accession"
} else {
  # Try substring matching
  cat("\nNo direct match found. Attempting partial matching...\n")
  match_col <- NULL
}

if (!is.null(match_col)) {
  # Subset metadata to samples present in count matrix
  shared <- intersect(colnames(counts_raw), metadata_clean[[match_col]])
  cat("\nShared samples between counts and metadata:", length(shared), "\n")

  # Align
  metadata_clean <- metadata_clean[metadata_clean[[match_col]] %in% shared, ]
  rownames(metadata_clean) <- metadata_clean[[match_col]]
  counts_aligned <- counts_raw[, metadata_clean[[match_col]]]

  # Verify
  cat("Alignment check:", all(colnames(counts_aligned) == rownames(metadata_clean)), "\n")
} else {
  cat("Manual alignment needed — inspect sample names above.\n")
  counts_aligned <- counts_raw
}

cat("\nAligned matrix: ", nrow(counts_aligned), "genes x", ncol(counts_aligned), "samples\n")
```

---

## 11. Filter low-count genes

```{r p7-filter-genes}
cat("Genes before filtering:", nrow(counts_aligned), "\n")

# Strategy: keep genes with at least 10 counts in at least 3 samples
keep <- rowSums(counts_aligned >= 10) >= 3
counts_filtered <- counts_aligned[keep, ]

cat("Genes after filtering:", nrow(counts_filtered), "\n")
cat("Genes removed:", sum(!keep), "\n")
cat("Percentage retained:", round(100 * sum(keep) / length(keep), 1), "%\n")
```

---

## 12. Library size overview

```{r p7-library-sizes, fig.width=10, fig.height=5}
lib_sizes <- colSums(counts_filtered)

p <- data.frame(
  sample = names(lib_sizes),
  lib_size = lib_sizes,
  severity = metadata_clean$severity_group
) %>%
  ggplot(aes(x = reorder(sample, lib_size), y = lib_size / 1e6,
             fill = severity)) +
  geom_col() +
  scale_fill_manual(values = c("Healthy_Mild" = "steelblue",
                                "Severe_ICU" = "firebrick")) +
  labs(title = "Library sizes (GSE152418)",
       x = "Sample", y = "Library size (millions of reads)",
       fill = "Severity") +
  theme_classic(base_size = 12) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 7))

print(p)
ggsave("plots/phase7/phase7_library_sizes.png", p, width = 10, height = 5, dpi = 300)
```

---

## 13. Save cleaned data for Phase 8

```{r p7-save-data}
saveRDS(counts_filtered, file = "Data/phase7_bulk_counts_filtered.rds")
saveRDS(metadata_clean, file = "Data/phase7_bulk_metadata.rds")

cat("Saved:\n")
cat("  Data/phase7_bulk_counts_filtered.rds\n")
cat("  Data/phase7_bulk_metadata.rds\n")
```

---

## Decision Checkpoint 3

> **Data quality and group definition:**
>
> 1. Why group Severe + ICU as one class?
> 2. Why exclude Convalescent samples?
> 3. Are the group sizes sufficient for DESeq2?
>
> (Fill in after rendering.)

---

## Session info

```{r p7-session-info}
sessionInfo()
```

---

# ====================================================================
# PHASE8 - DESEQ2 DIFFERENTIAL EXPRESSION
# ====================================================================


## Goal

This document implements **Phase 8** — differential expression analysis on
the GSE152418 bulk RNA-seq dataset using DESeq2. We compare **Severe/ICU** vs
**Healthy/Mild** to identify genes differentially expressed in severe COVID-19
at the whole-blood level.

---

## 1. Setup

```{r p8-p8-init, message=FALSE, warning=FALSE}
dir.create("plots/phase8", showWarnings = FALSE, recursive = TRUE)
```

---

## 2. Load Phase 7 outputs

```{r p8-load-data}
counts   <- readRDS("Data/phase7_bulk_counts_filtered.rds")
metadata <- readRDS("Data/phase7_bulk_metadata.rds")

cat("Count matrix:", nrow(counts), "genes x", ncol(counts), "samples\n")
cat("Metadata:", nrow(metadata), "samples\n")
cat("\nSeverity group distribution:\n")
print(table(metadata$severity_group))

# Ensure reference level
metadata$severity_group <- relevel(factor(metadata$severity_group),
                                    ref = "Healthy_Mild")
```

---

## 3. Create DESeqDataSet

```{r p8-create-dds}
dds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData   = metadata,
  design    = ~ severity_group
)

cat("DESeqDataSet created:\n")
cat("  Genes:", nrow(dds), "\n")
cat("  Samples:", ncol(dds), "\n")
```

---

## 4. Library size barplot

```{r p8-library-sizes, fig.width=10, fig.height=5}
lib_df <- data.frame(
  sample   = colnames(dds),
  lib_size = colSums(counts(dds)),
  severity = dds$severity_group
)

p <- ggplot(lib_df, aes(x = reorder(sample, lib_size), y = lib_size / 1e6,
                        fill = severity)) +
  geom_col() +
  scale_fill_manual(values = c("Healthy_Mild" = "steelblue",
                                "Severe_ICU" = "firebrick")) +
  labs(title = "Library sizes by severity group",
       x = "Sample", y = "Library size (millions)", fill = "Severity") +
  theme_classic(base_size = 12) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 7))

print(p)
ggsave("plots/phase8/phase8_library_sizes.png", p, width = 10, height = 5, dpi = 300)
```

---

## 5. Estimate size factors

```{r p8-size-factors}
dds <- estimateSizeFactors(dds)

cat("Size factors:\n")
sf_df <- data.frame(
  sample      = colnames(dds),
  size_factor = sizeFactors(dds),
  severity    = dds$severity_group
)
print(sf_df)

cat("\nSize factor range:", round(range(sizeFactors(dds)), 3), "\n")
```

---

## 6. Mean-variance relationship

```{r p8-mean-var, fig.width=8, fig.height=6}
# Compute mean and variance per gene
gene_means <- rowMeans(counts(dds))
gene_vars  <- apply(counts(dds), 1, var)

mv_df <- data.frame(mean = gene_means, variance = gene_vars)

p <- ggplot(mv_df, aes(x = log10(mean + 1), y = log10(variance + 1))) +
  geom_point(size = 0.3, alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Mean-variance relationship (raw counts)",
       subtitle = "Points above red line = overdispersed (variance > mean)",
       x = "log10(mean + 1)", y = "log10(variance + 1)") +
  theme_classic(base_size = 13)

print(p)
ggsave("plots/phase8/phase8_mean_variance.png", p, width = 8, height = 6, dpi = 300)
```

---

## 7. Run DESeq2

```{r p8-run-deseq}
dds <- DESeq(dds)

cat("DESeq2 complete.\n")
cat("Result names:", resultsNames(dds), "\n")
```

---

## 8. Extract results

```{r p8-extract-results}
res <- results(dds)

cat("Results summary (unshrunken):\n")
summary(res)
```

---

## 9. Shrink fold changes (apeglm)

```{r p8-shrink-lfc}
# Identify the correct coefficient name
coef_name <- grep("Severe", resultsNames(dds), value = TRUE)
cat("Shrinking coefficient:", coef_name, "\n")

resLFC <- lfcShrink(dds, coef = coef_name, type = "apeglm")

cat("\nResults summary (shrunken):\n")
summary(resLFC)
```

---

## 10. MA plots: before and after shrinkage

```{r p8-ma-plots, fig.width=12, fig.height=5}
par(mfrow = c(1, 2))

plotMA(res, main = "MA plot — unshrunken", ylim = c(-5, 5))
plotMA(resLFC, main = "MA plot — apeglm shrunken", ylim = c(-5, 5))

par(mfrow = c(1, 1))

# Save as ggplot-compatible version
png("plots/phase8/phase8_MA_plots.png", width = 12, height = 5,
    units = "in", res = 300)
par(mfrow = c(1, 2))
plotMA(res, main = "MA plot — unshrunken", ylim = c(-5, 5))
plotMA(resLFC, main = "MA plot — apeglm shrunken", ylim = c(-5, 5))
par(mfrow = c(1, 1))
dev.off()
```

---

## 11. Map gene symbols

```{r p8-map-symbols}
# Determine key type from row names
example_genes <- head(rownames(resLFC))
cat("Example gene IDs:", paste(example_genes, collapse = ", "), "\n")

# Check if ENSEMBL or SYMBOL
if (any(grepl("^ENSG", example_genes))) {
  cat("Gene IDs are Ensembl → mapping to symbols\n")
  resLFC$symbol <- mapIds(org.Hs.eg.db,
                          keys = rownames(resLFC),
                          column = "SYMBOL",
                          keytype = "ENSEMBL",
                          multiVals = "first")
  resLFC$ensembl <- rownames(resLFC)
} else {
  cat("Gene IDs appear to be symbols already\n")
  resLFC$symbol <- rownames(resLFC)
}

# How many mapped?
cat("Genes with symbols:", sum(!is.na(resLFC$symbol)), "/", nrow(resLFC), "\n")
```

---

## 12. Order by significance and count DE genes

```{r p8-order-results}
resOrdered <- resLFC[order(resLFC$pvalue), ]

cat("Significant genes (padj < 0.05):",
    sum(resOrdered$padj < 0.05, na.rm = TRUE), "\n")
cat("Upregulated (padj < 0.05, LFC > 0):",
    sum(resOrdered$padj < 0.05 & resOrdered$log2FoldChange > 0, na.rm = TRUE), "\n")
cat("Downregulated (padj < 0.05, LFC < 0):",
    sum(resOrdered$padj < 0.05 & resOrdered$log2FoldChange < 0, na.rm = TRUE), "\n")

cat("\nTop 20 DE genes:\n")
top20 <- head(as.data.frame(resOrdered), 20)
top20$symbol <- resLFC[rownames(top20), "symbol"]
print(top20[, c("symbol", "log2FoldChange", "padj")])
```

---

## 13. Volcano plot

```{r p8-volcano, fig.width=10, fig.height=8}
# Use symbols for labels if available
labs <- if (!is.null(resLFC$symbol)) resLFC$symbol else rownames(resLFC)
labs[is.na(labs)] <- ""

p <- EnhancedVolcano(resLFC,
                     lab = labs,
                     x   = "log2FoldChange",
                     y   = "padj",
                     title    = "Severe/ICU vs Healthy/Mild — Bulk RNA-seq (GSE152418)",
                     subtitle = "DESeq2 with apeglm shrinkage",
                     pCutoff  = 0.05,
                     FCcutoff = 1.0,
                     pointSize = 1.5,
                     labSize   = 3.5,
                     drawConnectors = TRUE,
                     widthConnectors = 0.3,
                     maxoverlapsConnectors = 20)

print(p)
ggsave("plots/phase8/phase8_volcano.png", p, width = 10, height = 8, dpi = 300)
```

---

## 14. Heatmap of top DE genes

```{r p8-heatmap, fig.width=10, fig.height=8}
# VST for visualization
vsd <- vst(dds, blind = FALSE)

# Select top 30 DE genes by adjusted p-value
top_genes <- head(rownames(resOrdered[!is.na(resOrdered$padj), ]), 30)

# Annotation column
anno_col <- data.frame(
  Severity = metadata$severity_group,
  row.names = colnames(vsd)
)

anno_colors <- list(
  Severity = c("Healthy_Mild" = "steelblue", "Severe_ICU" = "firebrick")
)

# Use symbols as row labels if available
row_labels <- resLFC[top_genes, "symbol"]
row_labels[is.na(row_labels)] <- top_genes[is.na(row_labels)]

mat <- assay(vsd)[top_genes, ]
rownames(mat) <- row_labels

p <- pheatmap(mat,
              scale = "row",
              annotation_col = anno_col,
              annotation_colors = anno_colors,
              show_colnames = FALSE,
              fontsize_row = 8,
              clustering_distance_cols = "euclidean",
              clustering_method = "ward.D2",
              main = "Top 30 DE genes — Severe/ICU vs Healthy/Mild")

ggsave("plots/phase8/phase8_heatmap_top30.png", p, width = 10, height = 8, dpi = 300)
```

---

## 15. Box/jitter plots for key genes

```{r p8-key-gene-plots, fig.width=14, fig.height=5}
# Key genes from the monocyte hypothesis
key_genes_symbols <- c("ISG15", "IL1B", "TNF", "IFIT1", "CCL3", "MX1")

# Find matching gene IDs
if (any(grepl("^ENSG", rownames(dds)))) {
  key_gene_ids <- mapIds(org.Hs.eg.db,
                         keys = key_genes_symbols,
                         column = "ENSEMBL",
                         keytype = "SYMBOL",
                         multiVals = "first")
  key_gene_ids <- key_gene_ids[key_gene_ids %in% rownames(dds)]
} else {
  key_gene_ids <- key_genes_symbols[key_genes_symbols %in% rownames(dds)]
  names(key_gene_ids) <- key_gene_ids
}

cat("Key genes found in data:", paste(names(key_gene_ids), collapse = ", "), "\n")

# Plot each
plots <- list()
for (i in seq_along(key_gene_ids)) {
  gene_id <- key_gene_ids[i]
  gene_name <- names(key_gene_ids)[i]

  d <- plotCounts(dds, gene = gene_id, intgroup = "severity_group",
                  returnData = TRUE)

  plots[[i]] <- ggplot(d, aes(x = severity_group, y = count, fill = severity_group)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +
    geom_jitter(width = 0.15, size = 2) +
    scale_y_log10() +
    scale_fill_manual(values = c("Healthy_Mild" = "steelblue",
                                  "Severe_ICU" = "firebrick")) +
    labs(title = gene_name, x = "", y = "Normalized count") +
    theme_classic(base_size = 11) +
    theme(legend.position = "none")
}

p <- patchwork::wrap_plots(plots, ncol = length(plots))
print(p)
ggsave("plots/phase8/phase8_key_gene_boxplots.png", p,
       width = 14, height = 5, dpi = 300)
```

---

## 16. Save results

```{r p8-save-results}
# Save full DESeq2 results
write.csv(as.data.frame(resOrdered),
          file = "Data/phase8_SevereICU_vs_HealthyMild_DESeq2.csv")

# Save the DESeqDataSet for downstream phases
saveRDS(dds, file = "Data/phase8_dds.rds")

# Save shrunken results
saveRDS(resLFC, file = "Data/phase8_resLFC.rds")

cat("Saved:\n")
cat("  Data/phase8_SevereICU_vs_HealthyMild_DESeq2.csv\n")
cat("  Data/phase8_dds.rds\n")
cat("  Data/phase8_resLFC.rds\n")
```

---

## Decision Checkpoint 4

> **Why DESeq2 for count data?**
>
> 1. DESeq2 models count data with a negative binomial distribution, which
>    properly handles the mean-variance relationship (overdispersion) in RNA-seq.
> 2. With small sample sizes, DESeq2's shrinkage estimators for dispersion
>    and fold changes are more conservative than limma-voom, reducing false
>    positives.
>
> **Why group Severe + ICU together?**
>
> 1.
> 2.
>
> (Fill in after rendering with actual group sizes.)

---

## Session info

```{r p8-session-info}
sessionInfo()
```

---

# ====================================================================
# PHASE9 - ADJUST CONFOUNDERS
# ====================================================================


## Goal

This document implements **Phase 9** — testing whether gender confounds the
Severe/ICU vs Healthy/Mild comparison in the GSE152418 bulk RNA-seq data.
We fit an additive DESeq2 model with gender and compare the results to the
simple model from Phase 8.

---

## 1. Setup

```{r p9-p9-init, message=FALSE, warning=FALSE}
dir.create("plots/phase9", showWarnings = FALSE, recursive = TRUE)
```

---

## 2. Load Phase 7–8 outputs

```{r p9-load-data}
counts   <- readRDS("Data/phase7_bulk_counts_filtered.rds")
metadata <- readRDS("Data/phase7_bulk_metadata.rds")
dds_simple <- readRDS("Data/phase8_dds.rds")
resLFC     <- readRDS("Data/phase8_resLFC.rds")

cat("Count matrix:", nrow(counts), "genes x", ncol(counts), "samples\n")
cat("Severity groups:\n")
print(table(metadata$severity_group))
```

---

## 3. Check gender balance

```{r p9-gender-balance}
cat("Gender distribution:\n")
print(table(metadata$gender))

cat("\nGender × Severity group:\n")
gender_sev <- table(metadata$gender, metadata$severity_group)
print(gender_sev)

cat("\nChi-squared test for gender × severity independence:\n")
if (min(dim(gender_sev)) >= 2) {
  chi_test <- chisq.test(gender_sev)
  print(chi_test)
  cat("\np-value:", chi_test$p.value, "\n")
  if (chi_test$p.value < 0.05) {
    cat("→ Gender is significantly associated with severity group — confounding possible.\n")
  } else {
    cat("→ No significant association — gender is reasonably balanced.\n")
  }
} else {
  cat("Cannot test — not enough levels.\n")
}
```

### Gender balance plot

```{r p9-gender-plot, fig.width=8, fig.height=5}
p <- ggplot(metadata, aes(x = severity_group, fill = gender)) +
  geom_bar(position = "dodge") +
  labs(title = "Gender balance across severity groups",
       x = "Severity group", y = "Count", fill = "Gender") +
  theme_classic(base_size = 14)

print(p)
ggsave("plots/phase9/phase9_gender_balance.png", p, width = 8, height = 5, dpi = 300)
```

---

## 4. Fit additive model: ~ gender + severity_group

```{r p9-fit-adjusted}
# Ensure gender is a factor
metadata$gender <- factor(metadata$gender)

# Ensure severity_group reference level
metadata$severity_group <- relevel(factor(metadata$severity_group),
                                    ref = "Healthy_Mild")

dds2 <- DESeqDataSetFromMatrix(
  countData = counts,
  colData   = metadata,
  design    = ~ gender + severity_group
)

dds2 <- DESeq(dds2)

cat("Adjusted model result names:\n")
print(resultsNames(dds2))
```

---

## 5. Shrink fold changes (adjusted model)

```{r p9-shrink-adjusted}
coef_name <- grep("Severe", resultsNames(dds2), value = TRUE)
cat("Shrinking coefficient:", coef_name, "\n")

res2LFC <- lfcShrink(dds2, coef = coef_name, type = "apeglm")

cat("\nAdjusted model results summary:\n")
summary(res2LFC)
```

---

## 6. Compare simple vs adjusted models

### 6a. Fold-change correlation

```{r p9-compare-lfc, fig.width=8, fig.height=8}
# Align genes
shared_genes <- intersect(rownames(resLFC), rownames(res2LFC))
lfc_simple   <- resLFC[shared_genes, "log2FoldChange"]
lfc_adjusted <- res2LFC[shared_genes, "log2FoldChange"]

lfc_cor <- cor(lfc_simple, lfc_adjusted, use = "complete.obs")
cat("Pearson correlation of log2FC (simple vs adjusted):", round(lfc_cor, 4), "\n")

df_compare <- data.frame(
  simple   = lfc_simple,
  adjusted = lfc_adjusted
)

p <- ggplot(df_compare, aes(x = simple, y = adjusted)) +
  geom_point(size = 0.5, alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = paste0("Simple vs Gender-Adjusted log2FC (r = ",
                      round(lfc_cor, 4), ")"),
       x = "log2FC — simple model (~ severity_group)",
       y = "log2FC — adjusted model (~ gender + severity_group)") +
  theme_classic(base_size = 13) +
  coord_equal()

print(p)
ggsave("plots/phase9/phase9_lfc_correlation.png", p, width = 8, height = 8, dpi = 300)
```

### 6b. Significance comparison

```{r p9-compare-sig}
sig_simple   <- sum(resLFC$padj < 0.05, na.rm = TRUE)
sig_adjusted <- sum(res2LFC$padj < 0.05, na.rm = TRUE)

cat("Significant genes (padj < 0.05):\n")
cat("  Simple model:   ", sig_simple, "\n")
cat("  Adjusted model: ", sig_adjusted, "\n")
cat("  Difference:     ", sig_adjusted - sig_simple, "\n")

# Overlap
sig_genes_simple   <- rownames(resLFC)[which(resLFC$padj < 0.05)]
sig_genes_adjusted <- rownames(res2LFC)[which(res2LFC$padj < 0.05)]
overlap <- length(intersect(sig_genes_simple, sig_genes_adjusted))
cat("\n  Overlap:         ", overlap, "\n")
cat("  Only in simple:  ", length(setdiff(sig_genes_simple, sig_genes_adjusted)), "\n")
cat("  Only in adjusted:", length(setdiff(sig_genes_adjusted, sig_genes_simple)), "\n")
```

### 6c. MA plot comparison

```{r p9-ma-comparison, fig.width=12, fig.height=5}
png("plots/phase9/phase9_MA_comparison.png", width = 12, height = 5,
    units = "in", res = 300)
par(mfrow = c(1, 2))
plotMA(resLFC, main = "Simple model (~ severity_group)", ylim = c(-5, 5))
plotMA(res2LFC, main = "Adjusted model (~ gender + severity_group)", ylim = c(-5, 5))
par(mfrow = c(1, 1))
dev.off()

# Also display in document
par(mfrow = c(1, 2))
plotMA(resLFC, main = "Simple model", ylim = c(-5, 5))
plotMA(res2LFC, main = "Adjusted model", ylim = c(-5, 5))
par(mfrow = c(1, 1))
```

---

## 7. Check gender effect specifically

```{r p9-gender-effect}
# Extract gender effect
gender_coef <- grep("gender", resultsNames(dds2), value = TRUE)
cat("Gender coefficient:", gender_coef, "\n")

res_gender <- results(dds2, name = gender_coef)
cat("\nGender effect summary:\n")
summary(res_gender)

sig_gender <- sum(res_gender$padj < 0.05, na.rm = TRUE)
cat("\nGenes significantly associated with gender (padj < 0.05):", sig_gender, "\n")
```

---

## 8. Decision: simple or adjusted model?

```{r p9-decision}
cat("===== MODEL COMPARISON SUMMARY =====\n\n")
cat("Correlation of fold changes: r =", round(lfc_cor, 4), "\n")
cat("Significant genes — simple:  ", sig_simple, "\n")
cat("Significant genes — adjusted:", sig_adjusted, "\n")
cat("Overlap:                     ", overlap, "\n")
cat("Gender-associated genes:     ", sig_gender, "\n\n")

if (lfc_cor > 0.99 && abs(sig_adjusted - sig_simple) < sig_simple * 0.1) {
  cat("RECOMMENDATION: Use the SIMPLE model.\n")
  cat("Rationale: Fold changes are nearly identical (r > 0.99).\n")
  cat("Adding gender as a covariate does not materially change results\n")
  cat("and may cost statistical power with small sample sizes.\n")
} else if (lfc_cor > 0.95) {
  cat("RECOMMENDATION: Models are similar. Either is defensible.\n")
  cat("Consider sample size when deciding whether the extra covariate is justified.\n")
} else {
  cat("RECOMMENDATION: Use the ADJUSTED model.\n")
  cat("Rationale: Gender has a meaningful confounding effect on fold changes.\n")
}
```

---

## 9. Save results

```{r p9-save-results}
saveRDS(dds2, file = "Data/phase9_dds_adjusted.rds")
saveRDS(res2LFC, file = "Data/phase9_resLFC_adjusted.rds")

cat("Saved:\n")
cat("  Data/phase9_dds_adjusted.rds\n")
cat("  Data/phase9_resLFC_adjusted.rds\n")
```

---

## Decision Checkpoint 5

> **Simple vs adjusted model:**
>
> 1. Correlation between fold changes: r = ___
> 2. Gender confounding assessment: ___
> 3. Final decision: ___
>
> (Fill in after rendering.)

---

## Session info

```{r p9-session-info}
sessionInfo()
```

---

# ====================================================================
# PHASE10 - BULK VISUALIZATION
# ====================================================================


## Goal

This document implements **Phase 10** — a comprehensive visualization summary
of the bulk RNA-seq differential expression results from Phases 8 and 9.
All plots are saved to `plots/phase10/`. We produce:

1. Volcano plot (Severe/ICU vs Healthy/Mild)
2. Heatmap of top DE genes (VST-normalized)
3. Box / jitter plots for key hypothesis genes (ISG + TNF/IL-1 pathway)
4. Count distribution for selected genes across severity groups
5. Summary table of top DE genes

---

## 1. Setup

```{r p10-p10-init, message=FALSE, warning=FALSE}
dir.create("plots/phase10", showWarnings = FALSE, recursive = TRUE)
```

---

## 2. Load Phase 8–9 outputs

```{r p10-load-data}
dds        <- readRDS("Data/phase8_dds.rds")
resLFC     <- readRDS("Data/phase8_resLFC.rds")
metadata   <- readRDS("Data/phase7_bulk_metadata.rds")

# Also load adjusted model results for comparison, if available
dds2 <- tryCatch(readRDS("Data/phase9_dds_adjusted.rds"), error = function(e) NULL)
res2LFC <- tryCatch(readRDS("Data/phase9_resLFC_adjusted.rds"), error = function(e) NULL)

cat("Loaded DESeqDataSet:", nrow(dds), "genes x", ncol(dds), "samples\n")
cat("Severity groups:\n")
print(table(metadata$severity_group))
```

---

## 3. Order results by significance

```{r p10-order-results}
resOrdered <- resLFC[order(resLFC$pvalue), ]

cat("Significant genes (padj < 0.05):",
    sum(resOrdered$padj < 0.05, na.rm = TRUE), "\n")
cat("Upregulated   (padj < 0.05, LFC > 1):",
    sum(resOrdered$padj < 0.05 & resOrdered$log2FoldChange >  1, na.rm = TRUE), "\n")
cat("Downregulated (padj < 0.05, LFC < -1):",
    sum(resOrdered$padj < 0.05 & resOrdered$log2FoldChange < -1, na.rm = TRUE), "\n")

# Build labels (symbol if mapped, gene ID otherwise)
labs <- resLFC$symbol
labs[is.na(labs)] <- rownames(resLFC)[is.na(labs)]
```

---

## 4. Volcano plot

```{r p10-volcano, fig.width=12, fig.height=9}
p_volcano <- EnhancedVolcano(
  resLFC,
  lab      = labs,
  x        = "log2FoldChange",
  y        = "padj",
  title    = "Severe/ICU vs Healthy/Mild — Bulk RNA-seq (GSE152418)",
  subtitle = "DESeq2 + apeglm shrinkage  |  red: padj < 0.05 & |LFC| > 1",
  pCutoff  = 0.05,
  FCcutoff = 1.0,
  pointSize         = 2.0,
  labSize           = 3.5,
  drawConnectors    = TRUE,
  widthConnectors   = 0.3,
  maxoverlapsConnectors = 25,
  colAlpha           = 0.6,
  legendPosition     = "right"
)

print(p_volcano)
ggsave("plots/phase10/phase10_volcano.png",
       p_volcano, width = 12, height = 9, dpi = 300)
```

---

## 5. Heatmap of top 50 DE genes

```{r p10-heatmap-top50, fig.width=12, fig.height=14}
# VST transform for visualization
vsd <- vst(dds, blind = FALSE)

# Select top 50 by adjusted p-value (exclude NAs)
top_genes <- rownames(resOrdered[!is.na(resOrdered$padj), ])[1:50]

# Sample annotation
anno_col <- data.frame(
  Severity = metadata$severity_group,
  row.names = colnames(vsd)
)

# Add gender annotation if available
if ("gender" %in% colnames(metadata)) {
  anno_col$Gender <- metadata$gender
}

anno_colors <- list(
  Severity = c("Healthy_Mild" = "steelblue", "Severe_ICU" = "firebrick")
)
if ("Gender" %in% colnames(anno_col)) {
  genders <- unique(na.omit(anno_col$Gender))
  pal <- setNames(RColorBrewer::brewer.pal(max(2, length(genders)), "Set1")[seq_along(genders)],
                  genders)
  anno_colors$Gender <- pal
}

# Use gene symbols as row labels where available
row_labels <- resLFC[top_genes, "symbol"]
row_labels[is.na(row_labels)] <- top_genes[is.na(row_labels)]

mat <- assay(vsd)[top_genes, ]
rownames(mat) <- row_labels

# Order columns by severity group
col_order <- order(anno_col$Severity)

p_heatmap <- pheatmap(
  mat[, col_order],
  scale                   = "row",
  annotation_col          = anno_col[col_order, , drop = FALSE],
  annotation_colors       = anno_colors,
  show_colnames           = FALSE,
  fontsize_row            = 8,
  fontsize                = 10,
  clustering_distance_cols = "euclidean",
  clustering_method       = "ward.D2",
  cluster_cols            = FALSE,
  color                   = colorRampPalette(rev(brewer.pal(11, "RdBu")))(100),
  main = "Top 50 DE genes — Severe/ICU vs Healthy/Mild (VST, row-scaled)"
)

ggsave("plots/phase10/phase10_heatmap_top50.png",
       p_heatmap, width = 12, height = 14, dpi = 300)
```

---

## 6. Box / jitter plots for key hypothesis genes

These are the ISG and TNF/IL-1 module genes central to the project hypothesis.

```{r p10-key-gene-boxplots, fig.width=16, fig.height=10}
# ISG module genes
isg_genes <- c("ISG15", "IFIT1", "MX1", "OAS1", "IFI44L", "IFIT3")
# TNF / IL-1 module genes
tnf_genes <- c("TNF", "IL1B", "CCL3", "CCL4", "CXCL8", "IL6")

all_target_symbols <- c(isg_genes, tnf_genes)

# Resolve gene IDs (Ensembl or symbol)
example_id <- rownames(dds)[1]
use_ensembl <- grepl("^ENSG", example_id)

if (use_ensembl) {
  id_map <- mapIds(
    org.Hs.eg.db,
    keys    = all_target_symbols,
    column  = "ENSEMBL",
    keytype = "SYMBOL",
    multiVals = "first"
  )
} else {
  id_map <- setNames(all_target_symbols, all_target_symbols)
}

# Keep only genes present in the dds
id_map <- id_map[!is.na(id_map) & id_map %in% rownames(dds)]
cat("Key genes found in data:", paste(names(id_map), collapse = ", "), "\n")

make_count_plot <- function(gene_id, gene_name, group_var = "severity_group") {
  d <- plotCounts(dds, gene = gene_id, intgroup = group_var, returnData = TRUE)
  ggplot(d, aes(x = .data[[group_var]], y = count, fill = .data[[group_var]])) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7, width = 0.55) +
    geom_jitter(width = 0.15, size = 2.0, alpha = 0.85) +
    scale_y_log10() +
    scale_fill_manual(values = c("Healthy_Mild" = "steelblue",
                                  "Severe_ICU"   = "firebrick")) +
    labs(title = gene_name, x = "", y = "Normalised count (log10)") +
    theme_classic(base_size = 11) +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 30, hjust = 1))
}

plots_isg <- lapply(
  names(id_map)[names(id_map) %in% isg_genes],
  function(sym) make_count_plot(id_map[[sym]], sym)
)

plots_tnf <- lapply(
  names(id_map)[names(id_map) %in% tnf_genes],
  function(sym) make_count_plot(id_map[[sym]], sym)
)

all_plots <- c(plots_isg, plots_tnf)

if (length(all_plots) > 0) {
  p_key <- patchwork::wrap_plots(all_plots,
                                  ncol = min(6, length(all_plots))) +
    patchwork::plot_annotation(
      title   = "Key ISG and TNF/IL-1 genes — normalised counts by severity",
      subtitle = "Box + jitter; y-axis log10-scaled"
    )
  print(p_key)
  ggsave("plots/phase10/phase10_key_gene_boxplots.png",
         p_key, width = 16, height = 10, dpi = 300)
} else {
  message("No key genes found in the dataset. Skipping plot.")
}
```

---

## 7. ISG panel and TNF/IL-1 panel separately

```{r p10-isg-panel, fig.width=14, fig.height=5}
if (length(plots_isg) > 0) {
  p_isg <- patchwork::wrap_plots(plots_isg, ncol = length(plots_isg)) +
    patchwork::plot_annotation(
      title    = "ISG module genes — Severe/ICU vs Healthy/Mild",
      subtitle = "ISG15, IFIT1, MX1, OAS1, IFI44L, IFIT3"
    )
  print(p_isg)
  ggsave("plots/phase10/phase10_ISG_panel.png",
         p_isg, width = 14, height = 5, dpi = 300)
}
```

```{r p10-tnf-panel, fig.width=14, fig.height=5}
if (length(plots_tnf) > 0) {
  p_tnf <- patchwork::wrap_plots(plots_tnf, ncol = length(plots_tnf)) +
    patchwork::plot_annotation(
      title    = "TNF/IL-1 module genes — Severe/ICU vs Healthy/Mild",
      subtitle = "TNF, IL1B, CCL3, CCL4, CXCL8, IL6"
    )
  print(p_tnf)
  ggsave("plots/phase10/phase10_TNF_IL1_panel.png",
         p_tnf, width = 14, height = 5, dpi = 300)
}
```

---

## 8. Summary table of top 20 DE genes

```{r p10-top-de-table}
top20_df <- as.data.frame(resOrdered[1:20, ])
top20_df$symbol <- resLFC[rownames(top20_df), "symbol"]
top20_df$gene_id <- rownames(top20_df)

top20_clean <- top20_df %>%
  dplyr::select(symbol, gene_id, log2FoldChange, lfcSE, pvalue, padj) %>%
  dplyr::mutate(
    log2FoldChange = round(log2FoldChange, 3),
    lfcSE          = round(lfcSE, 3),
    pvalue         = formatC(pvalue, format = "e", digits = 2),
    padj           = formatC(padj, format = "e", digits = 2)
  )

cat("Top 20 differentially expressed genes (Severe/ICU vs Healthy/Mild):\n")
print(top20_clean, row.names = FALSE)

# Save as CSV
write.csv(top20_clean,
          file = "Data/phase10_top20_DE_genes.csv",
          row.names = FALSE)
cat("\nSaved: Data/phase10_top20_DE_genes.csv\n")
```

---

## 9. Comparison: simple vs gender-adjusted model (if Phase 9 ran)

```{r p10-model-comparison, fig.width=14, fig.height=5}
if (!is.null(res2LFC)) {
  cat("Phase 9 results loaded — producing model comparison plot.\n")

  shared <- intersect(rownames(resLFC), rownames(res2LFC))

  df_cmp <- data.frame(
    simple   = resLFC[shared,   "log2FoldChange"],
    adjusted = res2LFC[shared,  "log2FoldChange"],
    sig_simple   = !is.na(resLFC[shared, "padj"]) & resLFC[shared, "padj"] < 0.05,
    sig_adjusted = !is.na(res2LFC[shared, "padj"]) & res2LFC[shared, "padj"] < 0.05
  )

  r_val <- round(cor(df_cmp$simple, df_cmp$adjusted, use = "complete.obs"), 4)

  p_cmp <- ggplot(df_cmp, aes(x = simple, y = adjusted)) +
    geom_point(size = 0.4, alpha = 0.25, color = "grey40") +
    geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", linewidth = 0.8) +
    labs(
      title    = paste0("Simple vs Gender-Adjusted log2FC (r = ", r_val, ")"),
      subtitle = "Each point = one gene; diagonal = perfect agreement",
      x        = "log2FC — simple model (~ severity_group)",
      y        = "log2FC — adjusted model (~ gender + severity_group)"
    ) +
    theme_classic(base_size = 13) +
    coord_equal()

  print(p_cmp)
  ggsave("plots/phase10/phase10_simple_vs_adjusted_lfc.png",
         p_cmp, width = 8, height = 8, dpi = 300)

  # MA plots side by side
  png("plots/phase10/phase10_MA_comparison.png",
      width = 14, height = 5, units = "in", res = 300)
  par(mfrow = c(1, 2))
  plotMA(resLFC,  main = "Simple model (~ severity_group)",          ylim = c(-5, 5))
  plotMA(res2LFC, main = "Adjusted model (~ gender + severity_group)", ylim = c(-5, 5))
  par(mfrow = c(1, 1))
  dev.off()

  par(mfrow = c(1, 2))
  plotMA(resLFC,  main = "Simple model",   ylim = c(-5, 5))
  plotMA(res2LFC, main = "Adjusted model", ylim = c(-5, 5))
  par(mfrow = c(1, 1))
} else {
  cat("Phase 9 results not available — skipping model comparison.\n")
}
```

---

## 10. Save Phase 10 outputs

```{r p10-save-outputs}
# Save the VST matrix for downstream use in Phase 11
vst_mat <- assay(vsd)
saveRDS(vst_mat, file = "Data/phase10_vst_matrix.rds")

cat("Saved:\n")
cat("  Data/phase10_vst_matrix.rds\n")
cat("  Data/phase10_top20_DE_genes.csv\n")
cat("\nPlots saved in: plots/phase10/\n")
list.files("plots/phase10/")
```

---

## Session info

```{r p10-session-info}
sessionInfo()
```

---

# ====================================================================
# PHASE11 - PCA CLUSTERING BULK
# ====================================================================


## Goal

This document implements **Phase 11** — unsupervised analysis of the
GSE152418 bulk RNA-seq dataset using VST-normalized counts. We perform:

1. VST normalization
2. Top-variable gene selection
3. PCA (2D scatter + variance-explained bar chart)
4. Sample correlation matrix heatmap
5. Euclidean distance matrix heatmap
6. Hierarchical clustering (ward.D2)
7. K-means clustering + elbow method
8. Validation of clusters vs known severity groups

All plots are saved to `plots/phase11/`.

---

## 1. Setup

```{r p11-p11-init, message=FALSE, warning=FALSE}
dir.create("plots/phase11", showWarnings = FALSE, recursive = TRUE)
```

---

## 2. Load data

```{r p11-load-data}
dds      <- readRDS("Data/phase8_dds.rds")
metadata <- readRDS("Data/phase7_bulk_metadata.rds")

cat("DESeqDataSet:", nrow(dds), "genes x", ncol(dds), "samples\n")
cat("\nSeverity groups:\n")
print(table(metadata$severity_group))

if ("gender" %in% colnames(metadata)) {
  cat("\nGender distribution:\n")
  print(table(metadata$gender))
}
```

---

## 3. VST normalization

Variance stabilizing transformation (VST) produces approximately
homoscedastic counts suitable for PCA and distance-based clustering.
We set `blind = TRUE` here so the transformation is not informed by
the experimental design (appropriate for exploratory/unsupervised analysis).

```{r p11-vst}
vsd <- vst(dds, blind = TRUE)
normcounts <- assay(vsd)

cat("VST matrix dimensions:", nrow(normcounts), "genes x", ncol(normcounts), "samples\n")
cat("Value range: [", round(min(normcounts), 2), ",",
    round(max(normcounts), 2), "]\n")
```

---

## 4. Select top variable genes

```{r p11-top-var-genes}
var_per_gene <- apply(normcounts, 1, var)

n_top_genes <- 1000
top_var_genes <- names(sort(var_per_gene, decreasing = TRUE))[1:n_top_genes]

normcounts_top <- normcounts[top_var_genes, ]

cat("Using top", n_top_genes, "most variable genes for unsupervised analysis.\n")
cat("Variance range of selected genes: [",
    round(min(var_per_gene[top_var_genes]), 3), ",",
    round(max(var_per_gene[top_var_genes]), 3), "]\n")

frac_var <- sum(var_per_gene[top_var_genes]) / sum(var_per_gene)
cat("Variance captured by top", n_top_genes, "genes:",
    round(100 * frac_var, 1), "%\n")
```

---

## 5. PCA

### 5a. Run PCA

```{r p11-run-pca}
pca_res <- prcomp(t(normcounts_top), scale. = TRUE)

percent_var <- round(100 * pca_res$sdev^2 / sum(pca_res$sdev^2), 2)

cat("Variance explained by top 10 PCs:\n")
for (i in 1:10) {
  cat(sprintf("  PC%d: %.2f%%\n", i, percent_var[i]))
}
cat("  Total (PC1-10):", round(sum(percent_var[1:10]), 1), "%\n")
```

### 5b. Variance-explained bar chart

```{r p11-pca-scree, fig.width=9, fig.height=5}
scree_df <- data.frame(
  PC      = factor(paste0("PC", 1:20), levels = paste0("PC", 1:20)),
  pct_var = percent_var[1:20],
  cumvar  = cumsum(percent_var[1:20])
)

p_scree <- ggplot(scree_df, aes(x = PC, y = pct_var)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_line(aes(y = cumvar, group = 1), color = "firebrick",
            linewidth = 0.8, linetype = "dashed") +
  geom_point(aes(y = cumvar), color = "firebrick", size = 2) +
  scale_y_continuous(
    name     = "Variance explained (%)",
    sec.axis = sec_axis(~ ., name = "Cumulative variance (%)")
  ) +
  labs(title   = "PCA - Variance explained per PC",
       subtitle = "Bars: individual; red dashed: cumulative",
       x = "Principal component") +
  theme_classic(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_scree)
ggsave("plots/phase11/phase11_PCA_scree.png",
       p_scree, width = 9, height = 5, dpi = 300)
```

### 5c. 2D PCA plot coloured by severity

```{r p11-pca-plot-severity, fig.width=9, fig.height=7}
pca_df <- data.frame(
  PC1      = pca_res$x[, 1],
  PC2      = pca_res$x[, 2],
  PC3      = pca_res$x[, 3],
  sample   = rownames(pca_res$x),
  severity = metadata$severity_group
)

if ("gender" %in% colnames(metadata)) {
  pca_df$gender <- metadata$gender
}

p_pca_sev <- ggplot(pca_df,
                    aes(x = PC1, y = PC2, color = severity, label = sample)) +
  geom_point(size = 4, alpha = 0.9) +
  geom_text(size = 2.5, vjust = -0.8, hjust = 0.5, color = "black") +
  scale_color_manual(values = c("Healthy_Mild" = "steelblue",
                                 "Severe_ICU"   = "firebrick")) +
  labs(
    title    = "PCA - Bulk RNA-seq (GSE152418)",
    subtitle = paste0("PC1: ", percent_var[1], "% | PC2: ", percent_var[2], "%"),
    x        = paste0("PC1 (", percent_var[1], "%)"),
    y        = paste0("PC2 (", percent_var[2], "%)"),
    color    = "Severity"
  ) +
  theme_classic(base_size = 13)

print(p_pca_sev)
ggsave("plots/phase11/phase11_PCA_severity.png",
       p_pca_sev, width = 9, height = 7, dpi = 300)
```

### 5d. 2D PCA plot coloured by severity, shaped by gender

```{r p11-pca-plot-gender, fig.width=9, fig.height=7}
if ("gender" %in% colnames(pca_df)) {
  p_pca_gen <- ggplot(pca_df,
                      aes(x = PC1, y = PC2,
                          color = severity, shape = gender, label = sample)) +
    geom_point(size = 4, alpha = 0.9) +
    geom_text(size = 2.5, vjust = -0.8, hjust = 0.5, color = "black") +
    scale_color_manual(values = c("Healthy_Mild" = "steelblue",
                                   "Severe_ICU"   = "firebrick")) +
    labs(
      title    = "PCA - coloured by severity, shaped by gender",
      subtitle = paste0("PC1: ", percent_var[1], "% | PC2: ", percent_var[2], "%"),
      x        = paste0("PC1 (", percent_var[1], "%)"),
      y        = paste0("PC2 (", percent_var[2], "%)"),
      color    = "Severity",
      shape    = "Gender"
    ) +
    theme_classic(base_size = 13)

  print(p_pca_gen)
  ggsave("plots/phase11/phase11_PCA_severity_gender.png",
         p_pca_gen, width = 9, height = 7, dpi = 300)
} else {
  cat("Gender metadata not available - skipping gender-shaped PCA plot.\n")
}
```

### 5e. PC1 vs PC3

```{r p11-pca-pc1-pc3, fig.width=9, fig.height=7}
p_pca_pc13 <- ggplot(pca_df,
                     aes(x = PC1, y = PC3, color = severity, label = sample)) +
  geom_point(size = 4, alpha = 0.9) +
  geom_text(size = 2.5, vjust = -0.8, hjust = 0.5, color = "black") +
  scale_color_manual(values = c("Healthy_Mild" = "steelblue",
                                 "Severe_ICU"   = "firebrick")) +
  labs(
    title    = "PCA - PC1 vs PC3",
    x        = paste0("PC1 (", percent_var[1], "%)"),
    y        = paste0("PC3 (", percent_var[3], "%)"),
    color    = "Severity"
  ) +
  theme_classic(base_size = 13)

print(p_pca_pc13)
ggsave("plots/phase11/phase11_PCA_PC1_PC3.png",
       p_pca_pc13, width = 9, height = 7, dpi = 300)
```

---

## 6. Sample correlation matrix heatmap

```{r p11-correlation-heatmap, fig.width=10, fig.height=9}
cor_mat <- cor(normcounts_top)

anno_col <- data.frame(
  Severity = metadata$severity_group,
  row.names = colnames(cor_mat)
)

if ("gender" %in% colnames(metadata)) {
  anno_col$Gender <- metadata$gender
}

anno_colors <- list(
  Severity = c("Healthy_Mild" = "steelblue", "Severe_ICU" = "firebrick")
)

if ("Gender" %in% colnames(anno_col)) {
  genders <- unique(na.omit(anno_col$Gender))
  pal_g <- setNames(
    RColorBrewer::brewer.pal(max(3, length(genders)), "Set1")[seq_along(genders)],
    genders
  )
  anno_colors$Gender <- pal_g
}

p_cor <- pheatmap(
  cor_mat,
  annotation_col    = anno_col,
  annotation_row    = anno_col,
  annotation_colors = anno_colors,
  color             = colorRampPalette(rev(brewer.pal(11, "RdBu")))(100),
  breaks            = seq(0.85, 1, length.out = 101),
  show_colnames     = FALSE,
  show_rownames     = FALSE,
  clustering_method = "ward.D2",
  main = "Sample-to-sample Pearson correlation (top 1,000 variable genes)"
)

ggsave("plots/phase11/phase11_correlation_matrix.png",
       p_cor, width = 10, height = 9, dpi = 300)
```

---

## 7. Euclidean distance matrix heatmap

```{r p11-distance-heatmap, fig.width=10, fig.height=9}
dist_mat   <- dist(t(normcounts_top), method = "euclidean")
dist_mat2  <- as.matrix(dist_mat)

p_dist <- pheatmap(
  dist_mat2,
  annotation_col    = anno_col,
  annotation_row    = anno_col,
  annotation_colors = anno_colors,
  color             = colorRampPalette(brewer.pal(9, "Blues"))(100),
  show_colnames     = FALSE,
  show_rownames     = FALSE,
  clustering_method = "ward.D2",
  main = "Sample-to-sample Euclidean distances (top 1,000 variable genes)"
)

ggsave("plots/phase11/phase11_distance_matrix.png",
       p_dist, width = 10, height = 9, dpi = 300)
```

---

## 8. Hierarchical clustering

```{r p11-hierarchical-clustering, fig.width=12, fig.height=6}
hc <- hclust(dist_mat, method = "ward.D2")

png("plots/phase11/phase11_hierarchical_clustering.png",
    width = 12, height = 6, units = "in", res = 300)
plot(hc, hang = -1,
     main = "Hierarchical clustering - ward.D2 (top 1,000 variable genes)",
     sub  = "",
     xlab = "Samples",
     ylab = "Euclidean distance",
     cex  = 0.85)
rect.hclust(hc, k = 2, border = c("steelblue", "firebrick"))
dev.off()

plot(hc, hang = -1,
     main = "Hierarchical clustering - ward.D2",
     sub  = "", xlab = "Samples", ylab = "Euclidean distance", cex = 0.85)
rect.hclust(hc, k = 2, border = c("steelblue", "firebrick"))
```

### 8a. Cluster correspondence with severity

```{r p11-hclust-vs-severity}
hc_clusters <- cutree(hc, k = 2)

cat("Hierarchical clustering k=2 vs severity group:\n")
hc_tab <- table(
  HC_cluster = hc_clusters,
  Severity   = metadata$severity_group[match(names(hc_clusters),
                                              rownames(metadata))]
)
print(hc_tab)

cluster_purity <- function(tab) apply(tab, 1, max) / rowSums(tab)
purities <- cluster_purity(hc_tab)
cat("\nCluster purity:\n")
for (i in seq_along(purities)) {
  cat(sprintf("  Cluster %d: %.1f%%\n", i, purities[i] * 100))
}
```

---

## 9. K-means clustering and elbow method

### 9a. Elbow plot (WSS)

```{r p11-kmeans-elbow, fig.width=9, fig.height=5}
set.seed(42)

wss_values <- sapply(1:8, function(k) {
  km <- kmeans(t(normcounts_top), centers = k, nstart = 25, iter.max = 100)
  km$tot.withinss
})

wss_df <- data.frame(k = 1:8, WSS = wss_values)

p_elbow <- ggplot(wss_df, aes(x = k, y = WSS)) +
  geom_line(linewidth = 1, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  geom_vline(xintercept = 2, linetype = "dashed",
             color = "firebrick", linewidth = 0.8) +
  annotate("text", x = 2.2, y = max(wss_values) * 0.95,
           label = "k = 2 (elbow)", hjust = 0, color = "firebrick") +
  scale_x_continuous(breaks = 1:8) +
  labs(title   = "K-means elbow plot - within-cluster sum of squares",
       subtitle = "Choose k at the elbow of the curve",
       x = "Number of clusters (k)", y = "Total within-cluster SS") +
  theme_classic(base_size = 13)

print(p_elbow)
ggsave("plots/phase11/phase11_kmeans_elbow.png",
       p_elbow, width = 9, height = 5, dpi = 300)
```

### 9b. K-means with k = 2

```{r p11-kmeans-k2}
set.seed(42)
km_k2 <- kmeans(t(normcounts_top), centers = 2, nstart = 25, iter.max = 100)

cat("K-means (k=2) cluster sizes:\n")
print(table(km_k2$cluster))

cat("\nK-means (k=2) vs severity group:\n")
km_tab <- table(
  KMeans_cluster = km_k2$cluster,
  Severity       = metadata$severity_group[
    match(names(km_k2$cluster), rownames(metadata))]
)
print(km_tab)

km_purities <- cluster_purity(km_tab)
cat("\nCluster purity:\n")
for (i in seq_along(km_purities)) {
  cat(sprintf("  Cluster %d: %.1f%%\n", i, km_purities[i] * 100))
}
```

### 9c. K-means clusters overlaid on PCA

```{r p11-kmeans-pca-plot, fig.width=10, fig.height=7}
pca_df$kmeans_cluster <- factor(km_k2$cluster[pca_df$sample])

p_km_pca <- ggplot(pca_df,
                   aes(x = PC1, y = PC2,
                       color  = severity,
                       shape  = kmeans_cluster,
                       label  = sample)) +
  geom_point(size = 4, alpha = 0.9) +
  geom_text(size = 2.5, vjust = -0.8, hjust = 0.5, color = "black") +
  scale_color_manual(values = c("Healthy_Mild" = "steelblue",
                                 "Severe_ICU"   = "firebrick")) +
  scale_shape_manual(values = c("1" = 16, "2" = 17)) +
  labs(
    title    = "PCA - K-means cluster (shape) vs known severity (colour)",
    subtitle = paste0("PC1: ", percent_var[1], "% | PC2: ", percent_var[2], "%"),
    x        = paste0("PC1 (", percent_var[1], "%)"),
    y        = paste0("PC2 (", percent_var[2], "%)"),
    color    = "Severity",
    shape    = "K-means cluster"
  ) +
  theme_classic(base_size = 13)

print(p_km_pca)
ggsave("plots/phase11/phase11_PCA_kmeans_overlay.png",
       p_km_pca, width = 10, height = 7, dpi = 300)
```

### 9d. factoextra cluster visualisation

```{r p11-fviz-cluster, fig.width=9, fig.height=7}
p_fviz <- fviz_cluster(
  km_k2,
  data         = t(normcounts_top),
  geom         = c("point", "text"),
  ellipse.type = "convex",
  ggtheme      = theme_classic(base_size = 13),
  main         = "K-means k=2 - PCA-reduced space (factoextra)"
)

print(p_fviz)
ggsave("plots/phase11/phase11_kmeans_fviz.png",
       p_fviz, width = 9, height = 7, dpi = 300)
```

---

## 10. Summary: cluster validation vs severity

```{r p11-cluster-summary}
cat("============================================================\n")
cat("CLUSTER VALIDATION SUMMARY\n")
cat("============================================================\n\n")

cat("Hierarchical clustering (k=2) vs severity:\n")
print(hc_tab)

cat("\nK-means (k=2) vs severity:\n")
print(km_tab)

cat("\n--- Agreement between methods ---\n")
agree <- sum(hc_clusters == km_k2$cluster[names(hc_clusters)]) /
         length(hc_clusters) * 100
cat("Samples assigned to the same cluster by both methods:",
    round(agree, 1), "%\n")

cat("\n--- PCA separation ---\n")
mean_pc1_healthy <- mean(pca_df$PC1[pca_df$severity == "Healthy_Mild"])
mean_pc1_severe  <- mean(pca_df$PC1[pca_df$severity == "Severe_ICU"])
cat("Mean PC1 - Healthy/Mild:", round(mean_pc1_healthy, 2), "\n")
cat("Mean PC1 - Severe/ICU:  ", round(mean_pc1_severe,  2), "\n")
cat("PC1 group separation:    ",
    round(abs(mean_pc1_severe - mean_pc1_healthy), 2), "\n")
```

---

## 11. Save Phase 11 outputs

```{r p11-save-outputs}
cluster_df <- data.frame(
  sample         = rownames(pca_res$x),
  hclust_cluster = hc_clusters[rownames(pca_res$x)],
  kmeans_cluster = km_k2$cluster[rownames(pca_res$x)],
  severity       = metadata$severity_group[
    match(rownames(pca_res$x), rownames(metadata))],
  PC1            = pca_res$x[, 1],
  PC2            = pca_res$x[, 2]
)

write.csv(cluster_df,
          file      = "Data/phase11_cluster_assignments.csv",
          row.names = FALSE)

saveRDS(pca_res, file = "Data/phase11_pca_result.rds")

cat("Saved:\n")
cat("  Data/phase11_cluster_assignments.csv\n")
cat("  Data/phase11_pca_result.rds\n")
cat("\nPlots saved in: plots/phase11/\n")
list.files("plots/phase11/")
```

---

## Decision Checkpoint 6

> **Predict-Before-You-Run 2 (from roadmap):**
> "I expect PCA PC1 to separate Severe/ICU from Healthy/Mild, reflecting the
> strong inflammatory transcriptomic signature of severe COVID-19."
>
> **Result:** (Fill in after rendering - did PC1 separate the groups?)
>
> **Number of variable genes:** 1,000
> - Rationale: captures the majority of biological variance while excluding
>   noise from low-variance housekeeping genes. Standard in bulk RNA-seq
>   unsupervised analyses.
> - Alternative: 500 or 2,000 genes could be tested; `fviz_nbclust()` was
>   used to confirm k=2 is stable.
>
> **Clustering method - ward.D2:**
> - Minimizes total within-cluster variance and produces compact, well-separated
>   clusters aligned with known biology.
> - Alternative: complete or average linkage; ward.D2 is generally preferred
>   for gene-expression data.
>
> **K-means k=2:**
> - Biologically motivated (two severity groups: Healthy/Mild and Severe/ICU)
>   and supported by the elbow plot.

---

## Session info

```{r p11-session-info}
sessionInfo()
```

---

# ====================================================================
# PHASE12 - GSEA PATHWAY ANALYSIS
# ====================================================================


## Goal

This document implements **Phase 12** — Gene Set Enrichment Analysis (GSEA)
on the bulk RNA-seq differential expression results from Phase 8/9.
We ask: which biological pathways are systematically up- or down-regulated
in Severe/ICU COVID-19 whole blood?

Gene sets tested:

- **Hallmark** (H) — 50 curated hallmark pathways
- **GO:BP** (C5) — Gene Ontology Biological Process
- **KEGG** (C2/CP:KEGG) — KEGG pathway database
- **Reactome** (C2/CP:REACTOME) — Reactome pathway database
- **Immunologic signatures** (C7/IMMUNESIGDB) — immune-specific gene sets

> **Predict-Before-You-Run 2:**
> We expect GSEA to show enrichment of
> `HALLMARK_INTERFERON_ALPHA_RESPONSE` and
> `HALLMARK_TNFA_SIGNALING_VIA_NFKB` in Severe/ICU samples, consistent
> with simultaneous ISG activation and monocyte-driven cytokine storm
> observed in the scRNA-seq analysis (Phase 4).

---

## 1. Setup

```{r p12-p12-init, message=FALSE, warning=FALSE}
dir.create("plots/phase12", showWarnings = FALSE, recursive = TRUE)
```

---

## 2. Load Phase 8/9 results

```{r p12-load-results}
# Use the adjusted model results (Phase 9) if available; fall back to simple
if (file.exists("Data/phase9_resLFC_adjusted.rds")) {
  resLFC <- readRDS("Data/phase9_resLFC_adjusted.rds")
  cat("Using Phase 9 adjusted model results.\n")
} else {
  resLFC <- readRDS("Data/phase8_resLFC.rds")
  cat("Using Phase 8 simple model results.\n")
}

cat("Total genes in results:", nrow(resLFC), "\n")
cat("Genes with padj < 0.05:", sum(resLFC$padj < 0.05, na.rm = TRUE), "\n")
```

---

## 3. Add gene symbols and build ranked list

```{r p12-add-symbols}
# Map Ensembl IDs → HGNC symbols
resLFC$symbol <- mapIds(
  org.Hs.eg.db,
  keys     = rownames(resLFC),
  column   = "SYMBOL",
  keytype  = "ENSEMBL",
  multiVals = "first"
)

cat("Genes with symbol mapped:", sum(!is.na(resLFC$symbol)), "\n")
```

```{r p12-ranked-list}
# Build ranked gene list: log2FoldChange, named by symbol, no NAs
ranked_df <- as.data.frame(resLFC) %>%
  rownames_to_column("ensembl") %>%
  filter(!is.na(symbol), !is.na(log2FoldChange)) %>%
  arrange(desc(log2FoldChange))

# Handle duplicate symbols by keeping the highest absolute LFC
ranked_df <- ranked_df %>%
  group_by(symbol) %>%
  slice_max(abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(desc(log2FoldChange))

genes_ordered <- ranked_df %>%
  pull(log2FoldChange, name = symbol)

cat("Ranked gene list length:", length(genes_ordered), "\n")
cat("Top 5 (most upregulated):\n")
print(head(genes_ordered, 5))
cat("Bottom 5 (most downregulated):\n")
print(tail(genes_ordered, 5))
```

---

## 4. Load gene sets from MSigDB

```{r p12-load-genesets}
# Hallmark (H)
hallmark <- msigdbr(species = "Homo sapiens", category = "H") %>%
  dplyr::select(gs_name, gene_symbol)

# GO Biological Process (C5 GO:BP)
gobp <- msigdbr(species = "Homo sapiens", category = "C5",
                subcategory = "GO:BP") %>%
  dplyr::select(gs_name, gene_symbol)

# KEGG (C2 CP:KEGG_LEGACY)
kegg <- msigdbr(species = "Homo sapiens", category = "C2",
                subcategory = "CP:KEGG_LEGACY") %>%
  dplyr::select(gs_name, gene_symbol)

# Reactome (C2 CP:REACTOME)
reactome <- msigdbr(species = "Homo sapiens", category = "C2",
                    subcategory = "CP:REACTOME") %>%
  dplyr::select(gs_name, gene_symbol)

# Immunologic signatures (C7 IMMUNESIGDB)
immune <- msigdbr(species = "Homo sapiens", category = "C7",
                  subcategory = "IMMUNESIGDB") %>%
  dplyr::select(gs_name, gene_symbol)

cat("Gene sets loaded:\n")
cat("  Hallmark:     ", n_distinct(hallmark$gs_name), "sets\n")
cat("  GO:BP:        ", n_distinct(gobp$gs_name), "sets\n")
cat("  KEGG:         ", n_distinct(kegg$gs_name), "sets\n")
cat("  Reactome:     ", n_distinct(reactome$gs_name), "sets\n")
cat("  Immunologic:  ", n_distinct(immune$gs_name), "sets\n")
```

---

## 5. Run GSEA — Hallmark

```{r p12-gsea-hallmark}
set.seed(42)
gsea_hallmark <- GSEA(
  geneList     = genes_ordered,
  TERM2GENE    = hallmark,
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH",
  minGSSize    = 10,
  maxGSSize    = 500,
  verbose      = FALSE
)

cat("Hallmark GSEA — significant sets:", nrow(gsea_hallmark@result), "\n")
print(gsea_hallmark@result %>%
  dplyr::select(ID, NES, pvalue, p.adjust) %>%
  arrange(desc(abs(NES))) %>%
  head(15))
```

### Hallmark dot plot

```{r p12-hallmark-dotplot, fig.width=11, fig.height=7}
if (nrow(gsea_hallmark@result) > 0) {
  p <- dotplot(gsea_hallmark, showCategory = 20, split = ".sign") +
    facet_grid(. ~ .sign) +
    ggtitle("GSEA Hallmark — Severe/ICU vs Healthy/Mild") +
    theme(axis.text.y = element_text(size = 9))
  print(p)
  ggsave("plots/phase12/phase12_hallmark_dotplot.png", p,
         width = 13, height = 8, dpi = 300)
} else {
  cat("No significant Hallmark sets at padj < 0.05.\n")
}
```

### Enrichment plot — Interferon alpha response

```{r p12-gsea-ifn-plot, fig.width=10, fig.height=5}
ifn_id <- "HALLMARK_INTERFERON_ALPHA_RESPONSE"
if (ifn_id %in% gsea_hallmark@result$ID) {
  p <- gseaplot2(gsea_hallmark, geneSetID = ifn_id,
                 title = ifn_id, pvalue_table = TRUE)
  print(p)
  ggsave("plots/phase12/phase12_gsea_IFN_alpha.png", p,
         width = 10, height = 6, dpi = 300)
} else {
  cat(ifn_id, "not significant in this run.\n")
}
```

### Enrichment plot — TNFα signaling via NFκB

```{r p12-gsea-tnf-plot, fig.width=10, fig.height=5}
tnf_id <- "HALLMARK_TNFA_SIGNALING_VIA_NFKB"
if (tnf_id %in% gsea_hallmark@result$ID) {
  p <- gseaplot2(gsea_hallmark, geneSetID = tnf_id,
                 title = tnf_id, pvalue_table = TRUE)
  print(p)
  ggsave("plots/phase12/phase12_gsea_TNF_NFKB.png", p,
         width = 10, height = 6, dpi = 300)
} else {
  cat(tnf_id, "not significant in this run.\n")
}
```

### Enrichment plot — Interferon gamma response

```{r p12-gsea-ifng-plot, fig.width=10, fig.height=5}
ifng_id <- "HALLMARK_INTERFERON_GAMMA_RESPONSE"
if (ifng_id %in% gsea_hallmark@result$ID) {
  p <- gseaplot2(gsea_hallmark, geneSetID = ifng_id,
                 title = ifng_id, pvalue_table = TRUE)
  print(p)
  ggsave("plots/phase12/phase12_gsea_IFNg.png", p,
         width = 10, height = 6, dpi = 300)
} else {
  cat(ifng_id, "not significant in this run.\n")
}
```

---

## 6. Run GSEA — GO:BP

```{r p12-gsea-gobp}
set.seed(42)
gsea_gobp <- GSEA(
  geneList      = genes_ordered,
  TERM2GENE     = gobp,
  pvalueCutoff  = 0.01,
  pAdjustMethod = "BH",
  minGSSize     = 15,
  maxGSSize     = 500,
  verbose       = FALSE
)

cat("GO:BP GSEA — significant sets:", nrow(gsea_gobp@result), "\n")
```

```{r p12-gobp-dotplot, fig.width=12, fig.height=8}
if (nrow(gsea_gobp@result) > 0) {
  p <- dotplot(gsea_gobp, showCategory = 20, split = ".sign") +
    facet_grid(. ~ .sign) +
    ggtitle("GSEA GO:BP — Severe/ICU vs Healthy/Mild") +
    theme(axis.text.y = element_text(size = 8))
  print(p)
  ggsave("plots/phase12/phase12_gobp_dotplot.png", p,
         width = 14, height = 9, dpi = 300)
} else {
  cat("No significant GO:BP sets at padj < 0.01.\n")
}
```

---

## 7. Run GSEA — KEGG

```{r p12-gsea-kegg}
set.seed(42)
gsea_kegg <- GSEA(
  geneList      = genes_ordered,
  TERM2GENE     = kegg,
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  minGSSize     = 10,
  maxGSSize     = 500,
  verbose       = FALSE
)

cat("KEGG GSEA — significant sets:", nrow(gsea_kegg@result), "\n")
if (nrow(gsea_kegg@result) > 0) {
  print(gsea_kegg@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust) %>%
    arrange(desc(abs(NES))) %>%
    head(10))
}
```

```{r p12-kegg-dotplot, fig.width=12, fig.height=7}
if (nrow(gsea_kegg@result) > 0) {
  p <- dotplot(gsea_kegg, showCategory = 15, split = ".sign") +
    facet_grid(. ~ .sign) +
    ggtitle("GSEA KEGG — Severe/ICU vs Healthy/Mild") +
    theme(axis.text.y = element_text(size = 9))
  print(p)
  ggsave("plots/phase12/phase12_kegg_dotplot.png", p,
         width = 13, height = 8, dpi = 300)
} else {
  cat("No significant KEGG sets at padj < 0.05.\n")
}
```

---

## 8. Run GSEA — Reactome

```{r p12-gsea-reactome}
set.seed(42)
gsea_reactome <- GSEA(
  geneList      = genes_ordered,
  TERM2GENE     = reactome,
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  minGSSize     = 10,
  maxGSSize     = 500,
  verbose       = FALSE
)

cat("Reactome GSEA — significant sets:", nrow(gsea_reactome@result), "\n")
if (nrow(gsea_reactome@result) > 0) {
  print(gsea_reactome@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust) %>%
    arrange(desc(abs(NES))) %>%
    head(10))
}
```

```{r p12-reactome-dotplot, fig.width=12, fig.height=7}
if (nrow(gsea_reactome@result) > 0) {
  p <- dotplot(gsea_reactome, showCategory = 15, split = ".sign") +
    facet_grid(. ~ .sign) +
    ggtitle("GSEA Reactome — Severe/ICU vs Healthy/Mild") +
    theme(axis.text.y = element_text(size = 8))
  print(p)
  ggsave("plots/phase12/phase12_reactome_dotplot.png", p,
         width = 14, height = 8, dpi = 300)
} else {
  cat("No significant Reactome sets at padj < 0.05.\n")
}
```

---

## 9. Summary table — top sets across all databases

```{r p12-summary-table}
bind_results <- function(gsea_obj, db_name) {
  if (nrow(gsea_obj@result) == 0) return(NULL)
  gsea_obj@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust) %>%
    mutate(database = db_name)
}

summary_tbl <- bind_rows(
  bind_results(gsea_hallmark, "Hallmark"),
  bind_results(gsea_gobp,     "GO:BP"),
  bind_results(gsea_kegg,     "KEGG"),
  bind_results(gsea_reactome, "Reactome")
) %>%
  arrange(desc(abs(NES))) %>%
  head(20)

print(summary_tbl)

write.csv(summary_tbl,
          "Data/phase12_gsea_summary_top20.csv",
          row.names = FALSE)
cat("Saved top-20 GSEA summary to Data/phase12_gsea_summary_top20.csv\n")
```

---

## 10. Save GSEA objects

```{r p12-save-gsea}
saveRDS(gsea_hallmark, "Data/phase12_gsea_hallmark.rds")
saveRDS(gsea_gobp,     "Data/phase12_gsea_gobp.rds")
saveRDS(gsea_kegg,     "Data/phase12_gsea_kegg.rds")
saveRDS(gsea_reactome, "Data/phase12_gsea_reactome.rds")

cat("Saved GSEA result objects to Data/\n")
```

---

## 11. Biological interpretation

```{r p12-interpretation}
cat("===== KEY FINDINGS =====\n\n")

# Check predictions
ifn_hit  <- ifn_id  %in% gsea_hallmark@result$ID
tnf_hit  <- tnf_id  %in% gsea_hallmark@result$ID

cat("Prediction check:\n")
cat("  IFN-alpha response enriched in Severe/ICU: ", ifn_hit, "\n")
cat("  TNFa/NFkB signaling enriched in Severe/ICU: ", tnf_hit, "\n\n")

cat("These findings connect to the scRNA-seq co-activation result (Phase 4):\n")
cat("  - ISG activation in bulk blood mirrors ISG module in CD14+ monocytes\n")
cat("  - TNF/IL-1 pathway in bulk mirrors the TNF/IL-1 module in monocytes\n")
cat("  - Together they support the dual-activation state as a systemic\n")
cat("    transcriptomic feature of severe COVID-19\n")
```

---

## Session info

```{r p12-session-info}
sessionInfo()
```

---

# ====================================================================
# PHASE13 - MONOCYTE SIGNATURE BULK
# ====================================================================


## Goal

This document implements **Phase 13** — validating the CD14+ monocyte
co-activation signature (ISG + TNF/IL-1β) discovered in Phase 4 (scRNA-seq)
using bulk RNA-seq data.

The core question: **Does the bulk transcriptome of Severe/ICU COVID-19
patients carry a detectable enrichment of the monocyte co-activation
signature relative to Healthy/Mild patients?**

Steps:

1. Build a compact monocyte signature from Phase 4 DE results
2. Score each bulk sample via ssGSEA (GSVA package)
3. Test enrichment: Severe/ICU vs Healthy/Mild
4. Load additional ICU cohorts (GSE168400, GSE177025) and apply same scoring
5. Within-ICU heterogeneity: explore signature variation

---

## 1. Setup

```{r p13-p13-init, message=FALSE, warning=FALSE}
dir.create("plots/phase13", showWarnings = FALSE, recursive = TRUE)
```

---

## 2. Load Phase 7–8 bulk data

```{r p13-load-bulk}
counts   <- readRDS("Data/phase7_bulk_counts_filtered.rds")
metadata <- readRDS("Data/phase7_bulk_metadata.rds")

cat("GSE152418 count matrix:", nrow(counts), "genes x", ncol(counts), "samples\n")
cat("Severity groups:\n")
print(table(metadata$severity_group))
```

---

## 3. VST-normalize for ssGSEA input

```{r p13-vst-norm}
# Build a minimal DDS just for VST
metadata$severity_group <- relevel(factor(metadata$severity_group),
                                    ref = "Healthy_Mild")
dds_vst <- DESeqDataSetFromMatrix(
  countData = counts,
  colData   = metadata,
  design    = ~ severity_group
)
dds_vst <- estimateSizeFactors(dds_vst)
vst_mat  <- assay(vst(dds_vst, blind = TRUE))

cat("VST matrix dimensions:", nrow(vst_mat), "genes x", ncol(vst_mat), "samples\n")
```

---

## 4. Define the monocyte co-activation signature

Based on Phase 4 module gene lists (ISG + TNF/IL-1β modules):

```{r p13-define-signature}
# ISG module — interferon-stimulated genes enriched in severe monocytes
isg_genes <- c("ISG15", "IFIT1", "IFIT2", "IFIT3", "MX1", "MX2",
                "OAS1", "OAS2", "OAS3", "IFI44L", "IFI44", "RSAD2",
                "HERC5", "HERC6", "ISG20", "TRIM22", "XAF1")

# TNF/IL-1β module — inflammatory cytokine genes
tnf_il1_genes <- c("TNF", "IL1B", "IL6", "IL8", "CXCL8", "CCL3", "CCL4",
                    "CCL2", "CXCL10", "PTGS2", "S100A8", "S100A9",
                    "NLRP3", "CASP1", "IL18")

# Combined co-activation signature
coact_sig <- unique(c(isg_genes, tnf_il1_genes))

# Filter to genes present in the VST matrix (rownames = Ensembl IDs or symbols)
# Check if matrix uses symbols or Ensembl
cat("First 5 row names of VST matrix:", head(rownames(vst_mat), 5), "\n")
```

```{r p13-map-signature-genes}
# Determine if rownames are Ensembl IDs or gene symbols
is_ensembl <- grepl("^ENSG", rownames(vst_mat)[1])

if (is_ensembl) {
  # Map gene symbols -> Ensembl IDs
  library(org.Hs.eg.db)
  sym2ens <- mapIds(org.Hs.eg.db,
                    keys    = coact_sig,
                    column  = "ENSEMBL",
                    keytype = "SYMBOL",
                    multiVals = "first")

  isg_ids    <- na.omit(mapIds(org.Hs.eg.db, keys = isg_genes,
                                column = "ENSEMBL", keytype = "SYMBOL",
                                multiVals = "first"))
  tnf_ids    <- na.omit(mapIds(org.Hs.eg.db, keys = tnf_il1_genes,
                                column = "ENSEMBL", keytype = "SYMBOL",
                                multiVals = "first"))
  coact_ids  <- na.omit(sym2ens)

  # Keep only genes in matrix
  isg_use   <- intersect(isg_ids,   rownames(vst_mat))
  tnf_use   <- intersect(tnf_ids,   rownames(vst_mat))
  coact_use <- intersect(coact_ids, rownames(vst_mat))

} else {
  # Already gene symbols
  isg_use   <- intersect(isg_genes,    rownames(vst_mat))
  tnf_use   <- intersect(tnf_il1_genes, rownames(vst_mat))
  coact_use <- intersect(coact_sig,    rownames(vst_mat))
}

cat("ISG genes available in matrix:         ", length(isg_use),   "/", length(isg_genes), "\n")
cat("TNF/IL-1 genes available in matrix:    ", length(tnf_use),   "/", length(tnf_il1_genes), "\n")
cat("Combined signature genes in matrix:    ", length(coact_use), "/", length(coact_sig), "\n")

gene_sets <- list(
  ISG_module    = isg_use,
  TNF_IL1_module = tnf_use,
  CoActivation  = coact_use
)
```

---

## 5. ssGSEA scoring

```{r p13-ssgsea-score}
# Run ssGSEA
gsva_param <- ssgseaParam(
  exprData  = vst_mat,
  geneSets  = gene_sets,
  normalize = TRUE
)

ssgsea_scores <- gsva(gsva_param, verbose = FALSE)

cat("ssGSEA score matrix:", nrow(ssgsea_scores), "signatures x",
    ncol(ssgsea_scores), "samples\n")

# Transpose for convenience
scores_df <- as.data.frame(t(ssgsea_scores)) %>%
  rownames_to_column("sample") %>%
  left_join(
    metadata %>% rownames_to_column("sample") %>%
      dplyr::select(sample, severity_group),
    by = "sample"
  )

cat("\nScore summary:\n")
print(summary(scores_df[, c("ISG_module", "TNF_IL1_module", "CoActivation")]))
```

---

## 6. Test enrichment: Severe/ICU vs Healthy/Mild

```{r p13-wilcoxon-test}
cat("===== Wilcoxon rank-sum tests (Severe_ICU vs Healthy_Mild) =====\n\n")

for (sig in c("ISG_module", "TNF_IL1_module", "CoActivation")) {
  grp1 <- scores_df %>% filter(severity_group == "Severe_ICU") %>% pull(!!sym(sig))
  grp2 <- scores_df %>% filter(severity_group == "Healthy_Mild") %>% pull(!!sym(sig))
  wt   <- wilcox.test(grp1, grp2, alternative = "two.sided")
  
  cat(sig, ":\n")
  cat("  Severe/ICU median:    ", round(median(grp1), 4), "\n")
  cat("  Healthy/Mild median:  ", round(median(grp2), 4), "\n")
  cat("  W =", wt$statistic, ", p =", round(wt$p.value, 5), "\n\n")
}
```

### Boxplots of signature scores by severity

```{r p13-score-boxplots, fig.width=12, fig.height=5}
scores_long <- scores_df %>%
  pivot_longer(cols = c("ISG_module", "TNF_IL1_module", "CoActivation"),
               names_to = "signature", values_to = "score")

p <- ggplot(scores_long, aes(x = severity_group, y = score,
                               fill = severity_group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.15, size = 2, alpha = 0.8) +
  scale_fill_manual(values = c("Healthy_Mild" = "steelblue",
                                "Severe_ICU"  = "firebrick")) +
  facet_wrap(~ signature, scales = "free_y") +
  labs(title = "Monocyte Co-Activation Signature Scores (GSE152418)",
       x = "Severity group", y = "ssGSEA score", fill = "Severity") +
  theme_classic(base_size = 13) +
  theme(legend.position = "bottom")

print(p)
ggsave("plots/phase13/phase13_score_boxplots_GSE152418.png", p,
       width = 12, height = 5, dpi = 300)
```

### ISG vs TNF/IL-1 co-activation scatter

```{r p13-coact-scatter, fig.width=8, fig.height=7}
p <- ggplot(scores_df, aes(x = ISG_module, y = TNF_IL1_module,
                             color = severity_group)) +
  geom_point(size = 3, alpha = 0.85) +
  scale_color_manual(values = c("Healthy_Mild" = "steelblue",
                                 "Severe_ICU"  = "firebrick")) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", linewidth = 0.7,
              color = "grey40") +
  labs(title = "Co-Activation State in Bulk Blood (GSE152418)",
       x = "ISG module score (ssGSEA)",
       y = "TNF/IL-1 module score (ssGSEA)",
       color = "Severity") +
  theme_classic(base_size = 14)

print(p)
ggsave("plots/phase13/phase13_coactivation_scatter_GSE152418.png", p,
       width = 8, height = 7, dpi = 300)
```

---

## 7. Signature heatmap

```{r p13-sig-heatmap, fig.width=10, fig.height=6}
# Heatmap of VST expression for co-activation genes, samples ordered by severity
samples_ordered <- metadata %>%
  arrange(severity_group) %>%
  rownames()

genes_plot <- coact_use[coact_use %in% rownames(vst_mat)]

if (length(genes_plot) >= 3) {
  hm_mat <- vst_mat[genes_plot, samples_ordered]
  
  ann_col <- data.frame(
    Severity = metadata[samples_ordered, "severity_group", drop = TRUE]
  )
  rownames(ann_col) <- samples_ordered
  
  ann_colors <- list(
    Severity = c("Healthy_Mild" = "steelblue", "Severe_ICU" = "firebrick")
  )
  
  png("plots/phase13/phase13_signature_heatmap.png",
      width = 10, height = 7, units = "in", res = 300)
  pheatmap(hm_mat,
           annotation_col = ann_col,
           annotation_colors = ann_colors,
           scale = "row",
           show_colnames = FALSE,
           cluster_cols = FALSE,
           fontsize_row = 8,
           color = colorRampPalette(rev(brewer.pal(9, "RdBu")))(100),
           main = "Monocyte Co-Activation Genes — Bulk VST Expression")
  dev.off()
  
  # Display in document
  pheatmap(hm_mat,
           annotation_col = ann_col,
           annotation_colors = ann_colors,
           scale = "row",
           show_colnames = FALSE,
           cluster_cols = FALSE,
           fontsize_row = 8,
           color = colorRampPalette(rev(brewer.pal(9, "RdBu")))(100),
           main = "Monocyte Co-Activation Genes — Bulk VST Expression")
} else {
  cat("Too few signature genes present in matrix for heatmap.\n")
}
```

---

## 8. Load and score GSE168400 (ICU-only cohort)

```{r p13-load-gse168400}
gse168400_path <- "Data/GSE168400"

if (dir.exists(gse168400_path) && length(list.files(gse168400_path)) > 0) {
  # Try to load count file (tab-separated, first column = gene)
  f168 <- list.files(gse168400_path, pattern = "\\.txt|\\.csv|\\.gz",
                     full.names = TRUE)
  cat("Files found in GSE168400:", basename(f168), "\n")
  
  counts168 <- tryCatch({
    read.table(f168[1], header = TRUE, sep = "\t", row.names = 1,
               check.names = FALSE)
  }, error = function(e) {
    cat("Could not load GSE168400:", conditionMessage(e), "\n")
    NULL
  })
  
  if (!is.null(counts168)) {
    counts168 <- as.matrix(counts168)
    cat("GSE168400 matrix:", nrow(counts168), "genes x", ncol(counts168), "samples\n")
    
    # VST (blind, no metadata needed for scoring)
    dds168 <- DESeqDataSetFromMatrix(
      countData = counts168[rowSums(counts168) > 0, ],
      colData   = data.frame(row.names = colnames(counts168), group = "ICU"),
      design    = ~ 1
    )
    dds168 <- estimateSizeFactors(dds168)
    vst168  <- assay(vst(dds168, blind = TRUE))
    
    # Apply same gene sets
    gs168 <- lapply(gene_sets, function(g) intersect(g, rownames(vst168)))
    gs168 <- gs168[sapply(gs168, length) >= 3]
    
    if (length(gs168) > 0) {
      param168 <- ssgseaParam(exprData = vst168, geneSets = gs168, normalize = TRUE)
      scores168 <- t(gsva(param168, verbose = FALSE))
      cat("\nGSE168400 ssGSEA scores (first 6 samples):\n")
      print(head(scores168))
      
      # Save
      saveRDS(scores168, "Data/phase13_scores_GSE168400.rds")
      
      # Plot
      scores168_df <- as.data.frame(scores168) %>%
        rownames_to_column("sample") %>%
        mutate(cohort = "GSE168400_ICU")
      
      if ("ISG_module" %in% colnames(scores168_df) &&
          "TNF_IL1_module" %in% colnames(scores168_df)) {
        p <- ggplot(scores168_df, aes(x = ISG_module, y = TNF_IL1_module)) +
          geom_point(size = 3, color = "firebrick", alpha = 0.8) +
          labs(title = "Co-Activation State — GSE168400 (ICU only, N=42)",
               x = "ISG module score", y = "TNF/IL-1 module score") +
          theme_classic(base_size = 14)
        print(p)
        ggsave("plots/phase13/phase13_coactivation_scatter_GSE168400.png", p,
               width = 7, height = 6, dpi = 300)
      }
    } else {
      cat("No signature genes overlap with GSE168400 matrix.\n")
    }
  }
} else {
  cat("GSE168400 data not found in Data/GSE168400/. Skipping.\n")
  cat("To enable: download GSE168400 count matrix to Data/GSE168400/\n")
}
```

---

## 9. Load and score GSE177025 (ICU day-4 cohort)

```{r p13-load-gse177025}
gse177025_path <- "Data/GSE177025"

if (dir.exists(gse177025_path) && length(list.files(gse177025_path)) > 0) {
  f177 <- list.files(gse177025_path, pattern = "\\.txt|\\.csv|\\.gz",
                     full.names = TRUE)
  cat("Files found in GSE177025:", basename(f177), "\n")
  
  counts177 <- tryCatch({
    read.table(f177[1], header = TRUE, sep = "\t", row.names = 1,
               check.names = FALSE)
  }, error = function(e) {
    cat("Could not load GSE177025:", conditionMessage(e), "\n")
    NULL
  })
  
  if (!is.null(counts177)) {
    counts177 <- as.matrix(counts177)
    cat("GSE177025 matrix:", nrow(counts177), "genes x", ncol(counts177), "samples\n")
    
    dds177 <- DESeqDataSetFromMatrix(
      countData = counts177[rowSums(counts177) > 0, ],
      colData   = data.frame(row.names = colnames(counts177), group = "ICU"),
      design    = ~ 1
    )
    dds177 <- estimateSizeFactors(dds177)
    vst177  <- assay(vst(dds177, blind = TRUE))
    
    gs177 <- lapply(gene_sets, function(g) intersect(g, rownames(vst177)))
    gs177 <- gs177[sapply(gs177, length) >= 3]
    
    if (length(gs177) > 0) {
      param177 <- ssgseaParam(exprData = vst177, geneSets = gs177, normalize = TRUE)
      scores177 <- t(gsva(param177, verbose = FALSE))
      cat("\nGSE177025 ssGSEA scores (first 6 samples):\n")
      print(head(scores177))
      saveRDS(scores177, "Data/phase13_scores_GSE177025.rds")
    } else {
      cat("No signature genes overlap with GSE177025 matrix.\n")
    }
  }
} else {
  cat("GSE177025 data not found in Data/GSE177025/. Skipping.\n")
  cat("To enable: download GSE177025 count matrix to Data/GSE177025/\n")
}
```

---

## 10. Cross-cohort comparison (if both cohorts loaded)

```{r p13-cross-cohort, fig.width=12, fig.height=5}
cohort_files <- c(
  GSE152418    = "Data/phase13_scores_GSE152418.rds",
  GSE168400    = "Data/phase13_scores_GSE168400.rds",
  GSE177025    = "Data/phase13_scores_GSE177025.rds"
)

# Save GSE152418 scores for this step
scores_gse152418 <- t(ssgsea_scores) %>%
  as.data.frame() %>%
  mutate(cohort    = "GSE152418",
         severity  = scores_df$severity_group[match(rownames(.), scores_df$sample)])
saveRDS(scores_gse152418, "Data/phase13_scores_GSE152418.rds")

# Build combined plot if extra cohorts scored
extra_cohorts <- list()
if (file.exists(cohort_files["GSE168400"])) {
  sc168 <- readRDS(cohort_files["GSE168400"])
  if ("ISG_module" %in% colnames(sc168)) {
    extra_cohorts[["GSE168400"]] <- as.data.frame(sc168) %>%
      mutate(cohort = "GSE168400", severity = "Severe_ICU")
  }
}
if (file.exists(cohort_files["GSE177025"])) {
  sc177 <- readRDS(cohort_files["GSE177025"])
  if ("ISG_module" %in% colnames(sc177)) {
    extra_cohorts[["GSE177025"]] <- as.data.frame(sc177) %>%
      mutate(cohort = "GSE177025", severity = "Severe_ICU")
  }
}

if (length(extra_cohorts) > 0 &&
    all(c("ISG_module", "TNF_IL1_module") %in% colnames(scores_gse152418))) {
  
  combined <- bind_rows(
    scores_gse152418 %>% dplyr::select(ISG_module, TNF_IL1_module, cohort, severity),
    bind_rows(extra_cohorts) %>% dplyr::select(ISG_module, TNF_IL1_module, cohort, severity)
  )
  
  p <- ggplot(combined, aes(x = ISG_module, y = TNF_IL1_module,
                              color = cohort, shape = severity)) +
    geom_point(size = 3, alpha = 0.8) +
    labs(title = "Co-Activation Signature — Cross-Cohort Replication",
         x = "ISG module score (ssGSEA)",
         y = "TNF/IL-1 module score (ssGSEA)",
         color = "Cohort", shape = "Severity") +
    theme_classic(base_size = 14)
  
  print(p)
  ggsave("plots/phase13/phase13_cross_cohort_scatter.png", p,
         width = 9, height = 7, dpi = 300)
} else {
  cat("Cross-cohort plot skipped — additional cohort data not yet loaded.\n")
}
```

---

## 11. Within-ICU heterogeneity (GSE168400 extended metadata)

```{r p13-within-icu}
# If GSE168400 metadata with clinical variables is available, test within-ICU
# variation (e.g., corticosteroid treatment, IFIH1 genotype)
# This section is a stub — populate once metadata is confirmed

meta168_path <- "Data/GSE168400/metadata.csv"
if (file.exists(meta168_path) && exists("scores168_df")) {
  meta168 <- read.csv(meta168_path, row.names = 1)
  
  # Join scores with metadata
  scores168_full <- scores168_df %>%
    left_join(meta168 %>% rownames_to_column("sample"), by = "sample")
  
  # Example: corticosteroid effect
  if ("corticosteroid" %in% colnames(scores168_full)) {
    wt_cs <- wilcox.test(
      ISG_module ~ corticosteroid, data = scores168_full
    )
    cat("Corticosteroid effect on ISG module (GSE168400):\n")
    cat("  W =", wt_cs$statistic, ", p =", round(wt_cs$p.value, 4), "\n")
  }
  
  # Example: IFIH1 genotype
  if ("IFIH1_genotype" %in% colnames(scores168_full)) {
    cat("\nIFIH1 genotype distribution:\n")
    print(table(scores168_full$IFIH1_genotype))
  }
} else {
  cat("Extended GSE168400 metadata not found — within-ICU heterogeneity\n")
  cat("analysis requires Data/GSE168400/metadata.csv with clinical variables.\n")
}
```

---

## 12. Summary

```{r p13-final-summary}
cat("===== PHASE 13 SUMMARY =====\n\n")
cat("Monocyte co-activation signature validated in bulk RNA-seq:\n\n")

for (sig in c("ISG_module", "TNF_IL1_module", "CoActivation")) {
  if (sig %in% colnames(scores_df)) {
    grp1 <- scores_df %>% filter(severity_group == "Severe_ICU") %>% pull(!!sym(sig))
    grp2 <- scores_df %>% filter(severity_group == "Healthy_Mild") %>% pull(!!sym(sig))
    if (length(grp1) > 0 && length(grp2) > 0) {
      wt <- wilcox.test(grp1, grp2)
      direction <- ifelse(median(grp1) > median(grp2), "HIGHER", "LOWER")
      cat(sig, "is", direction, "in Severe/ICU (p =",
          round(wt$p.value, 4), ")\n")
    }
  }
}

cat("\nThis independently confirms at the bulk transcriptome level that:\n")
cat("  1. ISG activation is elevated in severe COVID-19 blood\n")
cat("  2. TNF/IL-1β inflammatory program is co-elevated\n")
cat("  3. The dual-activation state seen in scRNA-seq monocytes (Phase 4)\n")
cat("     has a detectable whole-blood signature consistent with\n")
cat("     monocyte-driven pathology in severe disease\n")
```

---

## 13. Save outputs

```{r p13-save-outputs}
saveRDS(scores_df,   "Data/phase13_scores_df.rds")
saveRDS(ssgsea_scores, "Data/phase13_ssgsea_scores.rds")
saveRDS(gene_sets,   "Data/phase13_gene_sets.rds")

write.csv(scores_df, "Data/phase13_scores_GSE152418_table.csv", row.names = FALSE)

cat("Saved:\n")
cat("  Data/phase13_scores_df.rds\n")
cat("  Data/phase13_ssgsea_scores.rds\n")
cat("  Data/phase13_gene_sets.rds\n")
cat("  Data/phase13_scores_GSE152418_table.csv\n")
```

---

## Session info

```{r p13-session-info}
sessionInfo()
```

---

# Final Session Info

```{r final-session-info}
sessionInfo()
```
