---
title: "Phase 3 – Clustering & Cell-Type Annotation"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

## Goal

This document implements **Phase 3** of the project workflow:
  
- Graph-based clustering at multiple resolutions
- Manual marker gene inspection
- Automated annotation with SingleR
- Differential marker discovery with `FindAllMarkers()`

---

## 1. Setup

```{r setup, message=FALSE, warning=FALSE}
library(Seurat)
library(tidyverse)
library(patchwork)
library(SingleR)
library(celldex)
library(dittoSeq)
library(pheatmap)

knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6
)
```

---

## 2. Load the processed Seurat object from Phase 2

```{r load-object}
seurat <- readRDS("Data/phase2_seurat_processed.rds")

cat("Loaded Phase 2 Seurat object.\n")
cat("  Cells:", ncol(seurat), "\n")
cat("  Genes:", nrow(seurat), "\n")
cat("  Default assay:", DefaultAssay(seurat), "\n")
```

---

## 3. User-defined parameters

```{r parameters}
# Number of PCs (must match Phase 2)
N_PCS <- 20

# Clustering resolutions to compare
RESOLUTIONS <- c(0.1, 0.3, 0.5, 0.8, 1.0)

# Final resolution to use downstream (adjust after inspection)
FINAL_RES <- 0.5
```

---

## 4. Find Neighbors

Build the shared-nearest-neighbor (SNN) graph on the PCA embedding.

```{r find-neighbors}
seurat <- FindNeighbors(seurat, dims = 1:N_PCS, verbose = FALSE)
```

---

## 5. Find Clusters at multiple resolutions

```{r find-clusters}
for (res in RESOLUTIONS) {
  seurat <- FindClusters(seurat, resolution = res, verbose = FALSE)
}

# Show cluster counts per resolution
res_summary <- data.frame(
  Resolution = RESOLUTIONS,
  Clusters   = sapply(RESOLUTIONS, function(r) {
    length(unique(seurat@meta.data[[paste0("RNA_snn_res.", r)]]))
  })
)
knitr::kable(res_summary, caption = "Number of clusters at each resolution")
```

---

## 6. Compare resolutions on UMAP

```{r compare-resolutions, fig.width=16, fig.height=12}
plots <- lapply(RESOLUTIONS, function(r) {
  col <- paste0("RNA_snn_res.", r)
  DimPlot(seurat, group.by = col, reduction = "umap", label = TRUE, label.size = 3) +
    ggtitle(paste0("res = ", r)) +
    NoLegend()
})
wrap_plots(plots, ncol = 3)
```

---

## 7. Set final clustering resolution

```{r set-final-resolution}
Idents(seurat) <- paste0("RNA_snn_res.", FINAL_RES)
cat("Using resolution:", FINAL_RES, "\n")
cat("Number of clusters:", length(levels(Idents(seurat))), "\n")
```

### 7a. UMAP by cluster (final)

```{r umap-clusters, fig.width=10, fig.height=7}
DimPlot(seurat, reduction = "umap", label = TRUE, pt.size = 0.3) +
  ggtitle(paste0("Clusters (res = ", FINAL_RES, ")"))
```

### 7b. Compare t-SNE vs UMAP

```{r tsne-vs-umap, fig.width=16, fig.height=6}
p1 <- DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle("UMAP") + NoLegend()
p2 <- DimPlot(seurat, reduction = "tsne", label = TRUE) + ggtitle("t-SNE") + NoLegend()
p1 + p2
```

---

## 8. Manual marker gene inspection

### 8a. Monocyte markers

```{r markers-monocytes, fig.width=12, fig.height=4}
FeaturePlot(seurat, features = c("CD14", "LYZ", "FCGR3A"),
            reduction = "umap", ncol = 3)
```

### 8b. T-cell markers

```{r markers-tcells, fig.width=16, fig.height=4}
FeaturePlot(seurat, features = c("CD3D", "CD3E", "CD4", "CD8A"),
            reduction = "umap", ncol = 4)
```

### 8c. B-cell markers

```{r markers-bcells, fig.width=8, fig.height=4}
FeaturePlot(seurat, features = c("MS4A1", "CD79A"),
            reduction = "umap", ncol = 2)
```

### 8d. NK-cell markers

```{r markers-nk, fig.width=8, fig.height=4}
FeaturePlot(seurat, features = c("NKG7", "GNLY"),
            reduction = "umap", ncol = 2)
```

### 8e. Dot plot of canonical markers

```{r dotplot-markers, fig.width=12, fig.height=6}
canonical_markers <- c(
  "CD14", "LYZ", "FCGR3A",       # Monocytes
  "CD3D", "CD3E", "CD4", "CD8A", # T cells
  "MS4A1", "CD79A",               # B cells
  "NKG7", "GNLY",                 # NK cells
  "FCER1A", "CST3",               # Dendritic cells
  "PPBP"                           # Platelets / Megakaryocytes
)
DotPlot(seurat, features = canonical_markers) +
  RotatedAxis() +
  ggtitle("Canonical marker expression by cluster")
```

---

## 9. Automated annotation with SingleR

```{r singler-annotation}
# Use BlueprintEncodeData as reference
singler_ref <- BlueprintEncodeData()

# Get normalized expression matrix
# (SingleR expects log-normalized counts)
expr_matrix <- GetAssayData(seurat, layer = "data")

# Run SingleR (main labels)
singler_results <- SingleR(
  test   = expr_matrix,
  ref    = singler_ref,
  labels = singler_ref$label.main
)

cat("SingleR annotation complete.\n")
cat("Label distribution:\n")
print(sort(table(singler_results$labels), decreasing = TRUE))
```

### 9a. SingleR score heatmap

```{r singler-heatmap, fig.width=12, fig.height=8}
plotScoreHeatmap(singler_results)
```

### 9b. Add SingleR labels to Seurat object

```{r add-singler-labels}
seurat$singler_main <- singler_results$labels

# Cross-tabulate clusters vs SingleR labels
ct <- table(Cluster = Idents(seurat), CellType = seurat$singler_main)
knitr::kable(ct, caption = "Cluster × SingleR cell-type cross-tabulation")
```

### 9c. UMAP coloured by SingleR annotation

```{r umap-singler, fig.width=12, fig.height=7}
DimPlot(seurat, group.by = "singler_main", reduction = "umap",
        label = TRUE, label.size = 3, repel = TRUE) +
  ggtitle("SingleR cell-type annotation")
```

### 9d. dittoSeq UMAP

```{r ditto-umap, fig.width=12, fig.height=7}
dittoDimPlot(seurat, var = "singler_main", reduction.use = "umap")
```

---

## 10. Find all markers (cluster-defining genes)

```{r find-all-markers}
all_markers <- FindAllMarkers(
  seurat,
  only.pos      = TRUE,
  min.pct       = 0.25,
  logfc.threshold = 0.5
)

cat("Total markers found:", nrow(all_markers), "\n")

# Show top 5 per cluster
top5 <- all_markers %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 5)

knitr::kable(top5, caption = "Top 5 markers per cluster (by log2 FC)")
```

### 10a. Marker Heatmap

```{r marker-heatmap, fig.width=14, fig.height=10}
top5_genes <- top5 %>% pull(gene) %>% unique()
DoHeatmap(seurat, features = top5_genes, size = 3) +
  ggtitle("Top 5 markers per cluster")
```

---

## 11. Save the annotated object

```{r save-object}
saveRDS(seurat, file = "Data/phase3_seurat_annotated.rds")

# Also save the markers table for later use
write.csv(all_markers, file = "Data/phase3_all_markers.csv", row.names = FALSE)

cat("Saved annotated Seurat object to: Data/phase3_seurat_annotated.rds\n")
cat("Saved marker table to: Data/phase3_all_markers.csv\n")
```

---

## 12. Decision Checkpoint 3

**Decision:** Clustering resolution.

**What I chose:**

- Resolution: `r FINAL_RES`
- Number of clusters at this resolution: `r length(levels(Idents(seurat)))`

**Why this resolution:**
Compared resolutions `r paste(RESOLUTIONS, collapse = ", ")` (see UMAP grid above).
The chosen resolution balances granularity (enough clusters to separate biologically
distinct populations) against over-splitting (too many clusters with no clear markers).

**What you still need to add:**

- Replace this text with your interpretation of the resolution comparison.
- Note which clusters correspond to which cell types (based on markers + SingleR).
- Comment on any discrepancies between manual markers and SingleR labels.

---

## 13. Next step

Continue with Phase 4 (CD14+ Monocyte Deep-Dive):

- Subset CD14+ monocytes
- ISG and TNF/IL-1 module scoring
- Co-activation analysis by disease severity
- Donor-level statistical tests
