---
title: "Phase 3 – Clustering & Cell-Type Annotation"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

## Goal

This document implements **Phase 3** of the project workflow:
  
- Graph-based clustering at multiple resolutions
- Manual marker gene inspection
- Automated annotation with SingleR
- Differential marker discovery with `FindAllMarkers()`

---

## 1. Setup

```{r setup, message=FALSE, warning=FALSE}
library(Seurat)
library(tidyverse)
library(patchwork)
library(SingleR)
library(celldex)
library(dittoSeq)
library(pheatmap)

knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6
)

# Create plots output folder
PLOTS_DIR <- "plots/phase3"
dir.create(PLOTS_DIR, showWarnings = FALSE, recursive = TRUE)
```

---

## 2. Load the processed Seurat object from Phase 2

```{r load-object}
seurat <- readRDS("Data/phase2_seurat_processed.rds")

cat("Loaded Phase 2 Seurat object.\n")
cat("  Cells:", ncol(seurat), "\n")
cat("  Genes:", nrow(seurat), "\n")
cat("  Default assay:", DefaultAssay(seurat), "\n")
```

---

## 3. User-defined parameters

```{r parameters}
# Number of PCs (must match Phase 2)
N_PCS <- 20

# Clustering resolutions to compare
RESOLUTIONS <- c(0.1, 0.3, 0.5, 0.8, 1.0)

# Final resolution to use downstream (adjust after inspection)
FINAL_RES <- 0.5
```

---

## 4. Find Neighbors

Build the shared-nearest-neighbor (SNN) graph on the PCA embedding.

```{r find-neighbors}
seurat <- FindNeighbors(seurat, dims = 1:N_PCS, verbose = FALSE)
```

---

## 5. Find Clusters at multiple resolutions

```{r find-clusters}
for (res in RESOLUTIONS) {
  seurat <- FindClusters(seurat, resolution = res, verbose = FALSE)
}

# Show cluster counts per resolution
res_summary <- data.frame(
  Resolution = RESOLUTIONS,
  Clusters   = sapply(RESOLUTIONS, function(r) {
    length(unique(seurat@meta.data[[paste0("RNA_snn_res.", r)]]))
  })
)
knitr::kable(res_summary, caption = "Number of clusters at each resolution")
```

---

## 6. Compare resolutions on UMAP

```{r compare-resolutions, fig.width=16, fig.height=12}
plots <- lapply(RESOLUTIONS, function(r) {
  col <- paste0("RNA_snn_res.", r)
  DimPlot(seurat, group.by = col, reduction = "umap", label = TRUE, label.size = 3) +
    ggtitle(paste0("res = ", r)) +
    NoLegend()
})
p_compare_res <- wrap_plots(plots, ncol = 3)
p_compare_res
ggsave(file.path(PLOTS_DIR, "phase3_step06_resolutions_comparison.png"),
       plot = p_compare_res, width = 16, height = 12, dpi = 150)
```

---

## 7. Set final clustering resolution

```{r set-final-resolution}
Idents(seurat) <- paste0("RNA_snn_res.", FINAL_RES)
cat("Using resolution:", FINAL_RES, "\n")
cat("Number of clusters:", length(levels(Idents(seurat))), "\n")
```

### 7a. UMAP by cluster (final)

```{r umap-clusters, fig.width=10, fig.height=7}
p_umap_clusters <- DimPlot(seurat, reduction = "umap", label = TRUE, pt.size = 0.3) +
  ggtitle(paste0("Clusters (res = ", FINAL_RES, ")"))
p_umap_clusters
ggsave(file.path(PLOTS_DIR, "phase3_step07a_umap_clusters_final.png"),
       plot = p_umap_clusters, width = 10, height = 7, dpi = 150)
```

### 7b. Compare t-SNE vs UMAP

```{r tsne-vs-umap, fig.width=16, fig.height=6}
p1 <- DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle("UMAP") + NoLegend()
p2 <- DimPlot(seurat, reduction = "tsne", label = TRUE) + ggtitle("t-SNE") + NoLegend()
p_compare_embed <- p1 + p2
p_compare_embed
ggsave(file.path(PLOTS_DIR, "phase3_step07b_umap_vs_tsne.png"),
       plot = p_compare_embed, width = 16, height = 6, dpi = 150)
```

---

## 8. Manual marker gene inspection

### 8a. Monocyte markers

```{r markers-monocytes, fig.width=12, fig.height=4}
p_markers_mono <- FeaturePlot(seurat, features = c("CD14", "LYZ", "FCGR3A"),
            reduction = "umap", ncol = 3)
p_markers_mono
ggsave(file.path(PLOTS_DIR, "phase3_step08a_markers_monocytes.png"),
       plot = p_markers_mono, width = 12, height = 4, dpi = 150)
```

### 8b. T-cell markers

```{r markers-tcells, fig.width=16, fig.height=4}
p_markers_t <- FeaturePlot(seurat, features = c("CD3D", "CD3E", "CD4", "CD8A"),
            reduction = "umap", ncol = 4)
p_markers_t
ggsave(file.path(PLOTS_DIR, "phase3_step08b_markers_tcells.png"),
       plot = p_markers_t, width = 16, height = 4, dpi = 150)
```
```

### 8c. B-cell markers

```{r markers-bcells, fig.width=8, fig.height=4}
p_markers_b <- FeaturePlot(seurat, features = c("MS4A1", "CD79A"),
            reduction = "umap", ncol = 2)
p_markers_b
ggsave(file.path(PLOTS_DIR, "phase3_step08c_markers_bcells.png"),
       plot = p_markers_b, width = 8, height = 4, dpi = 150)
```

### 8d. NK-cell markers

```{r markers-nk, fig.width=8, fig.height=4}
p_markers_nk <- FeaturePlot(seurat, features = c("NKG7", "GNLY"),
            reduction = "umap", ncol = 2)
p_markers_nk
ggsave(file.path(PLOTS_DIR, "phase3_step08d_markers_nk.png"),
       plot = p_markers_nk, width = 8, height = 4, dpi = 150)
```

### 8e. Dot plot of canonical markers

```{r dotplot-markers, fig.width=12, fig.height=6}
canonical_markers <- c(
  "CD14", "LYZ", "FCGR3A",       # Monocytes
  "CD3D", "CD3E", "CD4", "CD8A", # T cells
  "MS4A1", "CD79A",               # B cells
  "NKG7", "GNLY",                 # NK cells
  "FCER1A", "CST3",               # Dendritic cells
  "PPBP"                           # Platelets / Megakaryocytes
)
p_dotplot <- DotPlot(seurat, features = canonical_markers) +
  RotatedAxis() +
  ggtitle("Canonical marker expression by cluster")
p_dotplot
ggsave(file.path(PLOTS_DIR, "phase3_step08e_dotplot_markers.png"),
       plot = p_dotplot, width = 12, height = 6, dpi = 150)
```

---

## 9. Automated annotation with SingleR

```{r singler-annotation}
# Use BlueprintEncodeData as reference
singler_ref <- BlueprintEncodeData()

# Get normalized expression matrix
# (SingleR expects log-normalized counts)
expr_matrix <- GetAssayData(seurat, layer = "data")

# Run SingleR (main labels)
singler_results <- SingleR(
  test   = expr_matrix,
  ref    = singler_ref,
  labels = singler_ref$label.main
)

cat("SingleR annotation complete.\n")
cat("Label distribution:\n")
print(sort(table(singler_results$labels), decreasing = TRUE))
```

### 9a. SingleR score heatmap

```{r singler-heatmap, fig.width=12, fig.height=8}
png(file.path(PLOTS_DIR, "phase3_step09a_singler_heatmap.png"), width = 1200, height = 800)
plotScoreHeatmap(singler_results)
dev.off()
plotScoreHeatmap(singler_results)
```

### 9b. Add SingleR labels to Seurat object

```{r add-singler-labels}
seurat$singler_main <- singler_results$labels

# Cross-tabulate clusters vs SingleR labels
ct <- table(Cluster = Idents(seurat), CellType = seurat$singler_main)
knitr::kable(ct, caption = "Cluster × SingleR cell-type cross-tabulation")
```

### 9c. UMAP coloured by SingleR annotation

```{r umap-singler, fig.width=12, fig.height=7}
p_umap_singler <- DimPlot(seurat, group.by = "singler_main", reduction = "umap",
        label = TRUE, label.size = 3, repel = TRUE) +
  ggtitle("SingleR cell-type annotation")
p_umap_singler
ggsave(file.path(PLOTS_DIR, "phase3_step09c_umap_singler.png"),
       plot = p_umap_singler, width = 12, height = 7, dpi = 150)
```

### 9d. dittoSeq UMAP

```{r ditto-umap, fig.width=12, fig.height=7}
p_ditto <- dittoDimPlot(seurat, var = "singler_main", reduction.use = "umap")
ggsave(file.path(PLOTS_DIR, "phase3_step09d_ditto_umap.png"),
       plot = p_ditto, width = 12, height = 7, dpi = 150)
p_ditto
```

---

## 10. Find all markers (cluster-defining genes)

```{r find-all-markers}
# FindAllMarkers on 55K cells x 23K genes exceeds the 16 GB vector memory limit
# because Seurat 5 materializes the full matrix for LFC calculation before any
# subsampling. Fix: downsample the object itself to 300 cells per cluster so
# both the test and background sides are small (300 x 23 = ~6900 total cells).
# 300 cells per cluster is sufficient to detect robust cluster-defining markers.
set.seed(42)
seurat_sub <- subset(seurat, cells = WhichCells(seurat, downsample = 300))
cat("Downsampled object:", ncol(seurat_sub), "cells\n")

all_markers <- FindAllMarkers(
  seurat_sub,
  only.pos        = TRUE,
  min.pct         = 0.25,
  logfc.threshold = 0.5,
  features        = VariableFeatures(seurat)  # limit to 3000 variable genes
)

cat("Total markers found:", nrow(all_markers), "\n")

# Show top 5 per cluster
top5 <- all_markers %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 5)

knitr::kable(top5, caption = "Top 5 markers per cluster (by log2 FC)")
```

### 10a. Marker Heatmap

```{r marker-heatmap, fig.width=14, fig.height=10}
top5_genes <- top5 %>% pull(gene) %>% unique()
# Use seurat_sub (already downsampled to ~6500 cells) — DoHeatmap on the full
# 55K-cell object exceeds the 16 GB limit. The downsampled object shows the
# same expression patterns since it was drawn proportionally from all clusters.
p_heatmap <- DoHeatmap(seurat_sub, features = top5_genes, size = 3) +
  ggtitle("Top 5 markers per cluster")
png(file.path(PLOTS_DIR, "phase3_step10a_marker_heatmap.png"), width = 1400, height = 1000)
print(p_heatmap)
dev.off()
p_heatmap
```

---

## 11. Save the annotated object

```{r save-object}
saveRDS(seurat, file = "Data/phase3_seurat_annotated.rds")

# Also save the markers table for later use
write.csv(all_markers, file = "Data/phase3_all_markers.csv", row.names = FALSE)

cat("Saved annotated Seurat object to: Data/phase3_seurat_annotated.rds\n")
cat("Saved marker table to: Data/phase3_all_markers.csv\n")
```

---

## 12. Decision Checkpoint 3

**Decision 1 — Clustering resolution: `r FINAL_RES`**

Five resolutions were tested: 0.1 (12 clusters), 0.3 (19), 0.5 (23), 0.8 (35), 1.0 (34).

Resolution 0.5 was chosen for three reasons:
1. **Stability test:** Cluster count *dropped* from 35 to 34 going from 0.8 to 1.0 — the algorithm is re-merging its own artificial splits. Anything ≥ 0.8 is over-split.
2. **UMAP inspection:** At res = 0.5 every visually distinct region of the UMAP has its own cluster and no continuous regions are fragmented. At res = 0.8 the T/NK right continent breaks into confetti.
3. **Marker validation:** `FindAllMarkers` returned biologically identifiable marker genes for all 23 clusters, confirming each cluster represents a real population.

**Decision 2 — FindAllMarkers on a downsampled object (300 cells/cluster)**

The full 55,548-cell object hits the 16 GB memory limit during DE testing. The object was downsampled to 300 cells per cluster (6,513 total) before running `FindAllMarkers`. This is a valid approach: 300 cells per cluster gives ample statistical power to detect highly expressed, cluster-specific markers, and the same 3,000 variable features from the full object were used.

**Population summary (from manual markers + SingleR cross-tabulation):**

- **Myeloid (right UMAP continent):** Clusters 2, 6, 7, 8, 10, 11, 12, 16, 18, 19 → monocyte sub-states (classical, non-classical, inflammasome-active, ISG-high, chemokine-secreting, proliferating)
- **Lymphoid (left UMAP continent):** Clusters 0, 1, 3, 4, 13, 20, 21 → CD8⁺ T, CD4⁺ T, NK, γδ T, Tregs
- **B cells (bottom island):** Clusters 5, 9 → naive and memory B cells
- **Dendritic cells:** Cluster 17 (FCER1A⁺)
- **Contamination:** Cluster 14 (platelets, PPBP/PTCRA), cluster 22 (erythrocytes, GYPA/GYPB)
- **Proliferating cells:** Cluster 15 (TYMS, CDC45 — S-phase genes)

**Discrepancies between manual markers and SingleR:**

- SingleR labels the entire left continent as NK cells + CD8⁺/CD4⁺ T cells, consistent with manual markers.
- Cluster 14 is labelled "Erythrocytes/HSC/Eosinophils" by SingleR — this matches the mixed/contamination interpretation from the dotplot (PPBP = platelets; GYPA-expressing cells in cluster 22 are the true erythrocytes).
- No major contradictions between the two approaches.

---

## 13. Next step

Continue with Phase 4 (CD14+ Monocyte Deep-Dive):

- Subset CD14+ monocytes
- ISG and TNF/IL-1 module scoring
- Co-activation analysis by disease severity
- Donor-level statistical tests
