---
title: "Phase 5 – Replication on GSE150861 (Independent Validation)"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

## Goal

This document implements **Phase 5** — independent validation on GSE150861
(Guo et al., Nat Commun 2020). This dataset contains PBMCs from **2 severe
COVID-19 patients** treated with **Tocilizumab** (anti-IL-6R), sampled at
multiple time points (day 1, day 5, day 7). We process this dataset
**independently** (no joint integration with GSE149689) and check whether:
 
1. The ISG + TNF/IL-1β co-activated monocyte state from Phase 4 replicates.
2. Co-activation changes over time (longitudinal dynamics).
3. Tocilizumab treatment reduces the TNF/IL-1β component.

---

## 1. Setup

```{r setup, message=FALSE, warning=FALSE}
library(Seurat)
library(tidyverse)
library(patchwork)
library(SingleR)
library(celldex)
library(dittoSeq)

knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6
)

dir.create("plots/phase5", showWarnings = FALSE, recursive = TRUE)
```

---

## 2. Load GSE150861 10X data

```{r load-data}
counts <- Read10X(data.dir = "Data/GSE150861/")

cat("Raw matrix dimensions:\n")
cat("  Genes:", nrow(counts), "\n")
cat("  Barcodes:", ncol(counts), "\n")
```

---

## 3. Load batch metadata and create Seurat object

```{r create-seurat}
# Load batch info (barcode → patient-day-rep mapping)
batch_info <- read.csv(gzfile("Data/GSE150861/cell_batch.csv.gz"),
                       stringsAsFactors = FALSE)
rownames(batch_info) <- batch_info$X

# Only keep barcodes that are in both the matrix and batch file
common_barcodes <- intersect(colnames(counts), batch_info$X)
cat("Barcodes in matrix:", ncol(counts), "\n")
cat("Barcodes in batch file:", nrow(batch_info), "\n")
cat("Common barcodes:", length(common_barcodes), "\n")

counts <- counts[, common_barcodes]

# Create Seurat object
rep_seurat <- CreateSeuratObject(
  counts  = counts,
  project = "GSE150861_COVID_Toci",
  min.cells    = 3,
  min.features = 200
)

# Add batch metadata
rep_seurat$batch <- batch_info[colnames(rep_seurat), "batch"]

# Parse batch into patient, day, replicate
rep_seurat$patient <- sub("-day.*", "", rep_seurat$batch)
rep_seurat$day     <- as.integer(sub(".*day(\\d+).*", "\\1", rep_seurat$batch))
rep_seurat$day_label <- paste0("Day ", rep_seurat$day)

cat("\nSeurat object created.\n")
cat("  Cells:", ncol(rep_seurat), "\n")
cat("  Genes:", nrow(rep_seurat), "\n")
cat("\nBatch distribution:\n")
print(table(rep_seurat$batch))
cat("\nPatient × Day:\n")
print(table(rep_seurat$patient, rep_seurat$day_label))
```

---

## 4. QC — same pipeline as Phase 1

```{r qc-metrics}
rep_seurat[["percent.mt"]] <- PercentageFeatureSet(rep_seurat, pattern = "^MT-")
```

### 4a. QC violin plots

```{r qc-violin, fig.width=12, fig.height=4}
p <- VlnPlot(rep_seurat,
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
        ncol = 3, pt.size = 0)
print(p)
ggsave("plots/phase5/phase5_qc_violin.png", p, width = 12, height = 4, dpi = 300)
```

### 4b. QC scatter plots

```{r qc-scatter, fig.width=12, fig.height=4}
p1 <- FeatureScatter(rep_seurat, feature1 = "nCount_RNA", feature2 = "percent.mt") + NoLegend()
p2 <- FeatureScatter(rep_seurat, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") + NoLegend()
p <- p1 + p2
print(p)
ggsave("plots/phase5/phase5_qc_scatter.png", p, width = 12, height = 4, dpi = 300)
```

### 4c. Filter cells

```{r filter-cells}
cat("Before filtering:", ncol(rep_seurat), "cells\n")

rep_seurat <- subset(rep_seurat,
                     subset = nFeature_RNA > 200 &
                              nFeature_RNA < 5000 &
                              percent.mt < 20)

cat("After filtering:", ncol(rep_seurat), "cells\n")
cat("\nBatch distribution after filtering:\n")
print(table(rep_seurat$batch))
```

---

## 5. Normalization (LogNormalize — same as Phase 2 draft)

#TODO: Switch to SCTransform on SSH server with more RAM.

```{r normalize}
rep_seurat <- NormalizeData(rep_seurat, normalization.method = "LogNormalize",
                            scale.factor = 10000)
rep_seurat <- FindVariableFeatures(rep_seurat, selection.method = "vst",
                                    nfeatures = 3000)
rep_seurat <- ScaleData(rep_seurat, vars.to.regress = "percent.mt")
```

---

## 6. PCA → UMAP → Clustering

```{r pca}
N_PCS <- 20

rep_seurat <- RunPCA(rep_seurat, verbose = FALSE)
```

```{r elbow, fig.width=8, fig.height=4}
p <- ElbowPlot(rep_seurat, ndims = 30) +
  geom_vline(xintercept = N_PCS, linetype = "dashed", color = "red")
print(p)
ggsave("plots/phase5/phase5_elbow_plot.png", p, width = 8, height = 4, dpi = 300)
```

```{r umap-cluster}
rep_seurat <- RunUMAP(rep_seurat, dims = 1:N_PCS, verbose = FALSE)
rep_seurat <- FindNeighbors(rep_seurat, dims = 1:N_PCS, verbose = FALSE)
rep_seurat <- FindClusters(rep_seurat, resolution = 0.5, verbose = FALSE)
```

```{r umap-plot, fig.width=14, fig.height=5}
p1 <- DimPlot(rep_seurat, reduction = "umap", label = TRUE) +
  ggtitle("Clusters") + NoLegend()
p2 <- DimPlot(rep_seurat, reduction = "umap", group.by = "patient") +
  ggtitle("Patient")
p3 <- DimPlot(rep_seurat, reduction = "umap", group.by = "day_label") +
  ggtitle("Time point")
p <- p1 + p2 + p3
print(p)
ggsave("plots/phase5/phase5_umap_overview.png", p, width = 14, height = 5, dpi = 300)
```

---

## 7. Cell-type annotation — manual markers + SingleR

### 7a. Manual marker feature plots

```{r marker-featureplots, fig.width=16, fig.height=12}
markers <- c("CD14", "LYZ", "FCGR3A",          # Monocytes
             "CD3D", "CD3E", "CD4", "CD8A",     # T cells
             "MS4A1", "CD79A",                    # B cells
             "NKG7", "GNLY",                      # NK cells
             "FCER1A", "CST3")                    # DCs
p <- FeaturePlot(rep_seurat, features = markers, reduction = "umap", ncol = 4)
print(p)
ggsave("plots/phase5/phase5_marker_featureplots.png", p, width = 16, height = 12, dpi = 300)
```

### 7b. SingleR annotation

```{r singler}
singler_ref <- BlueprintEncodeData()

expr_matrix <- GetAssayData(rep_seurat, layer = "data")
singler_results <- SingleR(
  test   = expr_matrix,
  ref    = singler_ref,
  labels = singler_ref$label.main
)

rep_seurat$singler_main <- singler_results$labels

cat("SingleR label distribution:\n")
print(sort(table(rep_seurat$singler_main), decreasing = TRUE))
```

### 7c. UMAP by cell type

```{r umap-celltype, fig.width=12, fig.height=7}
p <- DimPlot(rep_seurat, group.by = "singler_main", reduction = "umap",
        label = TRUE, repel = TRUE) +
  ggtitle("GSE150861 \u2014 Cell-type annotation (SingleR)")
print(p)
ggsave("plots/phase5/phase5_umap_celltype.png", p, width = 12, height = 7, dpi = 300)
```

---

## 8. Subset monocytes

```{r subset-mono}
mono_labels <- grep("Mono|Monocyte", unique(rep_seurat$singler_main),
                    value = TRUE, ignore.case = TRUE)
cat("Monocyte labels:", paste(mono_labels, collapse = ", "), "\n")

rep_mono <- subset(rep_seurat, singler_main %in% mono_labels)
cat("Monocyte cells:", ncol(rep_mono), "\n")
cat("\nMonocytes per time point:\n")
print(table(rep_mono$day_label, rep_mono$patient))
```

---

## 9. ISG and TNF/IL-1β module scoring

Using the **exact same gene lists** as Phase 4 (GSE149689).

```{r module-scores}
isg_genes     <- c("ISG15", "IFIT1", "MX1", "OAS1", "IFI44L")
tnf_il1_genes <- c("TNF", "IL1B", "CCL3", "CCL4", "CXCL8")

cat("ISG genes present:",
    paste(isg_genes[isg_genes %in% rownames(rep_mono)], collapse = ", "), "\n")
cat("TNF/IL-1 genes present:",
    paste(tnf_il1_genes[tnf_il1_genes %in% rownames(rep_mono)], collapse = ", "), "\n")

rep_mono <- AddModuleScore(rep_mono, features = list(isg_genes), name = "ISG")
rep_mono$ISG_score <- rep_mono$ISG1

rep_mono <- AddModuleScore(rep_mono, features = list(tnf_il1_genes), name = "TNF_IL1")
rep_mono$TNF_IL1_score <- rep_mono$TNF_IL11
```

---

## 10. Co-activation scatter — does the co-activated state replicate?

```{r replication-scatter, fig.width=12, fig.height=6}
df <- rep_mono@meta.data %>%
  select(ISG_score, TNF_IL1_score, patient, day_label, day)

p <- ggplot(df, aes(x = ISG_score, y = TNF_IL1_score, color = day_label)) +
  geom_point(size = 0.5, alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_color_brewer(palette = "Set1") +
  labs(x = "ISG module score", y = "TNF/IL-1\u03b2 module score",
       title = "Replication: ISG vs TNF/IL-1\u03b2 in GSE150861 monocytes",
       color = "Time point") +
  theme_minimal() +
  facet_wrap(~ patient)
print(p)
ggsave("plots/phase5/phase5_coactivation_scatter.png", p, width = 12, height = 6, dpi = 300)
```

---

## 11. Define co-activated state and quantify over time

```{r coactivation-time}
rep_mono$coactivated <- (rep_mono$ISG_score > 0) & (rep_mono$TNF_IL1_score > 0)

coact_by_time <- rep_mono@meta.data %>%
  group_by(patient, day, day_label) %>%
  summarise(
    n_cells         = n(),
    n_coactivated   = sum(coactivated),
    pct_coactivated = round(100 * n_coactivated / n_cells, 1),
    mean_ISG        = mean(ISG_score),
    mean_TNF_IL1    = mean(TNF_IL1_score),
    .groups = "drop"
  ) %>%
  arrange(patient, day)

knitr::kable(coact_by_time,
             caption = "Co-activation in monocytes over time (per patient-day)")
```

### 11a. Longitudinal dynamics plot

```{r longitudinal-plot, fig.width=12, fig.height=5}
p1 <- ggplot(coact_by_time, aes(x = day, y = pct_coactivated,
                                 color = patient, group = patient)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  labs(x = "Day", y = "% co-activated monocytes",
       title = "Co-activation over time") +
  theme_minimal()

p2 <- ggplot(coact_by_time, aes(x = day, y = mean_ISG,
                                 color = patient, group = patient)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  labs(x = "Day", y = "Mean ISG score",
       title = "ISG score over time") +
  theme_minimal()

p3 <- ggplot(coact_by_time, aes(x = day, y = mean_TNF_IL1,
                                 color = patient, group = patient)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  labs(x = "Day", y = "Mean TNF/IL-1β score",
       title = "TNF/IL-1β score over time") +
  theme_minimal()

p <- p1 + p2 + p3
print(p)
ggsave("plots/phase5/phase5_longitudinal_dynamics.png", p, width = 12, height = 5, dpi = 300)
```

---

## 12. Tocilizumab effect on TNF/IL-1β component

```{r toci-effect, fig.width=12, fig.height=5}
# Violin plot of module scores at Day 1 vs later time points
p1 <- VlnPlot(rep_mono, features = "ISG_score", group.by = "day_label",
               pt.size = 0) +
  ggtitle("ISG score by time point") + NoLegend()
p2 <- VlnPlot(rep_mono, features = "TNF_IL1_score", group.by = "day_label",
               pt.size = 0) +
  ggtitle("TNF/IL-1β score by time point") + NoLegend()
p <- p1 + p2
print(p)
ggsave("plots/phase5/phase5_toci_violin.png", p, width = 12, height = 5, dpi = 300)
```

### 12a. Statistical test: Day 1 vs later

```{r toci-stats}
mono_md <- rep_mono@meta.data

# Classify as early (day 1) vs post-treatment (day 5+)
mono_md$phase <- ifelse(mono_md$day == 1, "Pre-Toci (Day 1)", "Post-Toci (Day 5+)")

cat("=== Wilcoxon: ISG score Pre vs Post Tocilizumab ===\n")
print(wilcox.test(ISG_score ~ phase, data = mono_md))

cat("\n=== Wilcoxon: TNF/IL-1 score Pre vs Post Tocilizumab ===\n")
print(wilcox.test(TNF_IL1_score ~ phase, data = mono_md))

cat("\n=== Proportion co-activated: Pre vs Post ===\n")
prop_table <- mono_md %>%
  group_by(phase) %>%
  summarise(
    n = n(),
    coact = sum(coactivated),
    pct   = round(100 * coact / n, 1),
    .groups = "drop"
  )
print(prop_table)

cat("\nFisher's exact test (co-activation Pre vs Post):\n")
mat <- matrix(c(
  prop_table$coact[1], prop_table$n[1] - prop_table$coact[1],
  prop_table$coact[2], prop_table$n[2] - prop_table$coact[2]
), nrow = 2, byrow = TRUE)
print(fisher.test(mat))
```

---

## 13. Cell-type composition over time

```{r composition-time, fig.width=12, fig.height=6}
comp <- rep_seurat@meta.data %>%
  dplyr::count(day_label, patient, singler_main) %>%
  group_by(day_label, patient) %>%
  mutate(pct = n / sum(n) * 100)

p <- ggplot(comp, aes(x = day_label, y = pct, fill = singler_main)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~ patient) +
  labs(x = "Time point", y = "Proportion", fill = "Cell type",
       title = "Cell-type composition over time (GSE150861)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(p)
ggsave("plots/phase5/phase5_composition_time.png", p, width = 12, height = 6, dpi = 300)
```

---

## 14. Save results

```{r save-results}
saveRDS(rep_seurat, file = "Data/phase5_GSE150861_seurat.rds")
saveRDS(rep_mono,   file = "Data/phase5_GSE150861_monocytes.rds")
write.csv(coact_by_time, file = "Data/phase5_coactivation_longitudinal.csv",
          row.names = FALSE)

cat("Saved:\n")
cat("  Data/phase5_GSE150861_seurat.rds\n")
cat("  Data/phase5_GSE150861_monocytes.rds\n")
cat("  Data/phase5_coactivation_longitudinal.csv\n")
```

---

## 15. Replication summary

**Key questions and expected answers:**

1. **Does the co-activated monocyte state replicate?**
   *(Fill in after running — do the scatter plots show the upper-right quadrant
   population?)*

2. **Longitudinal dynamics:**
   *(Does co-activation % decrease at later time points?)*

3. **Tocilizumab effect:**
   *(Does IL-6 blockade reduce the TNF/IL-1β component specifically, while
   ISG scores may remain unchanged?)*

---

## 16. Next step

Continue with Phase 6 (scRNA-seq Visualization Summary) — collect the key
figures from Phases 3–5 into a comprehensive visual overview.
