---
title: "Phase 10 – Bulk RNA-seq Visualization"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

## Goal

This document implements **Phase 10** — a comprehensive visualization summary
of the bulk RNA-seq differential expression results from Phases 8 and 9.
All plots are saved to `plots/phase10/`. We produce:

1. Volcano plot (Severe/ICU vs Healthy/Mild)
2. Heatmap of top DE genes (VST-normalized)
3. Box / jitter plots for key hypothesis genes (ISG + TNF/IL-1 pathway)
4. Count distribution for selected genes across severity groups
5. Summary table of top DE genes

---

## 1. Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(DESeq2)
library(org.Hs.eg.db)
library(EnhancedVolcano)
library(pheatmap)
library(RColorBrewer)
library(patchwork)

knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6
)

dir.create("plots/phase10", showWarnings = FALSE, recursive = TRUE)
```

---

## 2. Load Phase 8–9 outputs

```{r load-data}
dds        <- readRDS("Data/phase8_dds.rds")
resLFC     <- readRDS("Data/phase8_resLFC.rds")
metadata   <- readRDS("Data/phase7_bulk_metadata.rds")

# Also load adjusted model results for comparison, if available
dds2 <- tryCatch(readRDS("Data/phase9_dds_adjusted.rds"), error = function(e) NULL)
res2LFC <- tryCatch(readRDS("Data/phase9_resLFC_adjusted.rds"), error = function(e) NULL)

cat("Loaded DESeqDataSet:", nrow(dds), "genes x", ncol(dds), "samples\n")
cat("Severity groups:\n")
print(table(metadata$severity_group))
```

---

## 3. Order results by significance

```{r order-results}
resOrdered <- resLFC[order(resLFC$pvalue), ]

cat("Significant genes (padj < 0.05):",
    sum(resOrdered$padj < 0.05, na.rm = TRUE), "\n")
cat("Upregulated   (padj < 0.05, LFC > 1):",
    sum(resOrdered$padj < 0.05 & resOrdered$log2FoldChange >  1, na.rm = TRUE), "\n")
cat("Downregulated (padj < 0.05, LFC < -1):",
    sum(resOrdered$padj < 0.05 & resOrdered$log2FoldChange < -1, na.rm = TRUE), "\n")

# Build labels (symbol if mapped, gene ID otherwise)
labs <- resLFC$symbol
labs[is.na(labs)] <- rownames(resLFC)[is.na(labs)]
```

---

## 4. Volcano plot

```{r volcano, fig.width=12, fig.height=9}
p_volcano <- EnhancedVolcano(
  resLFC,
  lab      = labs,
  x        = "log2FoldChange",
  y        = "padj",
  title    = "Severe/ICU vs Healthy/Mild — Bulk RNA-seq (GSE152418)",
  subtitle = "DESeq2 + apeglm shrinkage  |  red: padj < 0.05 & |LFC| > 1",
  pCutoff  = 0.05,
  FCcutoff = 1.0,
  pointSize         = 2.0,
  labSize           = 3.5,
  drawConnectors    = TRUE,
  widthConnectors   = 0.3,
  maxoverlapsConnectors = 25,
  colAlpha           = 0.6,
  legendPosition     = "right"
)

print(p_volcano)
ggsave("plots/phase10/phase10_volcano.png",
       p_volcano, width = 12, height = 9, dpi = 300)
```

---

## 5. Heatmap of top 50 DE genes

```{r heatmap-top50, fig.width=12, fig.height=14}
# VST transform for visualization
vsd <- vst(dds, blind = FALSE)

# Select top 50 by adjusted p-value (exclude NAs)
top_genes <- rownames(resOrdered[!is.na(resOrdered$padj), ])[1:50]

# Sample annotation
anno_col <- data.frame(
  Severity = metadata$severity_group,
  row.names = colnames(vsd)
)

# Add gender annotation if available
if ("gender" %in% colnames(metadata)) {
  anno_col$Gender <- metadata$gender
}

anno_colors <- list(
  Severity = c("Healthy_Mild" = "steelblue", "Severe_ICU" = "firebrick")
)
if ("Gender" %in% colnames(anno_col)) {
  genders <- unique(na.omit(anno_col$Gender))
  pal <- setNames(RColorBrewer::brewer.pal(max(2, length(genders)), "Set1")[seq_along(genders)],
                  genders)
  anno_colors$Gender <- pal
}

# Use gene symbols as row labels where available
row_labels <- resLFC[top_genes, "symbol"]
row_labels[is.na(row_labels)] <- top_genes[is.na(row_labels)]

mat <- assay(vsd)[top_genes, ]
rownames(mat) <- row_labels

# Order columns by severity group
col_order <- order(anno_col$Severity)

p_heatmap <- pheatmap(
  mat[, col_order],
  scale                   = "row",
  annotation_col          = anno_col[col_order, , drop = FALSE],
  annotation_colors       = anno_colors,
  show_colnames           = FALSE,
  fontsize_row            = 8,
  fontsize                = 10,
  clustering_distance_cols = "euclidean",
  clustering_method       = "ward.D2",
  cluster_cols            = FALSE,
  color                   = colorRampPalette(rev(brewer.pal(11, "RdBu")))(100),
  main = "Top 50 DE genes — Severe/ICU vs Healthy/Mild (VST, row-scaled)"
)

ggsave("plots/phase10/phase10_heatmap_top50.png",
       p_heatmap, width = 12, height = 14, dpi = 300)
```

---

## 6. Box / jitter plots for key hypothesis genes

These are the ISG and TNF/IL-1 module genes central to the project hypothesis.

```{r key-gene-boxplots, fig.width=16, fig.height=10}
# ISG module genes
isg_genes <- c("ISG15", "IFIT1", "MX1", "OAS1", "IFI44L", "IFIT3")
# TNF / IL-1 module genes
tnf_genes <- c("TNF", "IL1B", "CCL3", "CCL4", "CXCL8", "IL6")

all_target_symbols <- c(isg_genes, tnf_genes)

# Resolve gene IDs (Ensembl or symbol)
example_id <- rownames(dds)[1]
use_ensembl <- grepl("^ENSG", example_id)

if (use_ensembl) {
  id_map <- mapIds(
    org.Hs.eg.db,
    keys    = all_target_symbols,
    column  = "ENSEMBL",
    keytype = "SYMBOL",
    multiVals = "first"
  )
} else {
  id_map <- setNames(all_target_symbols, all_target_symbols)
}

# Keep only genes present in the dds
id_map <- id_map[!is.na(id_map) & id_map %in% rownames(dds)]
cat("Key genes found in data:", paste(names(id_map), collapse = ", "), "\n")

make_count_plot <- function(gene_id, gene_name, group_var = "severity_group") {
  d <- plotCounts(dds, gene = gene_id, intgroup = group_var, returnData = TRUE)
  ggplot(d, aes(x = .data[[group_var]], y = count, fill = .data[[group_var]])) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7, width = 0.55) +
    geom_jitter(width = 0.15, size = 2.0, alpha = 0.85) +
    scale_y_log10() +
    scale_fill_manual(values = c("Healthy_Mild" = "steelblue",
                                  "Severe_ICU"   = "firebrick")) +
    labs(title = gene_name, x = "", y = "Normalised count (log10)") +
    theme_classic(base_size = 11) +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 30, hjust = 1))
}

plots_isg <- lapply(
  names(id_map)[names(id_map) %in% isg_genes],
  function(sym) make_count_plot(id_map[[sym]], sym)
)

plots_tnf <- lapply(
  names(id_map)[names(id_map) %in% tnf_genes],
  function(sym) make_count_plot(id_map[[sym]], sym)
)

all_plots <- c(plots_isg, plots_tnf)

if (length(all_plots) > 0) {
  p_key <- patchwork::wrap_plots(all_plots,
                                  ncol = min(6, length(all_plots))) +
    patchwork::plot_annotation(
      title   = "Key ISG and TNF/IL-1 genes — normalised counts by severity",
      subtitle = "Box + jitter; y-axis log10-scaled"
    )
  print(p_key)
  ggsave("plots/phase10/phase10_key_gene_boxplots.png",
         p_key, width = 16, height = 10, dpi = 300)
} else {
  message("No key genes found in the dataset. Skipping plot.")
}
```

---

## 7. ISG panel and TNF/IL-1 panel separately

```{r isg-panel, fig.width=14, fig.height=5}
if (length(plots_isg) > 0) {
  p_isg <- patchwork::wrap_plots(plots_isg, ncol = length(plots_isg)) +
    patchwork::plot_annotation(
      title    = "ISG module genes — Severe/ICU vs Healthy/Mild",
      subtitle = "ISG15, IFIT1, MX1, OAS1, IFI44L, IFIT3"
    )
  print(p_isg)
  ggsave("plots/phase10/phase10_ISG_panel.png",
         p_isg, width = 14, height = 5, dpi = 300)
}
```

```{r tnf-panel, fig.width=14, fig.height=5}
if (length(plots_tnf) > 0) {
  p_tnf <- patchwork::wrap_plots(plots_tnf, ncol = length(plots_tnf)) +
    patchwork::plot_annotation(
      title    = "TNF/IL-1 module genes — Severe/ICU vs Healthy/Mild",
      subtitle = "TNF, IL1B, CCL3, CCL4, CXCL8, IL6"
    )
  print(p_tnf)
  ggsave("plots/phase10/phase10_TNF_IL1_panel.png",
         p_tnf, width = 14, height = 5, dpi = 300)
}
```

---

## 8. Summary table of top 20 DE genes

```{r top-de-table}
top20_df <- as.data.frame(resOrdered[1:20, ])
top20_df$symbol <- resLFC[rownames(top20_df), "symbol"]
top20_df$gene_id <- rownames(top20_df)

top20_clean <- top20_df %>%
  dplyr::select(symbol, gene_id, log2FoldChange, lfcSE, pvalue, padj) %>%
  dplyr::mutate(
    log2FoldChange = round(log2FoldChange, 3),
    lfcSE          = round(lfcSE, 3),
    pvalue         = formatC(pvalue, format = "e", digits = 2),
    padj           = formatC(padj, format = "e", digits = 2)
  )

cat("Top 20 differentially expressed genes (Severe/ICU vs Healthy/Mild):\n")
print(top20_clean, row.names = FALSE)

# Save as CSV
write.csv(top20_clean,
          file = "Data/phase10_top20_DE_genes.csv",
          row.names = FALSE)
cat("\nSaved: Data/phase10_top20_DE_genes.csv\n")
```

---

## 9. Comparison: simple vs gender-adjusted model (if Phase 9 ran)

```{r model-comparison, fig.width=14, fig.height=5}
if (!is.null(res2LFC)) {
  cat("Phase 9 results loaded — producing model comparison plot.\n")

  shared <- intersect(rownames(resLFC), rownames(res2LFC))

  df_cmp <- data.frame(
    simple   = resLFC[shared,   "log2FoldChange"],
    adjusted = res2LFC[shared,  "log2FoldChange"],
    sig_simple   = !is.na(resLFC[shared, "padj"]) & resLFC[shared, "padj"] < 0.05,
    sig_adjusted = !is.na(res2LFC[shared, "padj"]) & res2LFC[shared, "padj"] < 0.05
  )

  r_val <- round(cor(df_cmp$simple, df_cmp$adjusted, use = "complete.obs"), 4)

  p_cmp <- ggplot(df_cmp, aes(x = simple, y = adjusted)) +
    geom_point(size = 0.4, alpha = 0.25, color = "grey40") +
    geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", linewidth = 0.8) +
    labs(
      title    = paste0("Simple vs Gender-Adjusted log2FC (r = ", r_val, ")"),
      subtitle = "Each point = one gene; diagonal = perfect agreement",
      x        = "log2FC — simple model (~ severity_group)",
      y        = "log2FC — adjusted model (~ gender + severity_group)"
    ) +
    theme_classic(base_size = 13) +
    coord_equal()

  print(p_cmp)
  ggsave("plots/phase10/phase10_simple_vs_adjusted_lfc.png",
         p_cmp, width = 8, height = 8, dpi = 300)

  # MA plots side by side
  png("plots/phase10/phase10_MA_comparison.png",
      width = 14, height = 5, units = "in", res = 300)
  par(mfrow = c(1, 2))
  plotMA(resLFC,  main = "Simple model (~ severity_group)",          ylim = c(-5, 5))
  plotMA(res2LFC, main = "Adjusted model (~ gender + severity_group)", ylim = c(-5, 5))
  par(mfrow = c(1, 1))
  dev.off()

  par(mfrow = c(1, 2))
  plotMA(resLFC,  main = "Simple model",   ylim = c(-5, 5))
  plotMA(res2LFC, main = "Adjusted model", ylim = c(-5, 5))
  par(mfrow = c(1, 1))
} else {
  cat("Phase 9 results not available — skipping model comparison.\n")
}
```

---

## 10. Save Phase 10 outputs

```{r save-outputs}
# Save the VST matrix for downstream use in Phase 11
vst_mat <- assay(vsd)
saveRDS(vst_mat, file = "Data/phase10_vst_matrix.rds")

cat("Saved:\n")
cat("  Data/phase10_vst_matrix.rds\n")
cat("  Data/phase10_top20_DE_genes.csv\n")
cat("\nPlots saved in: plots/phase10/\n")
list.files("plots/phase10/")
```

---

## Session info

```{r session-info}
sessionInfo()
```
