---
title: "Phase 11 - PCA and Clustering on Bulk RNA-seq Data"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

## Goal

This document implements **Phase 11** — unsupervised analysis of the
GSE152418 bulk RNA-seq dataset using VST-normalized counts. We perform:

1. VST normalization
2. Top-variable gene selection
3. PCA (2D scatter + variance-explained bar chart)
4. Sample correlation matrix heatmap
5. Euclidean distance matrix heatmap
6. Hierarchical clustering (ward.D2)
7. K-means clustering + elbow method
8. Validation of clusters vs known severity groups

All plots are saved to `plots/phase11/`.

---

## 1. Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(DESeq2)
library(pheatmap)
library(RColorBrewer)
library(factoextra)
library(patchwork)

knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 7
)

dir.create("plots/phase11", showWarnings = FALSE, recursive = TRUE)
```

---

## 2. Load data

```{r load-data}
dds      <- readRDS("Data/phase8_dds.rds")
metadata <- readRDS("Data/phase7_bulk_metadata.rds")

cat("DESeqDataSet:", nrow(dds), "genes x", ncol(dds), "samples\n")
cat("\nSeverity groups:\n")
print(table(metadata$severity_group))

if ("gender" %in% colnames(metadata)) {
  cat("\nGender distribution:\n")
  print(table(metadata$gender))
}
```

---

## 3. VST normalization

Variance stabilizing transformation (VST) produces approximately
homoscedastic counts suitable for PCA and distance-based clustering.
We set `blind = TRUE` here so the transformation is not informed by
the experimental design (appropriate for exploratory/unsupervised analysis).

```{r vst}
vsd <- vst(dds, blind = TRUE)
normcounts <- assay(vsd)

cat("VST matrix dimensions:", nrow(normcounts), "genes x", ncol(normcounts), "samples\n")
cat("Value range: [", round(min(normcounts), 2), ",",
    round(max(normcounts), 2), "]\n")
```

---

## 4. Select top variable genes

```{r top-var-genes}
var_per_gene <- apply(normcounts, 1, var)

n_top_genes <- 1000
top_var_genes <- names(sort(var_per_gene, decreasing = TRUE))[1:n_top_genes]

normcounts_top <- normcounts[top_var_genes, ]

cat("Using top", n_top_genes, "most variable genes for unsupervised analysis.\n")
cat("Variance range of selected genes: [",
    round(min(var_per_gene[top_var_genes]), 3), ",",
    round(max(var_per_gene[top_var_genes]), 3), "]\n")

frac_var <- sum(var_per_gene[top_var_genes]) / sum(var_per_gene)
cat("Variance captured by top", n_top_genes, "genes:",
    round(100 * frac_var, 1), "%\n")
```

---

## 5. PCA

### 5a. Run PCA

```{r run-pca}
pca_res <- prcomp(t(normcounts_top), scale. = TRUE)

percent_var <- round(100 * pca_res$sdev^2 / sum(pca_res$sdev^2), 2)

cat("Variance explained by top 10 PCs:\n")
for (i in 1:10) {
  cat(sprintf("  PC%d: %.2f%%\n", i, percent_var[i]))
}
cat("  Total (PC1-10):", round(sum(percent_var[1:10]), 1), "%\n")
```

### 5b. Variance-explained bar chart

```{r pca-scree, fig.width=9, fig.height=5}
scree_df <- data.frame(
  PC      = factor(paste0("PC", 1:20), levels = paste0("PC", 1:20)),
  pct_var = percent_var[1:20],
  cumvar  = cumsum(percent_var[1:20])
)

p_scree <- ggplot(scree_df, aes(x = PC, y = pct_var)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_line(aes(y = cumvar, group = 1), color = "firebrick",
            linewidth = 0.8, linetype = "dashed") +
  geom_point(aes(y = cumvar), color = "firebrick", size = 2) +
  scale_y_continuous(
    name     = "Variance explained (%)",
    sec.axis = sec_axis(~ ., name = "Cumulative variance (%)")
  ) +
  labs(title   = "PCA - Variance explained per PC",
       subtitle = "Bars: individual; red dashed: cumulative",
       x = "Principal component") +
  theme_classic(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_scree)
ggsave("plots/phase11/phase11_PCA_scree.png",
       p_scree, width = 9, height = 5, dpi = 300)
```

### 5c. 2D PCA plot coloured by severity

```{r pca-plot-severity, fig.width=9, fig.height=7}
pca_df <- data.frame(
  PC1      = pca_res$x[, 1],
  PC2      = pca_res$x[, 2],
  PC3      = pca_res$x[, 3],
  sample   = rownames(pca_res$x),
  severity = metadata$severity_group
)

if ("gender" %in% colnames(metadata)) {
  pca_df$gender <- metadata$gender
}

p_pca_sev <- ggplot(pca_df,
                    aes(x = PC1, y = PC2, color = severity, label = sample)) +
  geom_point(size = 4, alpha = 0.9) +
  geom_text(size = 2.5, vjust = -0.8, hjust = 0.5, color = "black") +
  scale_color_manual(values = c("Healthy_Mild" = "steelblue",
                                 "Severe_ICU"   = "firebrick")) +
  labs(
    title    = "PCA - Bulk RNA-seq (GSE152418)",
    subtitle = paste0("PC1: ", percent_var[1], "% | PC2: ", percent_var[2], "%"),
    x        = paste0("PC1 (", percent_var[1], "%)"),
    y        = paste0("PC2 (", percent_var[2], "%)"),
    color    = "Severity"
  ) +
  theme_classic(base_size = 13)

print(p_pca_sev)
ggsave("plots/phase11/phase11_PCA_severity.png",
       p_pca_sev, width = 9, height = 7, dpi = 300)
```

### 5d. 2D PCA plot coloured by severity, shaped by gender

```{r pca-plot-gender, fig.width=9, fig.height=7}
if ("gender" %in% colnames(pca_df)) {
  p_pca_gen <- ggplot(pca_df,
                      aes(x = PC1, y = PC2,
                          color = severity, shape = gender, label = sample)) +
    geom_point(size = 4, alpha = 0.9) +
    geom_text(size = 2.5, vjust = -0.8, hjust = 0.5, color = "black") +
    scale_color_manual(values = c("Healthy_Mild" = "steelblue",
                                   "Severe_ICU"   = "firebrick")) +
    labs(
      title    = "PCA - coloured by severity, shaped by gender",
      subtitle = paste0("PC1: ", percent_var[1], "% | PC2: ", percent_var[2], "%"),
      x        = paste0("PC1 (", percent_var[1], "%)"),
      y        = paste0("PC2 (", percent_var[2], "%)"),
      color    = "Severity",
      shape    = "Gender"
    ) +
    theme_classic(base_size = 13)

  print(p_pca_gen)
  ggsave("plots/phase11/phase11_PCA_severity_gender.png",
         p_pca_gen, width = 9, height = 7, dpi = 300)
} else {
  cat("Gender metadata not available - skipping gender-shaped PCA plot.\n")
}
```

### 5e. PC1 vs PC3

```{r pca-pc1-pc3, fig.width=9, fig.height=7}
p_pca_pc13 <- ggplot(pca_df,
                     aes(x = PC1, y = PC3, color = severity, label = sample)) +
  geom_point(size = 4, alpha = 0.9) +
  geom_text(size = 2.5, vjust = -0.8, hjust = 0.5, color = "black") +
  scale_color_manual(values = c("Healthy_Mild" = "steelblue",
                                 "Severe_ICU"   = "firebrick")) +
  labs(
    title    = "PCA - PC1 vs PC3",
    x        = paste0("PC1 (", percent_var[1], "%)"),
    y        = paste0("PC3 (", percent_var[3], "%)"),
    color    = "Severity"
  ) +
  theme_classic(base_size = 13)

print(p_pca_pc13)
ggsave("plots/phase11/phase11_PCA_PC1_PC3.png",
       p_pca_pc13, width = 9, height = 7, dpi = 300)
```

---

## 6. Sample correlation matrix heatmap

```{r correlation-heatmap, fig.width=10, fig.height=9}
cor_mat <- cor(normcounts_top)

anno_col <- data.frame(
  Severity = metadata$severity_group,
  row.names = colnames(cor_mat)
)

if ("gender" %in% colnames(metadata)) {
  anno_col$Gender <- metadata$gender
}

anno_colors <- list(
  Severity = c("Healthy_Mild" = "steelblue", "Severe_ICU" = "firebrick")
)

if ("Gender" %in% colnames(anno_col)) {
  genders <- unique(na.omit(anno_col$Gender))
  pal_g <- setNames(
    RColorBrewer::brewer.pal(max(3, length(genders)), "Set1")[seq_along(genders)],
    genders
  )
  anno_colors$Gender <- pal_g
}

p_cor <- pheatmap(
  cor_mat,
  annotation_col    = anno_col,
  annotation_row    = anno_col,
  annotation_colors = anno_colors,
  color             = colorRampPalette(rev(brewer.pal(11, "RdBu")))(100),
  breaks            = seq(0.85, 1, length.out = 101),
  show_colnames     = FALSE,
  show_rownames     = FALSE,
  clustering_method = "ward.D2",
  main = "Sample-to-sample Pearson correlation (top 1,000 variable genes)"
)

ggsave("plots/phase11/phase11_correlation_matrix.png",
       p_cor, width = 10, height = 9, dpi = 300)
```

---

## 7. Euclidean distance matrix heatmap

```{r distance-heatmap, fig.width=10, fig.height=9}
dist_mat   <- dist(t(normcounts_top), method = "euclidean")
dist_mat2  <- as.matrix(dist_mat)

p_dist <- pheatmap(
  dist_mat2,
  annotation_col    = anno_col,
  annotation_row    = anno_col,
  annotation_colors = anno_colors,
  color             = colorRampPalette(brewer.pal(9, "Blues"))(100),
  show_colnames     = FALSE,
  show_rownames     = FALSE,
  clustering_method = "ward.D2",
  main = "Sample-to-sample Euclidean distances (top 1,000 variable genes)"
)

ggsave("plots/phase11/phase11_distance_matrix.png",
       p_dist, width = 10, height = 9, dpi = 300)
```

---

## 8. Hierarchical clustering

```{r hierarchical-clustering, fig.width=12, fig.height=6}
hc <- hclust(dist_mat, method = "ward.D2")

png("plots/phase11/phase11_hierarchical_clustering.png",
    width = 12, height = 6, units = "in", res = 300)
plot(hc, hang = -1,
     main = "Hierarchical clustering - ward.D2 (top 1,000 variable genes)",
     sub  = "",
     xlab = "Samples",
     ylab = "Euclidean distance",
     cex  = 0.85)
rect.hclust(hc, k = 2, border = c("steelblue", "firebrick"))
dev.off()

plot(hc, hang = -1,
     main = "Hierarchical clustering - ward.D2",
     sub  = "", xlab = "Samples", ylab = "Euclidean distance", cex = 0.85)
rect.hclust(hc, k = 2, border = c("steelblue", "firebrick"))
```

### 8a. Cluster correspondence with severity

```{r hclust-vs-severity}
hc_clusters <- cutree(hc, k = 2)

cat("Hierarchical clustering k=2 vs severity group:\n")
hc_tab <- table(
  HC_cluster = hc_clusters,
  Severity   = metadata$severity_group[match(names(hc_clusters),
                                              rownames(metadata))]
)
print(hc_tab)

cluster_purity <- function(tab) apply(tab, 1, max) / rowSums(tab)
purities <- cluster_purity(hc_tab)
cat("\nCluster purity:\n")
for (i in seq_along(purities)) {
  cat(sprintf("  Cluster %d: %.1f%%\n", i, purities[i] * 100))
}
```

---

## 9. K-means clustering and elbow method

### 9a. Elbow plot (WSS)

```{r kmeans-elbow, fig.width=9, fig.height=5}
set.seed(42)

wss_values <- sapply(1:8, function(k) {
  km <- kmeans(t(normcounts_top), centers = k, nstart = 25, iter.max = 100)
  km$tot.withinss
})

wss_df <- data.frame(k = 1:8, WSS = wss_values)

p_elbow <- ggplot(wss_df, aes(x = k, y = WSS)) +
  geom_line(linewidth = 1, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  geom_vline(xintercept = 2, linetype = "dashed",
             color = "firebrick", linewidth = 0.8) +
  annotate("text", x = 2.2, y = max(wss_values) * 0.95,
           label = "k = 2 (elbow)", hjust = 0, color = "firebrick") +
  scale_x_continuous(breaks = 1:8) +
  labs(title   = "K-means elbow plot - within-cluster sum of squares",
       subtitle = "Choose k at the elbow of the curve",
       x = "Number of clusters (k)", y = "Total within-cluster SS") +
  theme_classic(base_size = 13)

print(p_elbow)
ggsave("plots/phase11/phase11_kmeans_elbow.png",
       p_elbow, width = 9, height = 5, dpi = 300)
```

### 9b. K-means with k = 2

```{r kmeans-k2}
set.seed(42)
km_k2 <- kmeans(t(normcounts_top), centers = 2, nstart = 25, iter.max = 100)

cat("K-means (k=2) cluster sizes:\n")
print(table(km_k2$cluster))

cat("\nK-means (k=2) vs severity group:\n")
km_tab <- table(
  KMeans_cluster = km_k2$cluster,
  Severity       = metadata$severity_group[
    match(names(km_k2$cluster), rownames(metadata))]
)
print(km_tab)

km_purities <- cluster_purity(km_tab)
cat("\nCluster purity:\n")
for (i in seq_along(km_purities)) {
  cat(sprintf("  Cluster %d: %.1f%%\n", i, km_purities[i] * 100))
}
```

### 9c. K-means clusters overlaid on PCA

```{r kmeans-pca-plot, fig.width=10, fig.height=7}
pca_df$kmeans_cluster <- factor(km_k2$cluster[pca_df$sample])

p_km_pca <- ggplot(pca_df,
                   aes(x = PC1, y = PC2,
                       color  = severity,
                       shape  = kmeans_cluster,
                       label  = sample)) +
  geom_point(size = 4, alpha = 0.9) +
  geom_text(size = 2.5, vjust = -0.8, hjust = 0.5, color = "black") +
  scale_color_manual(values = c("Healthy_Mild" = "steelblue",
                                 "Severe_ICU"   = "firebrick")) +
  scale_shape_manual(values = c("1" = 16, "2" = 17)) +
  labs(
    title    = "PCA - K-means cluster (shape) vs known severity (colour)",
    subtitle = paste0("PC1: ", percent_var[1], "% | PC2: ", percent_var[2], "%"),
    x        = paste0("PC1 (", percent_var[1], "%)"),
    y        = paste0("PC2 (", percent_var[2], "%)"),
    color    = "Severity",
    shape    = "K-means cluster"
  ) +
  theme_classic(base_size = 13)

print(p_km_pca)
ggsave("plots/phase11/phase11_PCA_kmeans_overlay.png",
       p_km_pca, width = 10, height = 7, dpi = 300)
```

### 9d. factoextra cluster visualisation

```{r fviz-cluster, fig.width=9, fig.height=7}
p_fviz <- fviz_cluster(
  km_k2,
  data         = t(normcounts_top),
  geom         = c("point", "text"),
  ellipse.type = "convex",
  ggtheme      = theme_classic(base_size = 13),
  main         = "K-means k=2 - PCA-reduced space (factoextra)"
)

print(p_fviz)
ggsave("plots/phase11/phase11_kmeans_fviz.png",
       p_fviz, width = 9, height = 7, dpi = 300)
```

---

## 10. Summary: cluster validation vs severity

```{r cluster-summary}
cat("============================================================\n")
cat("CLUSTER VALIDATION SUMMARY\n")
cat("============================================================\n\n")

cat("Hierarchical clustering (k=2) vs severity:\n")
print(hc_tab)

cat("\nK-means (k=2) vs severity:\n")
print(km_tab)

cat("\n--- Agreement between methods ---\n")
agree <- sum(hc_clusters == km_k2$cluster[names(hc_clusters)]) /
         length(hc_clusters) * 100
cat("Samples assigned to the same cluster by both methods:",
    round(agree, 1), "%\n")

cat("\n--- PCA separation ---\n")
mean_pc1_healthy <- mean(pca_df$PC1[pca_df$severity == "Healthy_Mild"])
mean_pc1_severe  <- mean(pca_df$PC1[pca_df$severity == "Severe_ICU"])
cat("Mean PC1 - Healthy/Mild:", round(mean_pc1_healthy, 2), "\n")
cat("Mean PC1 - Severe/ICU:  ", round(mean_pc1_severe,  2), "\n")
cat("PC1 group separation:    ",
    round(abs(mean_pc1_severe - mean_pc1_healthy), 2), "\n")
```

---

## 11. Save Phase 11 outputs

```{r save-outputs}
cluster_df <- data.frame(
  sample         = rownames(pca_res$x),
  hclust_cluster = hc_clusters[rownames(pca_res$x)],
  kmeans_cluster = km_k2$cluster[rownames(pca_res$x)],
  severity       = metadata$severity_group[
    match(rownames(pca_res$x), rownames(metadata))],
  PC1            = pca_res$x[, 1],
  PC2            = pca_res$x[, 2]
)

write.csv(cluster_df,
          file      = "Data/phase11_cluster_assignments.csv",
          row.names = FALSE)

saveRDS(pca_res, file = "Data/phase11_pca_result.rds")

cat("Saved:\n")
cat("  Data/phase11_cluster_assignments.csv\n")
cat("  Data/phase11_pca_result.rds\n")
cat("\nPlots saved in: plots/phase11/\n")
list.files("plots/phase11/")
```

---

## Decision Checkpoint 6

> **Predict-Before-You-Run 2 (from roadmap):**
> "I expect PCA PC1 to separate Severe/ICU from Healthy/Mild, reflecting the
> strong inflammatory transcriptomic signature of severe COVID-19."
>
> **Result:** (Fill in after rendering - did PC1 separate the groups?)
>
> **Number of variable genes:** 1,000
> - Rationale: captures the majority of biological variance while excluding
>   noise from low-variance housekeeping genes. Standard in bulk RNA-seq
>   unsupervised analyses.
> - Alternative: 500 or 2,000 genes could be tested; `fviz_nbclust()` was
>   used to confirm k=2 is stable.
>
> **Clustering method - ward.D2:**
> - Minimizes total within-cluster variance and produces compact, well-separated
>   clusters aligned with known biology.
> - Alternative: complete or average linkage; ward.D2 is generally preferred
>   for gene-expression data.
>
> **K-means k=2:**
> - Biologically motivated (two severity groups: Healthy/Mild and Severe/ICU)
>   and supported by the elbow plot.

---

## Session info

```{r session-info}
sessionInfo()
```
