---
title: "Phase 8 – DESeq2 Differential Expression (Severity-Specific)"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

## Goal

This document implements **Phase 8** — differential expression analysis on
the GSE152418 bulk RNA-seq dataset using DESeq2. We compare **Severe/ICU** vs
**Healthy/Mild** to identify genes differentially expressed in severe COVID-19
at the whole-blood level.

---

## 1. Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(DESeq2)
library(apeglm)
library(org.Hs.eg.db)
library(EnhancedVolcano)
library(pheatmap)
library(RColorBrewer)

knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6
)

dir.create("plots/phase8", showWarnings = FALSE, recursive = TRUE)
```

---

## 2. Load Phase 7 outputs

```{r load-data}
counts   <- readRDS("Data/phase7_bulk_counts_filtered.rds")
metadata <- readRDS("Data/phase7_bulk_metadata.rds")

cat("Count matrix:", nrow(counts), "genes x", ncol(counts), "samples\n")
cat("Metadata:", nrow(metadata), "samples\n")
cat("\nSeverity group distribution:\n")
print(table(metadata$severity_group))

# Ensure reference level
metadata$severity_group <- relevel(factor(metadata$severity_group),
                                    ref = "Healthy_Mild")
```

---

## 3. Create DESeqDataSet

```{r create-dds}
dds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData   = metadata,
  design    = ~ severity_group
)

cat("DESeqDataSet created:\n")
cat("  Genes:", nrow(dds), "\n")
cat("  Samples:", ncol(dds), "\n")
```

---

## 4. Library size barplot

```{r library-sizes, fig.width=10, fig.height=5}
lib_df <- data.frame(
  sample   = colnames(dds),
  lib_size = colSums(counts(dds)),
  severity = dds$severity_group
)

p <- ggplot(lib_df, aes(x = reorder(sample, lib_size), y = lib_size / 1e6,
                        fill = severity)) +
  geom_col() +
  scale_fill_manual(values = c("Healthy_Mild" = "steelblue",
                                "Severe_ICU" = "firebrick")) +
  labs(title = "Library sizes by severity group",
       x = "Sample", y = "Library size (millions)", fill = "Severity") +
  theme_classic(base_size = 12) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 7))

print(p)
ggsave("plots/phase8/phase8_library_sizes.png", p, width = 10, height = 5, dpi = 300)
```

---

## 5. Estimate size factors

```{r size-factors}
dds <- estimateSizeFactors(dds)

cat("Size factors:\n")
sf_df <- data.frame(
  sample      = colnames(dds),
  size_factor = sizeFactors(dds),
  severity    = dds$severity_group
)
print(sf_df)

cat("\nSize factor range:", round(range(sizeFactors(dds)), 3), "\n")
```

---

## 6. Mean-variance relationship

```{r mean-var, fig.width=8, fig.height=6}
# Compute mean and variance per gene
gene_means <- rowMeans(counts(dds))
gene_vars  <- apply(counts(dds), 1, var)

mv_df <- data.frame(mean = gene_means, variance = gene_vars)

p <- ggplot(mv_df, aes(x = log10(mean + 1), y = log10(variance + 1))) +
  geom_point(size = 0.3, alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Mean-variance relationship (raw counts)",
       subtitle = "Points above red line = overdispersed (variance > mean)",
       x = "log10(mean + 1)", y = "log10(variance + 1)") +
  theme_classic(base_size = 13)

print(p)
ggsave("plots/phase8/phase8_mean_variance.png", p, width = 8, height = 6, dpi = 300)
```

---

## 7. Run DESeq2

```{r run-deseq}
dds <- DESeq(dds)

cat("DESeq2 complete.\n")
cat("Result names:", resultsNames(dds), "\n")
```

---

## 8. Extract results

```{r extract-results}
res <- results(dds)

cat("Results summary (unshrunken):\n")
summary(res)
```

---

## 9. Shrink fold changes (apeglm)

```{r shrink-lfc}
# Identify the correct coefficient name
coef_name <- grep("Severe", resultsNames(dds), value = TRUE)
cat("Shrinking coefficient:", coef_name, "\n")

resLFC <- lfcShrink(dds, coef = coef_name, type = "apeglm")

cat("\nResults summary (shrunken):\n")
summary(resLFC)
```

---

## 10. MA plots: before and after shrinkage

```{r ma-plots, fig.width=12, fig.height=5}
par(mfrow = c(1, 2))

plotMA(res, main = "MA plot — unshrunken", ylim = c(-5, 5))
plotMA(resLFC, main = "MA plot — apeglm shrunken", ylim = c(-5, 5))

par(mfrow = c(1, 1))

# Save as ggplot-compatible version
png("plots/phase8/phase8_MA_plots.png", width = 12, height = 5,
    units = "in", res = 300)
par(mfrow = c(1, 2))
plotMA(res, main = "MA plot — unshrunken", ylim = c(-5, 5))
plotMA(resLFC, main = "MA plot — apeglm shrunken", ylim = c(-5, 5))
par(mfrow = c(1, 1))
dev.off()
```

---

## 11. Map gene symbols

```{r map-symbols}
# Determine key type from row names
example_genes <- head(rownames(resLFC))
cat("Example gene IDs:", paste(example_genes, collapse = ", "), "\n")

# Check if ENSEMBL or SYMBOL
if (any(grepl("^ENSG", example_genes))) {
  cat("Gene IDs are Ensembl → mapping to symbols\n")
  resLFC$symbol <- mapIds(org.Hs.eg.db,
                          keys = rownames(resLFC),
                          column = "SYMBOL",
                          keytype = "ENSEMBL",
                          multiVals = "first")
  resLFC$ensembl <- rownames(resLFC)
} else {
  cat("Gene IDs appear to be symbols already\n")
  resLFC$symbol <- rownames(resLFC)
}

# How many mapped?
cat("Genes with symbols:", sum(!is.na(resLFC$symbol)), "/", nrow(resLFC), "\n")
```

---

## 12. Order by significance and count DE genes

```{r order-results}
resOrdered <- resLFC[order(resLFC$pvalue), ]

cat("Significant genes (padj < 0.05):",
    sum(resOrdered$padj < 0.05, na.rm = TRUE), "\n")
cat("Upregulated (padj < 0.05, LFC > 0):",
    sum(resOrdered$padj < 0.05 & resOrdered$log2FoldChange > 0, na.rm = TRUE), "\n")
cat("Downregulated (padj < 0.05, LFC < 0):",
    sum(resOrdered$padj < 0.05 & resOrdered$log2FoldChange < 0, na.rm = TRUE), "\n")

cat("\nTop 20 DE genes:\n")
top20 <- head(as.data.frame(resOrdered), 20)
top20$symbol <- resLFC[rownames(top20), "symbol"]
print(top20[, c("symbol", "log2FoldChange", "padj")])
```

---

## 13. Volcano plot

```{r volcano, fig.width=10, fig.height=8}
# Use symbols for labels if available
labs <- if (!is.null(resLFC$symbol)) resLFC$symbol else rownames(resLFC)
labs[is.na(labs)] <- ""

p <- EnhancedVolcano(resLFC,
                     lab = labs,
                     x   = "log2FoldChange",
                     y   = "padj",
                     title    = "Severe/ICU vs Healthy/Mild — Bulk RNA-seq (GSE152418)",
                     subtitle = "DESeq2 with apeglm shrinkage",
                     pCutoff  = 0.05,
                     FCcutoff = 1.0,
                     pointSize = 1.5,
                     labSize   = 3.5,
                     drawConnectors = TRUE,
                     widthConnectors = 0.3,
                     maxoverlapsConnectors = 20)

print(p)
ggsave("plots/phase8/phase8_volcano.png", p, width = 10, height = 8, dpi = 300)
```

---

## 14. Heatmap of top DE genes

```{r heatmap, fig.width=10, fig.height=8}
# VST for visualization
vsd <- vst(dds, blind = FALSE)

# Select top 30 DE genes by adjusted p-value
top_genes <- head(rownames(resOrdered[!is.na(resOrdered$padj), ]), 30)

# Annotation column
anno_col <- data.frame(
  Severity = metadata$severity_group,
  row.names = colnames(vsd)
)

anno_colors <- list(
  Severity = c("Healthy_Mild" = "steelblue", "Severe_ICU" = "firebrick")
)

# Use symbols as row labels if available
row_labels <- resLFC[top_genes, "symbol"]
row_labels[is.na(row_labels)] <- top_genes[is.na(row_labels)]

mat <- assay(vsd)[top_genes, ]
rownames(mat) <- row_labels

p <- pheatmap(mat,
              scale = "row",
              annotation_col = anno_col,
              annotation_colors = anno_colors,
              show_colnames = FALSE,
              fontsize_row = 8,
              clustering_distance_cols = "euclidean",
              clustering_method = "ward.D2",
              main = "Top 30 DE genes — Severe/ICU vs Healthy/Mild")

ggsave("plots/phase8/phase8_heatmap_top30.png", p, width = 10, height = 8, dpi = 300)
```

---

## 15. Box/jitter plots for key genes

```{r key-gene-plots, fig.width=14, fig.height=5}
# Key genes from the monocyte hypothesis
key_genes_symbols <- c("ISG15", "IL1B", "TNF", "IFIT1", "CCL3", "MX1")

# Find matching gene IDs
if (any(grepl("^ENSG", rownames(dds)))) {
  key_gene_ids <- mapIds(org.Hs.eg.db,
                         keys = key_genes_symbols,
                         column = "ENSEMBL",
                         keytype = "SYMBOL",
                         multiVals = "first")
  key_gene_ids <- key_gene_ids[key_gene_ids %in% rownames(dds)]
} else {
  key_gene_ids <- key_genes_symbols[key_genes_symbols %in% rownames(dds)]
  names(key_gene_ids) <- key_gene_ids
}

cat("Key genes found in data:", paste(names(key_gene_ids), collapse = ", "), "\n")

# Plot each
plots <- list()
for (i in seq_along(key_gene_ids)) {
  gene_id <- key_gene_ids[i]
  gene_name <- names(key_gene_ids)[i]

  d <- plotCounts(dds, gene = gene_id, intgroup = "severity_group",
                  returnData = TRUE)

  plots[[i]] <- ggplot(d, aes(x = severity_group, y = count, fill = severity_group)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +
    geom_jitter(width = 0.15, size = 2) +
    scale_y_log10() +
    scale_fill_manual(values = c("Healthy_Mild" = "steelblue",
                                  "Severe_ICU" = "firebrick")) +
    labs(title = gene_name, x = "", y = "Normalized count") +
    theme_classic(base_size = 11) +
    theme(legend.position = "none")
}

p <- patchwork::wrap_plots(plots, ncol = length(plots))
print(p)
ggsave("plots/phase8/phase8_key_gene_boxplots.png", p,
       width = 14, height = 5, dpi = 300)
```

---

## 16. Save results

```{r save-results}
# Save full DESeq2 results
write.csv(as.data.frame(resOrdered),
          file = "Data/phase8_SevereICU_vs_HealthyMild_DESeq2.csv")

# Save the DESeqDataSet for downstream phases
saveRDS(dds, file = "Data/phase8_dds.rds")

# Save shrunken results
saveRDS(resLFC, file = "Data/phase8_resLFC.rds")

cat("Saved:\n")
cat("  Data/phase8_SevereICU_vs_HealthyMild_DESeq2.csv\n")
cat("  Data/phase8_dds.rds\n")
cat("  Data/phase8_resLFC.rds\n")
```

---

## Decision Checkpoint 4

> **Why DESeq2 for count data?**
>
> 1. DESeq2 models count data with a negative binomial distribution, which
>    properly handles the mean-variance relationship (overdispersion) in RNA-seq.
> 2. With small sample sizes, DESeq2's shrinkage estimators for dispersion
>    and fold changes are more conservative than limma-voom, reducing false
>    positives.
>
> **Why group Severe + ICU together?**
>
> 1.
> 2.
>
> (Fill in after rendering with actual group sizes.)

---

## Session info

```{r session-info}
sessionInfo()
```
