---
title: "Phase 4 – CD14+ Monocyte Deep-Dive & Core Claim"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

## Goal

This document implements **Phase 4** of the project workflow:

- Annotate samples with severity metadata from GEO/paper
- Subset CD14+ monocytes
- ISG and TNF/IL-1β module scoring
- Co-activation analysis (scatter, thresholds)
- Cell-level and donor-level severity comparisons
- Differential expression in monocytes + volcano plot

---

## 1. Setup

```{r setup, message=FALSE, warning=FALSE}
library(Seurat)
library(tidyverse)
library(patchwork)
library(EnhancedVolcano)
library(ggrepel)

knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6
)

dir.create("plots/phase4", showWarnings = FALSE, recursive = TRUE)
```

---

## 2. Load the annotated Seurat object from Phase 3

```{r load-object}
seurat <- readRDS("Data/phase3_seurat_annotated.rds")

cat("Loaded Phase 3 Seurat object.\n")
cat("  Cells:", ncol(seurat), "\n")
cat("  Genes:", nrow(seurat), "\n")
```

---

## 3. Add sample / severity metadata

Barcode suffixes in GSE149689 encode sample identity (from GEO: Lee et al.,
Sci Immunol 2020). Severity assignment (mild vs severe COVID) is based on the
paper's Table S1.

```{r add-metadata}
# Extract barcode suffix → sample mapping from GEO
suffix_to_sample <- c(
  "1"  = "nCoV_1",  "2"  = "nCoV_2",  "3"  = "Flu_1",
  "4"  = "Flu_2",   "5"  = "Normal_1","6"  = "Flu_3",
  "7"  = "Flu_4",   "8"  = "Flu_5",   "9"  = "nCoV_3",
  "10" = "nCoV_4",  "11" = "nCoV_5",  "12" = "nCoV_6",
  "13" = "Normal_2","14" = "Normal_3","15" = "nCoV_7",
  "16" = "nCoV_8",  "17" = "nCoV_9",  "18" = "nCoV_10",
  "19" = "Normal_4","20" = "nCoV_11"
)

# Condition (broad)
suffix_to_condition <- c(
  "1"  = "COVID", "2"  = "COVID", "3"  = "Flu",
  "4"  = "Flu",   "5"  = "Healthy","6"  = "Flu",
  "7"  = "Flu",   "8"  = "Flu",   "9"  = "COVID",
  "10" = "COVID", "11" = "COVID", "12" = "COVID",
  "13" = "Healthy","14"= "Healthy","15"= "COVID",
  "16" = "COVID", "17" = "COVID", "18" = "COVID",
  "19" = "Healthy","20"= "COVID"
)

# Severity (from Lee et al. 2020, Table S1)
# nCoV 1-7 = severe COVID; nCoV 8-11 = mild COVID
suffix_to_severity <- c(
  "1"  = "Severe",  "2"  = "Severe",  "3"  = "Flu",
  "4"  = "Flu",     "5"  = "Healthy", "6"  = "Flu",
  "7"  = "Flu",     "8"  = "Flu",     "9"  = "Severe",
  "10" = "Severe",  "11" = "Severe",  "12" = "Severe",
  "13" = "Healthy", "14" = "Healthy", "15" = "Severe",
  "16" = "Mild",    "17" = "Mild",    "18" = "Mild",
  "19" = "Healthy", "20" = "Mild"
)

# Assign to Seurat object
barcodes <- colnames(seurat)
suffixes <- sub(".*-", "", barcodes)

seurat$sample    <- unname(suffix_to_sample[suffixes])
seurat$condition <- unname(suffix_to_condition[suffixes])
seurat$severity  <- unname(suffix_to_severity[suffixes])
seurat$donor     <- unname(suffix_to_sample[suffixes])   # donor = sample in this dataset

# Set severity as ordered factor
seurat$severity <- factor(seurat$severity,
                          levels = c("Healthy", "Mild", "Severe", "Flu"))

cat("Metadata summary:\n")
print(table(seurat$severity))
cat("\nDonors per severity group:\n")
print(tapply(seurat$donor, seurat$severity, function(x) length(unique(x))))
```

### 3a. UMAP split by severity

```{r umap-severity, fig.width=16, fig.height=5}
p <- DimPlot(seurat, group.by = "singler_main", split.by = "severity",
        reduction = "umap", label = FALSE, pt.size = 0.1) +
  ggtitle("Cell types split by severity")
print(p)
ggsave("plots/phase4/phase4_umap_split_severity.png", p, width = 16, height = 5, dpi = 300)
```

### 3b. Cell-type composition by severity

```{r composition-barplot, fig.width=10, fig.height=6}
comp <- seurat@meta.data %>%
  dplyr::count(severity, singler_main) %>%
  group_by(severity) %>%
  mutate(pct = n / sum(n) * 100)

p <- ggplot(comp, aes(x = severity, y = pct, fill = singler_main)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Severity", y = "Proportion", fill = "Cell type",
       title = "Cell-type composition by severity group") +
  theme_minimal()
print(p)
ggsave("plots/phase4/phase4_composition_barplot.png", p, width = 10, height = 6, dpi = 300)
```

---

## 4. Subset CD14+ monocytes

```{r subset-monocytes}
# Identify monocyte labels from SingleR
mono_labels <- grep("Mono", unique(seurat$singler_main), value = TRUE, ignore.case = TRUE)
cat("Monocyte labels found:", paste(mono_labels, collapse = ", "), "\n")

# If no "Mono" label, try "Monocyte"
if (length(mono_labels) == 0) {
  mono_labels <- grep("Monocyte", unique(seurat$singler_main), value = TRUE, ignore.case = TRUE)
  cat("Trying 'Monocyte' pattern:", paste(mono_labels, collapse = ", "), "\n")
}

mono <- subset(seurat, singler_main %in% mono_labels)
cat("Monocyte subset:\n")
cat("  Cells:", ncol(mono), "\n")
cat("  Severity distribution:\n")
print(table(mono$severity))
```

---

## 5. ISG module scoring

```{r isg-module}
isg_genes <- c("ISG15", "IFIT1", "MX1", "OAS1", "IFI44L")

# Check which genes are present
cat("ISG genes present:",
    paste(isg_genes[isg_genes %in% rownames(mono)], collapse = ", "), "\n")

mono <- AddModuleScore(mono,
                       features = list(isg_genes),
                       name = "ISG")
# Rename the auto-generated column (ISG1) to ISG_score
mono$ISG_score <- mono$ISG1
```

---

## 6. TNF/IL-1β module scoring

```{r tnf-il1-module}
tnf_il1_genes <- c("TNF", "IL1B", "CCL3", "CCL4", "CXCL8")

cat("TNF/IL-1 genes present:",
    paste(tnf_il1_genes[tnf_il1_genes %in% rownames(mono)], collapse = ", "), "\n")

mono <- AddModuleScore(mono,
                       features = list(tnf_il1_genes),
                       name = "TNF_IL1")
mono$TNF_IL1_score <- mono$TNF_IL11
```

---

## 7. Co-activation scatter plot

```{r co-activation-scatter, fig.width=12, fig.height=8}
df <- mono@meta.data %>%
  select(ISG_score, TNF_IL1_score, severity, donor)

p <- ggplot(df, aes(x = ISG_score, y = TNF_IL1_score, color = severity)) +
  geom_point(size = 0.3, alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_color_manual(values = c("Healthy" = "grey60", "Mild" = "steelblue",
                                "Severe" = "firebrick", "Flu" = "darkorange")) +
  labs(x = "ISG module score", y = "TNF/IL-1β module score",
       title = "ISG vs TNF/IL-1β co-activation in monocytes",
       color = "Severity") +
  theme_minimal() +
  facet_wrap(~ severity, ncol = 2)
print(p)
ggsave("plots/phase4/phase4_coactivation_scatter.png", p, width = 12, height = 8, dpi = 300)
```

---

## 8. Define co-activated state

A cell is "co-activated" if **both** its ISG score and TNF/IL-1β score are
above zero (i.e., above the background expectation from `AddModuleScore`).

```{r define-coactivated}
mono$coactivated <- (mono$ISG_score > 0) & (mono$TNF_IL1_score > 0)

cat("Co-activated cells per severity:\n")
coact_summary <- mono@meta.data %>%
  group_by(severity) %>%
  summarise(
    total      = n(),
    coactivated = sum(coactivated),
    pct        = round(100 * coactivated / total, 1),
    .groups    = "drop"
  )
print(coact_summary)
```

---

## 9. Cell-level severity comparison — Violin plots

```{r cell-violin, fig.width=14, fig.height=5}
p1 <- VlnPlot(mono, features = "ISG_score", group.by = "severity",
               pt.size = 0) + ggtitle("ISG score by severity") + NoLegend()
p2 <- VlnPlot(mono, features = "TNF_IL1_score", group.by = "severity",
               pt.size = 0) + ggtitle("TNF/IL-1β score by severity") + NoLegend()
p <- p1 + p2
print(p)
ggsave("plots/phase4/phase4_violin_scores_severity.png", p, width = 14, height = 5, dpi = 300)
```

---

## 10. Donor-level aggregation

```{r donor-aggregation}
donor_summary <- mono@meta.data %>%
  group_by(donor, severity) %>%
  summarise(
    n_cells          = n(),
    mean_ISG         = mean(ISG_score),
    mean_TNF_IL1     = mean(TNF_IL1_score),
    pct_coactivated  = 100 * mean(coactivated),
    .groups = "drop"
  )

knitr::kable(donor_summary, digits = 3,
             caption = "Donor-level monocyte module-score summary")
```

---

## 11. Donor-level statistical tests

```{r donor-stats}
# Subset to COVID patients only for mild vs severe comparison
covid_donors <- donor_summary %>% filter(severity %in% c("Mild", "Severe"))

# Wilcoxon rank-sum tests
cat("=== Wilcoxon rank-sum: Mild vs Severe COVID ===\n\n")

cat("ISG score (donor-level mean):\n")
w_isg <- wilcox.test(mean_ISG ~ severity, data = covid_donors)
print(w_isg)

cat("\nTNF/IL-1 score (donor-level mean):\n")
w_tnf <- wilcox.test(mean_TNF_IL1 ~ severity, data = covid_donors)
print(w_tnf)

cat("\n% Co-activated (donor-level):\n")
w_coact <- wilcox.test(pct_coactivated ~ severity, data = covid_donors)
print(w_coact)

# Kruskal-Wallis across all groups (Healthy, Mild, Severe, Flu)
cat("\n=== Kruskal-Wallis across all severity groups ===\n")
kw <- kruskal.test(pct_coactivated ~ severity, data = donor_summary)
print(kw)
```

---

## 12. Donor-level boxplots

```{r donor-boxplots, fig.width=14, fig.height=5}
p1 <- ggplot(donor_summary, aes(x = severity, y = mean_ISG, fill = severity)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 2) +
  scale_fill_manual(values = c("Healthy" = "grey70", "Mild" = "steelblue",
                               "Severe" = "firebrick", "Flu" = "darkorange")) +
  labs(y = "Mean ISG score", title = "ISG score per donor") +
  theme_minimal() + NoLegend()

p2 <- ggplot(donor_summary, aes(x = severity, y = mean_TNF_IL1, fill = severity)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 2) +
  scale_fill_manual(values = c("Healthy" = "grey70", "Mild" = "steelblue",
                               "Severe" = "firebrick", "Flu" = "darkorange")) +
  labs(y = "Mean TNF/IL-1β score", title = "TNF/IL-1β score per donor") +
  theme_minimal() + NoLegend()

p3 <- ggplot(donor_summary, aes(x = severity, y = pct_coactivated, fill = severity)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 2) +
  scale_fill_manual(values = c("Healthy" = "grey70", "Mild" = "steelblue",
                               "Severe" = "firebrick", "Flu" = "darkorange")) +
  labs(y = "% co-activated monocytes", title = "Co-activation per donor") +
  theme_minimal() + NoLegend()

p <- p1 + p2 + p3
print(p)
ggsave("plots/phase4/phase4_donor_boxplots.png", p, width = 14, height = 5, dpi = 300)
```

---

## 13. Cell-type-specific differential expression (monocytes)

```{r mono-de}
# Compare severe COVID monocytes vs healthy monocytes
mono$severity_ident <- mono$severity
Idents(mono) <- "severity_ident"

mono_de <- FindMarkers(
  mono,
  ident.1 = "Severe",
  ident.2 = "Healthy",
  min.pct = 0.1,
  logfc.threshold = 0.25
)

cat("DE genes (Severe vs Healthy monocytes):", nrow(mono_de), "\n")
cat("\nTop 20 upregulated:\n")
print(head(mono_de[order(-mono_de$avg_log2FC), ], 20))
cat("\nTop 20 downregulated:\n")
print(head(mono_de[order(mono_de$avg_log2FC), ], 20))
```

---

## 14. Volcano plot

```{r volcano, fig.width=10, fig.height=8}
p <- EnhancedVolcano(
  mono_de,
  lab        = rownames(mono_de),
  x          = "avg_log2FC",
  y          = "p_val_adj",
  pCutoff    = 0.05,
  FCcutoff   = 1.0,
  title      = "Severe vs Healthy COVID — CD14+ Monocytes",
  subtitle   = "Differential expression",
  pointSize  = 1.5,
  labSize    = 3.5
)
print(p)
ggsave("plots/phase4/phase4_volcano_monocyte_DE.png", p, width = 10, height = 8, dpi = 300)
```

---

## 15. Save results

```{r save-results}
# Save monocyte subset with scores
saveRDS(mono, file = "Data/phase4_monocytes.rds")

# Save DE results
write.csv(mono_de, file = "Data/phase4_mono_de_severe_vs_healthy.csv")

# Save donor summary
write.csv(donor_summary, file = "Data/phase4_donor_summary.csv",
          row.names = FALSE)

# Save the full Seurat with severity metadata for downstream phases
saveRDS(seurat, file = "Data/phase4_seurat_with_severity.rds")

cat("Saved:\n")
cat("  Data/phase4_monocytes.rds\n")
cat("  Data/phase4_mono_de_severe_vs_healthy.csv\n")
cat("  Data/phase4_donor_summary.csv\n")
cat("  Data/phase4_seurat_with_severity.rds\n")
```

---

## 16. Predict-Before-You-Run 1

**Prediction:** "I expect ISG + TNF/IL-1β co-activation frequency to be higher
in severe COVID-19 monocytes, because severe disease is characterized by both
interferon dysregulation and monocyte-driven cytokine storm (as described by
Lee et al., Sci Immunol 2020)."

**Outcome:** *(Fill in after running — does the data support the prediction?)*

- % co-activated in Severe: `r coact_summary$pct[coact_summary$severity == "Severe"]`%
- % co-activated in Mild: `r coact_summary$pct[coact_summary$severity == "Mild"]`%
- % co-activated in Healthy: `r coact_summary$pct[coact_summary$severity == "Healthy"]`%
- Wilcoxon p-value (Severe vs Mild, donor-level): `r w_coact$p.value`

---

## 17. Next step

Continue with Phase 5 (Replication on GSE150861):

- Independent validation dataset
- Same QC → normalize → cluster → annotate pipeline
- Apply identical ISG and TNF/IL-1β module scores
- Confirm co-activated monocyte state in independent cohort
