---
title: "Phase 12 – GSEA / Pathway Analysis"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

## Goal

This document implements **Phase 12** — Gene Set Enrichment Analysis (GSEA)
on the bulk RNA-seq differential expression results from Phase 8/9.
We ask: which biological pathways are systematically up- or down-regulated
in Severe/ICU COVID-19 whole blood?

Gene sets tested:

- **Hallmark** (H) — 50 curated hallmark pathways
- **GO:BP** (C5) — Gene Ontology Biological Process
- **KEGG** (C2/CP:KEGG) — KEGG pathway database
- **Reactome** (C2/CP:REACTOME) — Reactome pathway database
- **Immunologic signatures** (C7/IMMUNESIGDB) — immune-specific gene sets

> **Predict-Before-You-Run 2:**
> We expect GSEA to show enrichment of
> `HALLMARK_INTERFERON_ALPHA_RESPONSE` and
> `HALLMARK_TNFA_SIGNALING_VIA_NFKB` in Severe/ICU samples, consistent
> with simultaneous ISG activation and monocyte-driven cytokine storm
> observed in the scRNA-seq analysis (Phase 4).

---

## 1. Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(clusterProfiler)
library(enrichplot)
library(msigdbr)
library(org.Hs.eg.db)
library(patchwork)

knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 12,
  fig.height = 7
)

dir.create("plots/phase12", showWarnings = FALSE, recursive = TRUE)
```

---

## 2. Load Phase 8/9 results

```{r load-results}
# Use the adjusted model results (Phase 9) if available; fall back to simple
if (file.exists("Data/phase9_resLFC_adjusted.rds")) {
  resLFC <- readRDS("Data/phase9_resLFC_adjusted.rds")
  cat("Using Phase 9 adjusted model results.\n")
} else {
  resLFC <- readRDS("Data/phase8_resLFC.rds")
  cat("Using Phase 8 simple model results.\n")
}

cat("Total genes in results:", nrow(resLFC), "\n")
cat("Genes with padj < 0.05:", sum(resLFC$padj < 0.05, na.rm = TRUE), "\n")
```

---

## 3. Add gene symbols and build ranked list

```{r add-symbols}
# Map Ensembl IDs → HGNC symbols
resLFC$symbol <- mapIds(
  org.Hs.eg.db,
  keys     = rownames(resLFC),
  column   = "SYMBOL",
  keytype  = "ENSEMBL",
  multiVals = "first"
)

cat("Genes with symbol mapped:", sum(!is.na(resLFC$symbol)), "\n")
```

```{r ranked-list}
# Build ranked gene list: log2FoldChange, named by symbol, no NAs
ranked_df <- as.data.frame(resLFC) %>%
  rownames_to_column("ensembl") %>%
  filter(!is.na(symbol), !is.na(log2FoldChange)) %>%
  arrange(desc(log2FoldChange))

# Handle duplicate symbols by keeping the highest absolute LFC
ranked_df <- ranked_df %>%
  group_by(symbol) %>%
  slice_max(abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(desc(log2FoldChange))

genes_ordered <- ranked_df %>%
  pull(log2FoldChange, name = symbol)

cat("Ranked gene list length:", length(genes_ordered), "\n")
cat("Top 5 (most upregulated):\n")
print(head(genes_ordered, 5))
cat("Bottom 5 (most downregulated):\n")
print(tail(genes_ordered, 5))
```

---

## 4. Load gene sets from MSigDB

```{r load-genesets}
# Hallmark (H)
hallmark <- msigdbr(species = "Homo sapiens", category = "H") %>%
  dplyr::select(gs_name, gene_symbol)

# GO Biological Process (C5 GO:BP)
gobp <- msigdbr(species = "Homo sapiens", category = "C5",
                subcategory = "GO:BP") %>%
  dplyr::select(gs_name, gene_symbol)

# KEGG (C2 CP:KEGG_LEGACY)
kegg <- msigdbr(species = "Homo sapiens", category = "C2",
                subcategory = "CP:KEGG_LEGACY") %>%
  dplyr::select(gs_name, gene_symbol)

# Reactome (C2 CP:REACTOME)
reactome <- msigdbr(species = "Homo sapiens", category = "C2",
                    subcategory = "CP:REACTOME") %>%
  dplyr::select(gs_name, gene_symbol)

# Immunologic signatures (C7 IMMUNESIGDB)
immune <- msigdbr(species = "Homo sapiens", category = "C7",
                  subcategory = "IMMUNESIGDB") %>%
  dplyr::select(gs_name, gene_symbol)

cat("Gene sets loaded:\n")
cat("  Hallmark:     ", n_distinct(hallmark$gs_name), "sets\n")
cat("  GO:BP:        ", n_distinct(gobp$gs_name), "sets\n")
cat("  KEGG:         ", n_distinct(kegg$gs_name), "sets\n")
cat("  Reactome:     ", n_distinct(reactome$gs_name), "sets\n")
cat("  Immunologic:  ", n_distinct(immune$gs_name), "sets\n")
```

---

## 5. Run GSEA — Hallmark

```{r gsea-hallmark}
set.seed(42)
gsea_hallmark <- GSEA(
  geneList     = genes_ordered,
  TERM2GENE    = hallmark,
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH",
  minGSSize    = 10,
  maxGSSize    = 500,
  verbose      = FALSE
)

cat("Hallmark GSEA — significant sets:", nrow(gsea_hallmark@result), "\n")
print(gsea_hallmark@result %>%
  dplyr::select(ID, NES, pvalue, p.adjust) %>%
  arrange(desc(abs(NES))) %>%
  head(15))
```

### Hallmark dot plot

```{r hallmark-dotplot, fig.width=11, fig.height=7}
if (nrow(gsea_hallmark@result) > 0) {
  p <- dotplot(gsea_hallmark, showCategory = 20, split = ".sign") +
    facet_grid(. ~ .sign) +
    ggtitle("GSEA Hallmark — Severe/ICU vs Healthy/Mild") +
    theme(axis.text.y = element_text(size = 9))
  print(p)
  ggsave("plots/phase12/phase12_hallmark_dotplot.png", p,
         width = 13, height = 8, dpi = 300)
} else {
  cat("No significant Hallmark sets at padj < 0.05.\n")
}
```

### Enrichment plot — Interferon alpha response

```{r gsea-ifn-plot, fig.width=10, fig.height=5}
ifn_id <- "HALLMARK_INTERFERON_ALPHA_RESPONSE"
if (ifn_id %in% gsea_hallmark@result$ID) {
  p <- gseaplot2(gsea_hallmark, geneSetID = ifn_id,
                 title = ifn_id, pvalue_table = TRUE)
  print(p)
  ggsave("plots/phase12/phase12_gsea_IFN_alpha.png", p,
         width = 10, height = 6, dpi = 300)
} else {
  cat(ifn_id, "not significant in this run.\n")
}
```

### Enrichment plot — TNFα signaling via NFκB

```{r gsea-tnf-plot, fig.width=10, fig.height=5}
tnf_id <- "HALLMARK_TNFA_SIGNALING_VIA_NFKB"
if (tnf_id %in% gsea_hallmark@result$ID) {
  p <- gseaplot2(gsea_hallmark, geneSetID = tnf_id,
                 title = tnf_id, pvalue_table = TRUE)
  print(p)
  ggsave("plots/phase12/phase12_gsea_TNF_NFKB.png", p,
         width = 10, height = 6, dpi = 300)
} else {
  cat(tnf_id, "not significant in this run.\n")
}
```

### Enrichment plot — Interferon gamma response

```{r gsea-ifng-plot, fig.width=10, fig.height=5}
ifng_id <- "HALLMARK_INTERFERON_GAMMA_RESPONSE"
if (ifng_id %in% gsea_hallmark@result$ID) {
  p <- gseaplot2(gsea_hallmark, geneSetID = ifng_id,
                 title = ifng_id, pvalue_table = TRUE)
  print(p)
  ggsave("plots/phase12/phase12_gsea_IFNg.png", p,
         width = 10, height = 6, dpi = 300)
} else {
  cat(ifng_id, "not significant in this run.\n")
}
```

---

## 6. Run GSEA — GO:BP

```{r gsea-gobp}
set.seed(42)
gsea_gobp <- GSEA(
  geneList      = genes_ordered,
  TERM2GENE     = gobp,
  pvalueCutoff  = 0.01,
  pAdjustMethod = "BH",
  minGSSize     = 15,
  maxGSSize     = 500,
  verbose       = FALSE
)

cat("GO:BP GSEA — significant sets:", nrow(gsea_gobp@result), "\n")
```

```{r gobp-dotplot, fig.width=12, fig.height=8}
if (nrow(gsea_gobp@result) > 0) {
  p <- dotplot(gsea_gobp, showCategory = 20, split = ".sign") +
    facet_grid(. ~ .sign) +
    ggtitle("GSEA GO:BP — Severe/ICU vs Healthy/Mild") +
    theme(axis.text.y = element_text(size = 8))
  print(p)
  ggsave("plots/phase12/phase12_gobp_dotplot.png", p,
         width = 14, height = 9, dpi = 300)
} else {
  cat("No significant GO:BP sets at padj < 0.01.\n")
}
```

---

## 7. Run GSEA — KEGG

```{r gsea-kegg}
set.seed(42)
gsea_kegg <- GSEA(
  geneList      = genes_ordered,
  TERM2GENE     = kegg,
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  minGSSize     = 10,
  maxGSSize     = 500,
  verbose       = FALSE
)

cat("KEGG GSEA — significant sets:", nrow(gsea_kegg@result), "\n")
if (nrow(gsea_kegg@result) > 0) {
  print(gsea_kegg@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust) %>%
    arrange(desc(abs(NES))) %>%
    head(10))
}
```

```{r kegg-dotplot, fig.width=12, fig.height=7}
if (nrow(gsea_kegg@result) > 0) {
  p <- dotplot(gsea_kegg, showCategory = 15, split = ".sign") +
    facet_grid(. ~ .sign) +
    ggtitle("GSEA KEGG — Severe/ICU vs Healthy/Mild") +
    theme(axis.text.y = element_text(size = 9))
  print(p)
  ggsave("plots/phase12/phase12_kegg_dotplot.png", p,
         width = 13, height = 8, dpi = 300)
} else {
  cat("No significant KEGG sets at padj < 0.05.\n")
}
```

---

## 8. Run GSEA — Reactome

```{r gsea-reactome}
set.seed(42)
gsea_reactome <- GSEA(
  geneList      = genes_ordered,
  TERM2GENE     = reactome,
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  minGSSize     = 10,
  maxGSSize     = 500,
  verbose       = FALSE
)

cat("Reactome GSEA — significant sets:", nrow(gsea_reactome@result), "\n")
if (nrow(gsea_reactome@result) > 0) {
  print(gsea_reactome@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust) %>%
    arrange(desc(abs(NES))) %>%
    head(10))
}
```

```{r reactome-dotplot, fig.width=12, fig.height=7}
if (nrow(gsea_reactome@result) > 0) {
  p <- dotplot(gsea_reactome, showCategory = 15, split = ".sign") +
    facet_grid(. ~ .sign) +
    ggtitle("GSEA Reactome — Severe/ICU vs Healthy/Mild") +
    theme(axis.text.y = element_text(size = 8))
  print(p)
  ggsave("plots/phase12/phase12_reactome_dotplot.png", p,
         width = 14, height = 8, dpi = 300)
} else {
  cat("No significant Reactome sets at padj < 0.05.\n")
}
```

---

## 9. Summary table — top sets across all databases

```{r summary-table}
bind_results <- function(gsea_obj, db_name) {
  if (nrow(gsea_obj@result) == 0) return(NULL)
  gsea_obj@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust) %>%
    mutate(database = db_name)
}

summary_tbl <- bind_rows(
  bind_results(gsea_hallmark, "Hallmark"),
  bind_results(gsea_gobp,     "GO:BP"),
  bind_results(gsea_kegg,     "KEGG"),
  bind_results(gsea_reactome, "Reactome")
) %>%
  arrange(desc(abs(NES))) %>%
  head(20)

print(summary_tbl)

write.csv(summary_tbl,
          "Data/phase12_gsea_summary_top20.csv",
          row.names = FALSE)
cat("Saved top-20 GSEA summary to Data/phase12_gsea_summary_top20.csv\n")
```

---

## 10. Save GSEA objects

```{r save-gsea}
saveRDS(gsea_hallmark, "Data/phase12_gsea_hallmark.rds")
saveRDS(gsea_gobp,     "Data/phase12_gsea_gobp.rds")
saveRDS(gsea_kegg,     "Data/phase12_gsea_kegg.rds")
saveRDS(gsea_reactome, "Data/phase12_gsea_reactome.rds")

cat("Saved GSEA result objects to Data/\n")
```

---

## 11. Biological interpretation

```{r interpretation}
cat("===== KEY FINDINGS =====\n\n")

# Check predictions
ifn_hit  <- ifn_id  %in% gsea_hallmark@result$ID
tnf_hit  <- tnf_id  %in% gsea_hallmark@result$ID

cat("Prediction check:\n")
cat("  IFN-alpha response enriched in Severe/ICU: ", ifn_hit, "\n")
cat("  TNFa/NFkB signaling enriched in Severe/ICU: ", tnf_hit, "\n\n")

cat("These findings connect to the scRNA-seq co-activation result (Phase 4):\n")
cat("  - ISG activation in bulk blood mirrors ISG module in CD14+ monocytes\n")
cat("  - TNF/IL-1 pathway in bulk mirrors the TNF/IL-1 module in monocytes\n")
cat("  - Together they support the dual-activation state as a systemic\n")
cat("    transcriptomic feature of severe COVID-19\n")
```

---

## Session info

```{r session-info}
sessionInfo()
```
