---
title: "Phase 6 – scRNA-seq Visualization Summary"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

## Goal

This document implements **Phase 6** of the project workflow:

Compile all key scRNA-seq visualizations from Phases 3–5 into a single
publication-ready summary document. Every plot is saved to `plots/phase6/`.

---

## 1. Setup

```{r setup, message=FALSE, warning=FALSE}
library(Seurat)
library(tidyverse)
library(patchwork)
library(EnhancedVolcano)
library(ggrepel)
library(RColorBrewer)
library(pheatmap)

knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 12,
  fig.height = 7
)

dir.create("plots/phase6", showWarnings = FALSE, recursive = TRUE)
```

---

## 2. Load Phase 3–5 outputs

```{r load-data}
# ── Phase 3 outputs ──
seurat  <- readRDS("Data/phase3_seurat_annotated.rds")
markers <- read.csv("Data/phase3_all_markers.csv")

cat("Phase 3 Seurat object:", ncol(seurat), "cells,", nrow(seurat), "genes\n")

# ── Phase 4 outputs ──
seurat_sev  <- readRDS("Data/phase4_seurat_with_severity.rds")
mono        <- readRDS("Data/phase4_monocytes.rds")
mono_de     <- read.csv("Data/phase4_mono_de_severe_vs_healthy.csv")
donor_summary <- read.csv("Data/phase4_donor_summary.csv")

cat("Phase 4 monocytes:", ncol(mono), "cells\n")
cat("Phase 4 DE genes:", nrow(mono_de), "\n")

# ── Phase 5 outputs ──
rep_mono <- readRDS("Data/phase5_GSE150861_monocytes.rds")
coact_longitudinal <- read.csv("Data/phase5_coactivation_longitudinal.csv")

cat("Phase 5 replication monocytes:", ncol(rep_mono), "cells\n")
```

---

## 3. Plot 1 — UMAP by cell type (overview of PBMC populations in GSE149689)

```{r umap-celltype, fig.width=10, fig.height=7}
p1 <- DimPlot(seurat_sev, group.by = "singler_main", reduction = "umap",
              label = TRUE, repel = TRUE, pt.size = 0.3) +
  ggtitle("GSE149689 — UMAP coloured by cell type (SingleR)") +
  theme(legend.position = "right")

print(p1)

ggsave("plots/phase6/phase6_01_umap_celltype.png", p1,
       width = 10, height = 7, dpi = 300)
```

---

## 4. Plot 2 — UMAP split by severity

```{r umap-severity, fig.width=16, fig.height=5}
p2 <- DimPlot(seurat_sev, group.by = "singler_main", split.by = "severity",
              reduction = "umap", label = FALSE, pt.size = 0.1) +
  ggtitle("Cell types split by severity group")

print(p2)

ggsave("plots/phase6/phase6_02_umap_split_severity.png", p2,
       width = 16, height = 5, dpi = 300)
```

---

## 5. Plot 3 — Feature plots: ISG genes on UMAP

```{r feature-isg, fig.width=15, fig.height=5}
isg_genes <- c("ISG15", "IFIT1", "MX1")

p3 <- FeaturePlot(seurat_sev, features = isg_genes, reduction = "umap",
                  ncol = 3, pt.size = 0.1, order = TRUE) &
  scale_colour_viridis_c() &
  theme_minimal()

print(p3)

ggsave("plots/phase6/phase6_03_feature_ISG.png", p3,
       width = 15, height = 5, dpi = 300)
```

---

## 6. Plot 4 — Feature plots: TNF/IL-1β genes on UMAP

```{r feature-tnf, fig.width=15, fig.height=5}
tnf_genes <- c("TNF", "IL1B", "CCL3")

p4 <- FeaturePlot(seurat_sev, features = tnf_genes, reduction = "umap",
                  ncol = 3, pt.size = 0.1, order = TRUE) &
  scale_colour_viridis_c() &
  theme_minimal()

print(p4)

ggsave("plots/phase6/phase6_04_feature_TNF_IL1.png", p4,
       width = 15, height = 5, dpi = 300)
```

---

## 7. Plot 5 — Violin plots: module scores by severity (monocytes)

```{r violin-scores, fig.width=12, fig.height=5}
# Violin for ISG score
v_isg <- VlnPlot(mono, features = "ISG_score", group.by = "severity",
                 pt.size = 0) +
  ggtitle("ISG module score — monocytes") +
  xlab("Severity") + ylab("ISG score") +
  NoLegend()

# Violin for TNF/IL-1β score
v_tnf <- VlnPlot(mono, features = "TNF_IL1_score", group.by = "severity",
                 pt.size = 0) +
  ggtitle("TNF/IL-1β module score — monocytes") +
  xlab("Severity") + ylab("TNF/IL-1β score") +
  NoLegend()

p5 <- v_isg | v_tnf

print(p5)

ggsave("plots/phase6/phase6_05_violin_module_scores.png", p5,
       width = 12, height = 5, dpi = 300)
```

---

## 8. Plot 6 — Donor-level boxplot: co-activation % by severity (key figure)

```{r donor-boxplot, fig.width=8, fig.height=6}
p6 <- ggplot(donor_summary, aes(x = severity, y = pct_coactivated,
                                fill = severity)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.15, size = 3, alpha = 0.8) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Co-activated monocyte percentage per donor",
       subtitle = "ISG score > 0 AND TNF/IL-1β score > 0",
       x = "Severity group",
       y = "% co-activated monocytes") +
  theme_classic(base_size = 14) +
  theme(legend.position = "none")

print(p6)

ggsave("plots/phase6/phase6_06_donor_coactivation_boxplot.png", p6,
       width = 8, height = 6, dpi = 300)
```

---

## 9. Plot 7 — Volcano plot: monocyte DE (severe vs healthy)

```{r volcano, fig.width=10, fig.height=8}
# Restore gene names as row names if loaded without row.names
if ("X" %in% colnames(mono_de)) {
  rownames(mono_de) <- mono_de$X
}

p7 <- EnhancedVolcano(mono_de,
                      lab = rownames(mono_de),
                      x   = "avg_log2FC",
                      y   = "p_val_adj",
                      title    = "Monocyte DE: Severe COVID vs Healthy",
                      subtitle = "Wilcoxon rank-sum test",
                      pCutoff  = 0.05,
                      FCcutoff = 0.5,
                      pointSize = 1.5,
                      labSize   = 4,
                      drawConnectors = TRUE,
                      widthConnectors = 0.3,
                      maxoverlapsConnectors = 20)

print(p7)

ggsave("plots/phase6/phase6_07_volcano_monocyte_DE.png", p7,
       width = 10, height = 8, dpi = 300)
```

---

## 10. Plot 8 — Marker heatmap: top 5 markers per cluster

```{r marker-heatmap, fig.width=14, fig.height=10}
top5 <- markers %>%
  group_by(cluster) %>%
  top_n(n = 5, wt = avg_log2FC) %>%
  ungroup()

# Downsample for heatmap rendering
set.seed(42)
cells_sub <- seurat[, sample(ncol(seurat),
                             min(ncol(seurat), 5000))]

p8 <- DoHeatmap(cells_sub, features = top5$gene, group.by = "seurat_clusters",
                size = 3) +
  ggtitle("Top 5 markers per cluster (downsampled)") +
  theme(axis.text.y = element_text(size = 6))

print(p8)

ggsave("plots/phase6/phase6_08_marker_heatmap.png", p8,
       width = 14, height = 10, dpi = 300)
```

---

## 11. Plot 9 — Cell-type composition barplot by severity

```{r composition-barplot, fig.width=10, fig.height=6}
comp <- seurat_sev@meta.data %>%
  dplyr::count(severity, singler_main) %>%
  group_by(severity) %>%
  mutate(pct = n / sum(n) * 100)

p9 <- ggplot(comp, aes(x = severity, y = pct, fill = singler_main)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Cell-type composition by severity",
       x = "Severity group", y = "Proportion",
       fill = "Cell type") +
  theme_classic(base_size = 13)

print(p9)

ggsave("plots/phase6/phase6_09_composition_barplot.png", p9,
       width = 10, height = 6, dpi = 300)
```

---

## 12. Plot 10 — GSE150861 replication: co-activation scatter

```{r replication-scatter, fig.width=10, fig.height=6}
rep_df <- rep_mono@meta.data %>%
  select(ISG_score, TNF_IL1_score, day_label, patient)

p10 <- ggplot(rep_df, aes(x = ISG_score, y = TNF_IL1_score,
                          colour = day_label)) +
  geom_point(size = 0.5, alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  facet_wrap(~ patient, ncol = 2) +
  scale_colour_brewer(palette = "Dark2") +
  labs(title = "GSE150861 replication — monocyte co-activation scatter",
       subtitle = "Upper-right quadrant = co-activated (ISG > 0 AND TNF/IL-1β > 0)",
       x = "ISG module score",
       y = "TNF/IL-1β module score",
       colour = "Time point") +
  theme_bw(base_size = 12)

print(p10)

ggsave("plots/phase6/phase6_10_replication_coactivation.png", p10,
       width = 10, height = 6, dpi = 300)
```

---

## 13. Summary panel — key figures combined

```{r summary-panel, fig.width=20, fig.height=12}
# Combine the four most important figures into one panel
panel <- (p1 | p6) / (p7 | p10) +
  plot_annotation(
    title    = "scRNA-seq Analysis Summary — COVID-19 Monocyte Co-Activation",
    subtitle = "GSE149689 (discovery) + GSE150861 (replication)",
    tag_levels = "A"
  )

print(panel)

ggsave("plots/phase6/phase6_summary_panel.png", panel,
       width = 20, height = 12, dpi = 300)
```

---

## Decision Checkpoint 6

> **Summary of key findings visualized:**
>
> 1.
> 2.
> 3.
>
> (Fill in after rendering.)

---

## Session info

```{r session-info}
sessionInfo()
```
