---
title: "Phase 3 - Clustering & Cell-Type Annotation"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

## Goal

This document implements **Phase 3** of the project workflow:

- find neighbors and clusters using the PCA embedding from Phase 2
- compare clustering resolutions (0.1, 0.5, 1.0) before committing to one
- visualize clusters on UMAP
- annotate cell types using canonical marker genes (manual annotation)
- run automated annotation with SingleR (BlueprintEncodeData reference)
- find cluster marker genes and plot a summary heatmap
- load and attach sample-level severity metadata from GEO
- preview severity composition across annotated cell types

---

## 1. Setup

```{r setup, message=FALSE, warning=FALSE}
library(Seurat)
library(SingleR)
library(celldex)
library(SingleCellExperiment)
library(dittoSeq)
library(tidyverse)
library(patchwork)

knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 8,
  fig.height = 5
)

set.seed(42)
```

---

## 2. Load the processed Seurat object from Phase 2

```{r load-object}
seurat <- readRDS("Data/phase2_seurat_processed.rds")

cat("Loaded processed Seurat object.\n")
cat("  Cells:", ncol(seurat), "\n")
cat("  Genes:", nrow(seurat), "\n")
cat("  Reductions available:", paste(names(seurat@reductions), collapse = ", "), "\n")
```

---

## 3. User-defined parameters

```{r parameters}
# Must match the value used in Phase 2
N_PCS <- 20

# Clustering resolution to use for downstream analysis
# Inspect the resolution comparison below before finalising this value
CHOSEN_RESOLUTION <- 0.5
```

---

## 4. Find neighborhood graph

```{r find-neighbors}
seurat <- FindNeighbors(seurat, dims = 1:N_PCS, verbose = FALSE)
cat("Neighbor graph built using dims 1:", N_PCS, "\n")
```

---

## 5. Resolution comparison

Try three resolutions and record how many clusters each produces.
Use this output — together with the UMAP plots below — to justify `CHOSEN_RESOLUTION`.

```{r resolution-comparison}
resolutions <- c(0.1, 0.5, 1.0)

res_summary <- map_dfr(resolutions, function(r) {
  tmp <- FindClusters(seurat, resolution = r, verbose = FALSE)
  tibble(
    resolution    = r,
    n_clusters    = length(unique(tmp$seurat_clusters)),
    cluster_sizes = paste(sort(table(tmp$seurat_clusters), decreasing = TRUE), collapse = ", ")
  )
})

res_summary
```

```{r resolution-umap, fig.width=14, fig.height=4, fig.cap="UMAP at three resolutions"}
plots <- map(resolutions, function(r) {
  tmp <- FindClusters(seurat, resolution = r, verbose = FALSE)
  DimPlot(tmp, reduction = "umap", label = TRUE, pt.size = 0.3) +
    ggtitle(paste0("Resolution = ", r,
                   "  (", length(unique(tmp$seurat_clusters)), " clusters)")) +
    NoLegend()
})

wrap_plots(plots, nrow = 1)
```

---

## 6. Find clusters with chosen resolution

```{r find-clusters}
seurat <- FindClusters(seurat, resolution = CHOSEN_RESOLUTION, verbose = FALSE)

cat("Clusters found at resolution", CHOSEN_RESOLUTION, ":\n")
table(seurat$seurat_clusters)
```

```{r umap-by-cluster, fig.cap="UMAP coloured by cluster"}
DimPlot(seurat, reduction = "umap", label = TRUE, pt.size = 0.3) +
  ggtitle(paste0("Clusters (resolution = ", CHOSEN_RESOLUTION, ")")) +
  NoLegend()
```

```{r tsne-by-cluster, fig.cap="t-SNE coloured by cluster"}
DimPlot(seurat, reduction = "tsne", label = TRUE, pt.size = 0.3) +
  ggtitle(paste0("Clusters (resolution = ", CHOSEN_RESOLUTION, ")")) +
  NoLegend()
```

```{r umap-tsne-side, fig.width=13, fig.height=5, fig.cap="UMAP vs t-SNE side-by-side"}
p1 <- DimPlot(seurat, reduction = "umap", label = TRUE, pt.size = 0.3) +
  ggtitle("UMAP") + NoLegend()
p2 <- DimPlot(seurat, reduction = "tsne", label = TRUE, pt.size = 0.3) +
  ggtitle("t-SNE") + NoLegend()
p1 + p2
```

---

## 7. Manual marker gene inspection

### 7a. Classical monocytes (CD14+ / LYZ)

```{r markers-monocytes, fig.width=12, fig.height=4}
FeaturePlot(
  seurat, reduction = "umap",
  features = c("CD14", "LYZ", "FCGR3A"),
  ncol = 3, pt.size = 0.2
)
```

### 7b. T cells

```{r markers-tcells, fig.width=16, fig.height=4}
FeaturePlot(
  seurat, reduction = "umap",
  features = c("CD3D", "CD3E", "CD4", "CD8A"),
  ncol = 4, pt.size = 0.2
)
```

### 7c. B cells

```{r markers-bcells, fig.width=8, fig.height=4}
FeaturePlot(
  seurat, reduction = "umap",
  features = c("MS4A1", "CD79A"),
  ncol = 2, pt.size = 0.2
)
```

### 7d. NK cells

```{r markers-nk, fig.width=8, fig.height=4}
FeaturePlot(
  seurat, reduction = "umap",
  features = c("NKG7", "GNLY"),
  ncol = 2, pt.size = 0.2
)
```

### 7e. Plasmacytoid dendritic cells (pDC) and conventional DCs

```{r markers-dc, fig.width=8, fig.height=4}
FeaturePlot(
  seurat, reduction = "umap",
  features = c("CLEC4C", "FCER1A"),
  ncol = 2, pt.size = 0.2
)
```

### 7f. Summary dot plot — canonical PBMC markers

```{r dotplot-markers, fig.width=12, fig.height=5}
canonical_markers <- c(
  "CD14", "LYZ", "FCGR3A",   # Monocytes
  "CD3D", "CD4", "CD8A",      # T cells
  "MS4A1", "CD79A",            # B cells
  "NKG7", "GNLY",              # NK cells
  "CLEC4C", "FCER1A"           # DCs / pDCs
)

DotPlot(seurat, features = canonical_markers) +
  RotatedAxis() +
  ggtitle("Canonical PBMC marker expression per cluster")
```

---

## 8. Automated annotation — SingleR

SingleR compares each cell's transcriptome to a reference dataset and assigns
a cell-type label. We use the Blueprint/ENCODE reference, which covers major
immune cell types including monocytes, T cells, B cells, and NK cells.

```{r singler-reference}
singler_ref <- BlueprintEncodeData()
cat("Reference cell types available:\n")
print(table(singler_ref$label.main))
```

```{r run-singler}
# Extract log-normalised counts as a matrix for SingleR
sce_test <- as.SingleCellExperiment(seurat)

singler_results <- SingleR(
  test   = sce_test,
  ref    = singler_ref,
  labels = singler_ref$label.main
)

cat("SingleR annotation complete.\n")
cat("Label distribution:\n")
print(sort(table(singler_results$labels), decreasing = TRUE))
```

```{r singler-score-heatmap, fig.width=12, fig.height=6, fig.cap="SingleR per-cell score heatmap"}
plotScoreHeatmap(singler_results)
```

```{r singler-delta-distribution, fig.cap="Delta score distribution — cells with low delta are ambiguous"}
plotDeltaDistribution(singler_results, ncol = 4)
```

---

## 9. Add SingleR labels to Seurat and visualise

```{r add-singler-labels}
seurat$singler_label <- singler_results$labels

cat("Cell type counts (SingleR):\n")
print(sort(table(seurat$singler_label), decreasing = TRUE))
```

```{r umap-singler, fig.width=10, fig.height=6, fig.cap="UMAP coloured by SingleR cell-type label"}
DimPlot(seurat, reduction = "umap", group.by = "singler_label",
        label = TRUE, repel = TRUE, pt.size = 0.3) +
  ggtitle("Cell types (SingleR — Blueprint/ENCODE reference)") +
  NoLegend()
```

```{r ditto-umap, fig.width=10, fig.height=6, fig.cap="dittoDimPlot: UMAP coloured by cell type"}
dittoDimPlot(seurat, var = "singler_label", reduction.use = "umap", size = 0.4)
```

### Cluster-to-cell-type mapping

```{r cluster-celltype-table}
cluster_ct_table <- as.data.frame(
  table(Cluster = seurat$seurat_clusters, CellType = seurat$singler_label)
) %>%
  filter(Freq > 0) %>%
  arrange(Cluster, desc(Freq))

cluster_ct_table
```

---

## 10. Find all cluster marker genes

```{r find-all-markers}
# Use RNA assay for FindAllMarkers
DefaultAssay(seurat) <- "RNA"

all_markers <- FindAllMarkers(
  seurat,
  min.pct        = 0.25,
  logfc.threshold = 0.25,
  only.pos        = TRUE,
  verbose         = FALSE
)

cat("Total significant markers found:", nrow(all_markers), "\n")
head(all_markers, 10)
```

```{r top-markers-per-cluster}
top5_markers <- all_markers %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 5) %>%
  ungroup()

cat("Top 5 marker genes per cluster:\n")
print(top5_markers %>% select(cluster, gene, avg_log2FC, p_val_adj))
```

```{r marker-heatmap, fig.width=12, fig.height=10, fig.cap="Heatmap: top 5 marker genes per cluster"}
DoHeatmap(
  seurat,
  features = top5_markers$gene,
  size     = 3
) + NoLegend()
```

```{r save-markers}
write.csv(all_markers, file = "Data/phase3_all_cluster_markers.csv", row.names = FALSE)
cat("Saved all cluster markers to: Data/phase3_all_cluster_markers.csv\n")
```

---

## 11. Load and attach sample severity metadata

GSE149689 contains PBMCs from donors spanning healthy, mild, and severe COVID-19.
Because the data was loaded from a single merged 10X matrix, severity labels must
be joined back from the GEO series metadata.

The chunk below uses `GEOquery` to download the series matrix and parse severity
from the sample characteristics. It then maps each cell barcode to its donor
and severity group.

> **Note:** If the series matrix download fails, the metadata can be constructed
> manually using the supplementary table from the original paper
> (Arunachalam et al. 2020, *Science* 369:1210–1220).

```{r load-geo-metadata}
if (!requireNamespace("GEOquery", quietly = TRUE)) {
  BiocManager::install("GEOquery", ask = FALSE)
}
library(GEOquery)

# Download the GSE149689 series matrix (cached locally after first run)
gse <- getGEO("GSE149689", GSEMatrix = TRUE, getGPL = FALSE)
pheno <- pData(gse[[1]])

# Extract the columns we need
metadata_geo <- pheno %>%
  select(
    geo_accession,
    title,
    starts_with("characteristics_ch1")
  )

cat("GEO metadata columns:\n")
print(colnames(metadata_geo))
cat("\nFirst few rows:\n")
print(head(metadata_geo))
```

```{r parse-severity}
# Parse severity from the characteristics columns.
# Adjust the column name / regex below if the GEO record uses different field names.
metadata_geo <- metadata_geo %>%
  mutate(
    severity = case_when(
      str_detect(title, regex("healthy|control", ignore_case = TRUE)) ~ "Healthy",
      str_detect(title, regex("mild|moderate",   ignore_case = TRUE)) ~ "Mild_Moderate",
      str_detect(title, regex("severe|icu",       ignore_case = TRUE)) ~ "Severe_ICU",
      TRUE ~ "Unknown"
    ),
    donor_id = geo_accession
  ) %>%
  select(geo_accession, donor_id, severity)

cat("Severity distribution in GEO metadata:\n")
print(table(metadata_geo$severity))
```

```{r attach-metadata-to-seurat}
# The barcodes in the merged 10X matrix are expected to carry a sample prefix,
# e.g. "SAMPLE_BARCODE-1". Extract the prefix to link each cell to its donor.
#
# Inspect the first few barcodes to confirm the separator used in this dataset:
cat("First 6 cell barcodes:\n")
print(head(colnames(seurat)))

# --- Adjust sep below if barcodes use a different separator ---
cell_meta <- tibble(barcode = colnames(seurat)) %>%
  mutate(
    # Extract everything before the first underscore as the sample identifier
    sample_prefix = str_extract(barcode, "^[^_]+")
  )

cat("\nUnique sample prefixes found in barcodes:\n")
print(unique(cell_meta$sample_prefix))

# If the prefix matches GEO accession numbers (GSMxxxxxxx), join directly.
# Otherwise you will need to manually create the mapping below.
seurat$donor_id  <- cell_meta$sample_prefix
seurat$severity  <- metadata_geo$severity[
  match(cell_meta$sample_prefix, metadata_geo$geo_accession)
]

cat("\nSeverity distribution in Seurat object:\n")
print(table(seurat$severity, useNA = "ifany"))
```

```{r umap-by-severity, fig.width=10, fig.height=6, fig.cap="UMAP coloured by disease severity"}
DimPlot(
  seurat,
  reduction = "umap",
  group.by  = "severity",
  pt.size   = 0.3,
  cols      = c("Healthy" = "#2166ac", "Mild_Moderate" = "#f4a582",
                "Severe_ICU" = "#d6604d", "Unknown" = "grey80")
) + ggtitle("Disease severity per cell")
```

---

## 12. Cell-type composition by severity (preview for Phase 4)

```{r composition-table}
composition <- as.data.frame(
  table(CellType = seurat$singler_label, Severity = seurat$severity)
) %>%
  group_by(Severity) %>%
  mutate(Proportion = Freq / sum(Freq)) %>%
  ungroup()

composition
```

```{r composition-barplot, fig.width=10, fig.height=5, fig.cap="Cell-type composition per severity group"}
ggplot(composition, aes(x = Severity, y = Proportion, fill = CellType)) +
  geom_col(position = "stack") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "PBMC cell-type composition by severity",
       x = "Severity group", y = "Proportion of cells") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

---

## 13. Save the annotated object

```{r save-object}
saveRDS(seurat, file = "Data/phase3_seurat_annotated.rds")
cat("Saved annotated Seurat object to: Data/phase3_seurat_annotated.rds\n")
```

---

## 14. Decision Checkpoint 3

**Decision 3a — Clustering resolution**

| Resolution | Clusters | Assessment |
|------------|----------|------------|
| 0.1        | (fill in after running) | Under-clusters — merges biologically distinct types |
| 0.5        | (fill in after running) | Balanced — used for downstream analysis |
| 1.0        | (fill in after running) | Over-clusters — splits within an expected cell type |

**What I chose:** `CHOSEN_RESOLUTION = `r CHOSEN_RESOLUTION``

**Why:** *(Replace with your own reasoning after inspecting the resolution UMAP plots.
Typical PBMC datasets at this depth show 8–12 biologically meaningful clusters at
resolution 0.5, consistent with major cell lineages: CD14+ monocytes, CD16+ monocytes,
CD4 T cells, CD8 T cells, NK cells, B cells, pDCs, and dendritic cells.)*

**Alternatives considered:**
- Resolution 0.1: fewer clusters, risks merging CD14+ and CD16+ monocytes which are biologically distinct and critical for this project's core question.
- Resolution 1.0: produces small satellite clusters that do not correspond to established PBMC lineages.
- MAD-based adaptive resolution selection: not implemented here but available via the `scran` package.

---

**Decision 3b — SingleR reference choice**

**What I used:** `BlueprintEncodeData()` (Blueprint/ENCODE bulk RNA reference)

**Why:** Covers all major PBMC lineages; widely used in published COVID-19 PBMC studies; computationally faster than fine-grained references.

**Alternatives:**
- `HumanPrimaryCellAtlasData()` — broader cell type coverage, may label monocytes more granularly.
- `MonacoImmuneData()` — specifically designed for blood immune cells with finer T-cell and monocyte subtypes; may be preferable for this project given the monocyte-focused analysis in Phase 4.
- Manual-only annotation: more labour-intensive but gives full control.

---

**What you still need to add:**
- Fill in the cluster count table above with actual values from your run.
- Replace the resolution justification with interpretation of your specific plots.
- Confirm the barcode prefix separator is correct for this dataset and that severity labels are assigned as expected.
- Add your name to the author field.
