---
title: "Phase 2 - Normalization & Dimensionality Reduction"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

## Goal

This document implements **Phase 2** of the project workflow:

- normalize with SCTransform (regressing out mitochondrial percentage)
- run PCA and inspect loadings / heatmaps
- choose the number of PCs via elbow plot
- run UMAP and t-SNE embeddings

---

## 1. Setup

```{r setup, message=FALSE, warning=FALSE}
library(Seurat)
library(tidyverse)
library(patchwork)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5
)

# Create plots output folder
dir.create(PLOTS_DIR, showWarnings = FALSE, recursive = TRUE)
```

---

## 2. Load the filtered Seurat object from Phase 1

```{r load-object}
seurat <- readRDS("Data/phase1_seurat_filtered.rds")

cat("Loaded filtered Seurat object.\n")
cat("  Cells:", ncol(seurat), "\n")
cat("  Genes:", nrow(seurat), "\n")
```

---

## 3. User-defined parameters

```{r parameters}
# Number of PCs to use for UMAP / t-SNE (adjust after inspecting elbow plot)
N_PCS <- 20

# Folder where all Phase 2 plots will be saved
PLOTS_DIR <- "plots/phase2"
```

---

## 4. Normalization

We use the standard LogNormalize workflow (NormalizeData → FindVariableFeatures →
ScaleData) because **both SCTransform v1 and v2** fail with an out-of-memory error
at the `sctransform::correct()` step, which allocates a dense cells × genes matrix
(~9.1 GB) — this is inherent to SCTransform's architecture and not resolved by
the `glmGamPoi` backend or `vst.flavor = "v2"`.

The LogNormalize workflow is well-validated, widely used in published scRNA-seq
studies, and produces comparable downstream results for clustering and
differential expression.

```{r normalize}
seurat <- NormalizeData(seurat, normalization.method = "LogNormalize", scale.factor = 10000)
seurat <- FindVariableFeatures(seurat, selection.method = "vst", nfeatures = 3000)
seurat <- ScaleData(seurat, vars.to.regress = "percent.mt")
```

---

## 5. Run PCA

```{r run-pca}
seurat <- RunPCA(seurat, verbose = FALSE)
```

### 5a. Inspect PC loadings

```{r pc-loadings, fig.width=10, fig.height=8}
p_loadings <- VizDimLoadings(seurat, dims = 1:3, reduction = "pca")
p_loadings
ggsave(file.path(PLOTS_DIR, "phase2_step05a_pc_loadings.png"),
       plot = p_loadings, width = 10, height = 8, dpi = 150)
```

### 5b. PC heatmaps

```{r pc-heatmaps, fig.width=10, fig.height=16}
png(file.path(PLOTS_DIR, "phase2_step05b_pc_heatmaps.png"), width = 1000, height = 1600)
DimHeatmap(seurat, dims = 1:20, cells = 500, balanced = TRUE)
dev.off()
DimHeatmap(seurat, dims = 1:20, cells = 500, balanced = TRUE)
```

---

## 6. Choose number of PCs — Elbow plot

```{r elbow-plot, fig.cap="Elbow plot — choose PCs where the curve flattens"}
p_elbow <- ElbowPlot(seurat, ndims = 30) +
  geom_vline(xintercept = N_PCS, linetype = "dashed", color = "red") +
  annotate("text", x = N_PCS + 1, y = max(Stdev(seurat[["pca"]])) * 0.9,
           label = paste("N_PCS =", N_PCS), hjust = 0, color = "red")
p_elbow
ggsave(file.path(PLOTS_DIR, "phase2_step06_elbow_plot.png"),
       plot = p_elbow, width = 8, height = 5, dpi = 150)
```

---

## 7. Run UMAP

```{r run-umap}
seurat <- RunUMAP(seurat, dims = 1:N_PCS, verbose = FALSE)
```

```{r plot-umap, fig.cap="UMAP embedding"}
p_umap <- DimPlot(seurat, reduction = "umap") + ggtitle("UMAP")
p_umap
ggsave(file.path(PLOTS_DIR, "phase2_step07_umap.png"),
       plot = p_umap, width = 8, height = 6, dpi = 150)
```

---

## 8. Run t-SNE

```{r run-tsne}
seurat <- RunTSNE(seurat, dims = 1:N_PCS, verbose = FALSE)
```

```{r plot-tsne, fig.cap="t-SNE embedding"}
p_tsne <- DimPlot(seurat, reduction = "tsne") + ggtitle("t-SNE")
p_tsne
ggsave(file.path(PLOTS_DIR, "phase2_step08_tsne.png"),
       plot = p_tsne, width = 8, height = 6, dpi = 150)
```

---

## 9. Side-by-side comparison

```{r umap-vs-tsne, fig.width=14, fig.height=5}
p1 <- DimPlot(seurat, reduction = "umap") + ggtitle("UMAP")
p2 <- DimPlot(seurat, reduction = "tsne") + ggtitle("t-SNE")
p_compare <- p1 + p2
p_compare
ggsave(file.path(PLOTS_DIR, "phase2_step09_umap_vs_tsne.png"),
       plot = p_compare, width = 14, height = 5, dpi = 150)
```

---

## 10. Save the processed object

```{r save-object}
saveRDS(seurat, file = "Data/phase2_seurat_processed.rds")
cat("Saved processed Seurat object to: Data/phase2_seurat_processed.rds\n")
```

---

## 11. Decision Checkpoint 2

**Decision 1 — Normalization method:** LogNormalize

- **What I chose:** NormalizeData (LogNormalize, scale.factor = 10,000) → FindVariableFeatures (VST, 3,000 genes) → ScaleData (regress `percent.mt`)
- **Why not SCTransform:** Both SCTransform v1 and v2 (`vst.flavor = "v2"` with `glmGamPoi`) crash with an OOM error at `sctransform::correct()`, which attempts to allocate a ~9.1 GB dense matrix. This is an architectural limitation for datasets of this size. LogNormalize is well-established in the literature and produces concordant results for clustering and differential expression.
- **TODO:** Run SCTransform on an SSH server with more RAM for the final version.

---

**Decision 2 — Number of PCs:** `r N_PCS`

- **What I chose:** 20 PCs
- **Justification from the elbow plot:** PC1 dominates with std ≈ 14 (the monocyte signal). The curve declines gradually through PC2–10 (std 8 → 2.5), then slows to a nearly flat plateau from PC10–20. After PC20 the curve is essentially flat at std ≈ 1–1.5 — the noise floor. There is no single sharp elbow; the transition is gradual, which is typical for complex multi-celltype datasets. PC20 is the start of the plateau.
- **Justification from PC heatmaps:** PC1–3 show extremely clean two-block patterns. PC4–15 retain visible structure. PC16–20 are patchier but not pure noise. Going beyond 20 would add PCs that are predominantly noise.
- **Justification from PC loadings:** PC1 = monocyte markers (FCN1, LYZ, S100A9), PC2 = platelet markers (CLEC1B, TUBB1), PC3 = B cells vs NK/T cells (CD79A vs NKG7, CCL5). All three PCs are biologically interpretable, confirming PCA captured real structure.
- **Alternatives considered:** 15 PCs would likely lose sub-population structure visible in PC16–20; 25 PCs would add noise. N_PCS = 20 is the best-supported cutoff given the evidence from all three diagnostic plots.

---

## 12. Next step

Continue with Phase 3:

- `FindNeighbors()` and `FindClusters()`
- Cell-type annotation with manual markers and SingleR
- `FindAllMarkers()` and marker heatmap
