---
title: "Phase 1 - scRNA-seq QC and Preprocessing"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

## Goal

This document implements **Phase 1** of the project workflow using the same logic as **Tutorial 10**:

- load 10X scRNA-seq data from `/Data/`
- inspect matrix dimensions and sparsity
- create a Seurat object
- calculate mitochondrial percentage
- visualize QC metrics
- choose filtering thresholds
- filter low-quality cells

> This file is written so you can use it directly in your project and later add your own explanations/decision checkpoint text.

---

## 1. Setup

```{r setup, message=FALSE, warning=FALSE}
library(Seurat)
library(Matrix)
library(tidyverse)
library(patchwork)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

---

## 2. User-defined parameters

Edit these values after inspecting the QC plots.

```{r parameters}
# Root folder containing the dataset
DATA_ROOT <- "Data/"

# Seurat project name
PROJECT_NAME <- "COVID_PBMC"

# QC thresholds (initial defaults; adjust after inspection)
MIN_FEATURES <- 500
MAX_FEATURES <- 6000
MAX_PERCENT_MT <- 20

# Optional: minimum number of cells expressing a gene to keep the gene at object creation
MIN_CELLS_PER_GENE <- 3
```

---

## 3. Locate and load the 10X data

This chunk tries to automatically find a valid 10X directory inside `/Data/`.
A valid directory contains files such as:

- `matrix.mtx` or `matrix.mtx.gz`
- `barcodes.tsv` or `barcodes.tsv.gz`
- `features.tsv` / `genes.tsv` (or gzipped versions)

```{r detect-10x-dir}
TENX_DIR <- "Data/GSE149689"
cat("Using 10X directory:\n", TENX_DIR, "\n")
```

```{r load-data}
counts <- Read10X(data.dir = TENX_DIR)

cat("Preview of the count matrix:\n")
counts[1:min(10, nrow(counts)), 1:min(3, ncol(counts))]
```

---

## 4. Inspect matrix dimensions and sparsity

```{r matrix-summary}
total_entries <- nrow(counts) * ncol(counts)
nonzero_entries <- length(counts@x)
sparsity_pct <- 100 * (1 - nonzero_entries / total_entries)
avg_genes_per_cell <- mean(Matrix::colSums(counts > 0))

cat("Matrix dimensions:\n")
cat("  Genes (rows): ", nrow(counts), "\n", sep = "")
cat("  Cells (columns): ", ncol(counts), "\n", sep = "")
cat("  Total entries: ", format(total_entries, big.mark = ","), "\n", sep = "")
cat("  Non-zero entries: ", format(nonzero_entries, big.mark = ","), "\n", sep = "")
cat("  Sparsity: ", round(sparsity_pct, 2), "%\n", sep = "")
cat("  Average genes per cell: ", round(avg_genes_per_cell, 1), "\n", sep = "")
```

---

## 5. Genes detected per cell

```{r genes-per-cell}
genes_per_cell <- Matrix::colSums(counts > 0)
summary(genes_per_cell)
```

```{r genes-per-cell-plot, fig.cap="Genes detected per cell (sorted, log scale)"}
plot(
  sort(genes_per_cell),
  xlab = "Cells (sorted)",
  ylab = "Genes detected",
  log = "y",
  main = "Genes per Cell Distribution",
  pch = 16,
  cex = 0.35
)

abline(h = MIN_FEATURES, col = "red", lwd = 2, lty = 2)
abline(h = MAX_FEATURES, col = "blue", lwd = 2, lty = 2)
legend(
  "topleft",
  legend = c(
    paste("MIN_FEATURES =", MIN_FEATURES),
    paste("MAX_FEATURES =", MAX_FEATURES)
  ),
  col = c("red", "blue"),
  lty = 2,
  lwd = 2,
  bty = "n"
)
```

---

## 6. Create Seurat object

```{r create-seurat}
seurat <- CreateSeuratObject(
  counts = counts,
  project = PROJECT_NAME,
  min.cells = MIN_CELLS_PER_GENE
)

cat("Seurat object created.\n")
cat("  Cells:", ncol(seurat), "\n")
cat("  Genes:", nrow(seurat), "\n")
```

---

## 7. Add mitochondrial QC metric

```{r percent-mt}
seurat[["percent.mt"]] <- PercentageFeatureSet(seurat, pattern = "^MT-")

summary(seurat$percent.mt)
```

```{r mito-distribution, fig.cap="Mitochondrial percentage per cell (sorted)"}
plot(
  sort(seurat$percent.mt),
  xlab = "Cells (sorted)",
  ylab = "% mitochondrial reads",
  main = "Mitochondrial Content Distribution",
  pch = 16,
  cex = 0.35
)

abline(h = MAX_PERCENT_MT, col = "darkgreen", lwd = 2, lty = 2)
legend(
  "topleft",
  legend = paste("MAX_PERCENT_MT =", MAX_PERCENT_MT),
  col = "darkgreen",
  lty = 2,
  lwd = 2,
  bty = "n"
)
```

---

## 8. Visualize QC metrics

```{r violin-qc, fig.width=11, fig.height=4}
VlnPlot(
  seurat,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
  ncol = 3,
  pt.size = 0.1
) + NoLegend()
```

```{r scatter-qc, fig.width=10, fig.height=4}
p1 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "percent.mt") +
  ggtitle("nCount_RNA vs percent.mt")

p2 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") +
  ggtitle("nCount_RNA vs nFeature_RNA")

p1 + p2
```

---

## 9. Summarize filtering rules before applying them

```{r qc-summary-before}
qc_summary_before <- tibble(
  metric = c("cells_total",
             paste0("nFeature_RNA <= ", MIN_FEATURES),
             paste0("nFeature_RNA >= ", MAX_FEATURES),
             paste0("percent.mt >= ", MAX_PERCENT_MT)),
  value = c(
    ncol(seurat),
    sum(seurat$nFeature_RNA <= MIN_FEATURES),
    sum(seurat$nFeature_RNA >= MAX_FEATURES),
    sum(seurat$percent.mt >= MAX_PERCENT_MT)
  )
)

qc_summary_before
```

---

## 10. Filter cells

```{r filter-cells}
seurat_filtered <- subset(
  seurat,
  subset = nFeature_RNA > MIN_FEATURES &
           nFeature_RNA < MAX_FEATURES &
           percent.mt < MAX_PERCENT_MT
)

cat("Cells before filtering:", ncol(seurat), "\n")
cat("Cells after filtering:", ncol(seurat_filtered), "\n")
cat("Cells removed:", ncol(seurat) - ncol(seurat_filtered), "\n")
```

---

## 11. Compare QC metrics before vs after filtering

```{r compare-violin, fig.width=11, fig.height=8}
p_before <- VlnPlot(
  seurat,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
  ncol = 3,
  pt.size = 0.1
) + plot_annotation(title = "Before filtering")

p_after <- VlnPlot(
  seurat_filtered,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
  ncol = 3,
  pt.size = 0.1
) + plot_annotation(title = "After filtering")

p_before / p_after
```

```{r compare-summary-after}
qc_summary_after <- tibble(
  metric = c("cells_remaining",
             "median_nFeature_RNA",
             "median_nCount_RNA",
             "median_percent.mt"),
  value = c(
    ncol(seurat_filtered),
    median(seurat_filtered$nFeature_RNA),
    median(seurat_filtered$nCount_RNA),
    median(seurat_filtered$percent.mt)
  )
)

qc_summary_after
```

---

## 12. Save the filtered object

```{r save-object}
saveRDS(seurat_filtered, file = file.path(DATA_ROOT, "phase1_seurat_filtered.rds"))
cat("Saved filtered Seurat object to:\n", file.path(DATA_ROOT, "phase1_seurat_filtered.rds"), "\n")
```

---

## 13. Decision checkpoint 1

Use this section in your final project write-up.

**Decision:** I filtered cells using thresholds on the number of detected genes and mitochondrial percentage.

**What I chose in this draft:**

- lower gene threshold = `r MIN_FEATURES`
- upper gene threshold = `r MAX_FEATURES`
- mitochondrial threshold = `r MAX_PERCENT_MT`%

**Why this choice:**  
These thresholds remove likely low-quality droplets/dead cells (too few genes), likely doublets or unusually complex droplets (too many genes), and stressed/dying cells (high mitochondrial percentage).

**Alternatives:**  
Possible alternatives include using different fixed thresholds, adaptive thresholds based on median absolute deviation (MAD), explicit doublet-detection tools, or more dataset-specific mitochondrial cutoffs.

**What you still need to add:**  
Replace this generic explanation with your own interpretation based on the actual QC plots from your dataset.

---

## 14. Next step

After Phase 1, continue with:

- `SCTransform()`
- `RunPCA()`
- `ElbowPlot()`
- `RunUMAP()` / `RunTSNE()`
- `FindNeighbors()` and `FindClusters()`

That corresponds to the next part of Tutorial 10 and your project roadmap.
