---
title: "Phase 9 – Adjust for Confounders (Gender)"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

## Goal

This document implements **Phase 9** — testing whether gender confounds the
Severe/ICU vs Healthy/Mild comparison in the GSE152418 bulk RNA-seq data.
We fit an additive DESeq2 model with gender and compare the results to the
simple model from Phase 8.

---

## 1. Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(DESeq2)
library(apeglm)
library(org.Hs.eg.db)
library(patchwork)

knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6
)

dir.create("plots/phase9", showWarnings = FALSE, recursive = TRUE)
```

---

## 2. Load Phase 7–8 outputs

```{r load-data}
counts   <- readRDS("Data/phase7_bulk_counts_filtered.rds")
metadata <- readRDS("Data/phase7_bulk_metadata.rds")
dds_simple <- readRDS("Data/phase8_dds.rds")
resLFC     <- readRDS("Data/phase8_resLFC.rds")

cat("Count matrix:", nrow(counts), "genes x", ncol(counts), "samples\n")
cat("Severity groups:\n")
print(table(metadata$severity_group))
```

---

## 3. Check gender balance

```{r gender-balance}
cat("Gender distribution:\n")
print(table(metadata$gender))

cat("\nGender × Severity group:\n")
gender_sev <- table(metadata$gender, metadata$severity_group)
print(gender_sev)

cat("\nChi-squared test for gender × severity independence:\n")
if (min(dim(gender_sev)) >= 2) {
  chi_test <- chisq.test(gender_sev)
  print(chi_test)
  cat("\np-value:", chi_test$p.value, "\n")
  if (chi_test$p.value < 0.05) {
    cat("→ Gender is significantly associated with severity group — confounding possible.\n")
  } else {
    cat("→ No significant association — gender is reasonably balanced.\n")
  }
} else {
  cat("Cannot test — not enough levels.\n")
}
```

### Gender balance plot

```{r gender-plot, fig.width=8, fig.height=5}
p <- ggplot(metadata, aes(x = severity_group, fill = gender)) +
  geom_bar(position = "dodge") +
  labs(title = "Gender balance across severity groups",
       x = "Severity group", y = "Count", fill = "Gender") +
  theme_classic(base_size = 14)

print(p)
ggsave("plots/phase9/phase9_gender_balance.png", p, width = 8, height = 5, dpi = 300)
```

---

## 4. Fit additive model: ~ gender + severity_group

```{r fit-adjusted}
# Ensure gender is a factor
metadata$gender <- factor(metadata$gender)

# Ensure severity_group reference level
metadata$severity_group <- relevel(factor(metadata$severity_group),
                                    ref = "Healthy_Mild")

dds2 <- DESeqDataSetFromMatrix(
  countData = counts,
  colData   = metadata,
  design    = ~ gender + severity_group
)

dds2 <- DESeq(dds2)

cat("Adjusted model result names:\n")
print(resultsNames(dds2))
```

---

## 5. Shrink fold changes (adjusted model)

```{r shrink-adjusted}
coef_name <- grep("Severe", resultsNames(dds2), value = TRUE)
cat("Shrinking coefficient:", coef_name, "\n")

res2LFC <- lfcShrink(dds2, coef = coef_name, type = "apeglm")

cat("\nAdjusted model results summary:\n")
summary(res2LFC)
```

---

## 6. Compare simple vs adjusted models

### 6a. Fold-change correlation

```{r compare-lfc, fig.width=8, fig.height=8}
# Align genes
shared_genes <- intersect(rownames(resLFC), rownames(res2LFC))
lfc_simple   <- resLFC[shared_genes, "log2FoldChange"]
lfc_adjusted <- res2LFC[shared_genes, "log2FoldChange"]

lfc_cor <- cor(lfc_simple, lfc_adjusted, use = "complete.obs")
cat("Pearson correlation of log2FC (simple vs adjusted):", round(lfc_cor, 4), "\n")

df_compare <- data.frame(
  simple   = lfc_simple,
  adjusted = lfc_adjusted
)

p <- ggplot(df_compare, aes(x = simple, y = adjusted)) +
  geom_point(size = 0.5, alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = paste0("Simple vs Gender-Adjusted log2FC (r = ",
                      round(lfc_cor, 4), ")"),
       x = "log2FC — simple model (~ severity_group)",
       y = "log2FC — adjusted model (~ gender + severity_group)") +
  theme_classic(base_size = 13) +
  coord_equal()

print(p)
ggsave("plots/phase9/phase9_lfc_correlation.png", p, width = 8, height = 8, dpi = 300)
```

### 6b. Significance comparison

```{r compare-sig}
sig_simple   <- sum(resLFC$padj < 0.05, na.rm = TRUE)
sig_adjusted <- sum(res2LFC$padj < 0.05, na.rm = TRUE)

cat("Significant genes (padj < 0.05):\n")
cat("  Simple model:   ", sig_simple, "\n")
cat("  Adjusted model: ", sig_adjusted, "\n")
cat("  Difference:     ", sig_adjusted - sig_simple, "\n")

# Overlap
sig_genes_simple   <- rownames(resLFC)[which(resLFC$padj < 0.05)]
sig_genes_adjusted <- rownames(res2LFC)[which(res2LFC$padj < 0.05)]
overlap <- length(intersect(sig_genes_simple, sig_genes_adjusted))
cat("\n  Overlap:         ", overlap, "\n")
cat("  Only in simple:  ", length(setdiff(sig_genes_simple, sig_genes_adjusted)), "\n")
cat("  Only in adjusted:", length(setdiff(sig_genes_adjusted, sig_genes_simple)), "\n")
```

### 6c. MA plot comparison

```{r ma-comparison, fig.width=12, fig.height=5}
png("plots/phase9/phase9_MA_comparison.png", width = 12, height = 5,
    units = "in", res = 300)
par(mfrow = c(1, 2))
plotMA(resLFC, main = "Simple model (~ severity_group)", ylim = c(-5, 5))
plotMA(res2LFC, main = "Adjusted model (~ gender + severity_group)", ylim = c(-5, 5))
par(mfrow = c(1, 1))
dev.off()

# Also display in document
par(mfrow = c(1, 2))
plotMA(resLFC, main = "Simple model", ylim = c(-5, 5))
plotMA(res2LFC, main = "Adjusted model", ylim = c(-5, 5))
par(mfrow = c(1, 1))
```

---

## 7. Check gender effect specifically

```{r gender-effect}
# Extract gender effect
gender_coef <- grep("gender", resultsNames(dds2), value = TRUE)
cat("Gender coefficient:", gender_coef, "\n")

res_gender <- results(dds2, name = gender_coef)
cat("\nGender effect summary:\n")
summary(res_gender)

sig_gender <- sum(res_gender$padj < 0.05, na.rm = TRUE)
cat("\nGenes significantly associated with gender (padj < 0.05):", sig_gender, "\n")
```

---

## 8. Decision: simple or adjusted model?

```{r decision}
cat("===== MODEL COMPARISON SUMMARY =====\n\n")
cat("Correlation of fold changes: r =", round(lfc_cor, 4), "\n")
cat("Significant genes — simple:  ", sig_simple, "\n")
cat("Significant genes — adjusted:", sig_adjusted, "\n")
cat("Overlap:                     ", overlap, "\n")
cat("Gender-associated genes:     ", sig_gender, "\n\n")

if (lfc_cor > 0.99 && abs(sig_adjusted - sig_simple) < sig_simple * 0.1) {
  cat("RECOMMENDATION: Use the SIMPLE model.\n")
  cat("Rationale: Fold changes are nearly identical (r > 0.99).\n")
  cat("Adding gender as a covariate does not materially change results\n")
  cat("and may cost statistical power with small sample sizes.\n")
} else if (lfc_cor > 0.95) {
  cat("RECOMMENDATION: Models are similar. Either is defensible.\n")
  cat("Consider sample size when deciding whether the extra covariate is justified.\n")
} else {
  cat("RECOMMENDATION: Use the ADJUSTED model.\n")
  cat("Rationale: Gender has a meaningful confounding effect on fold changes.\n")
}
```

---

## 9. Save results

```{r save-results}
saveRDS(dds2, file = "Data/phase9_dds_adjusted.rds")
saveRDS(res2LFC, file = "Data/phase9_resLFC_adjusted.rds")

cat("Saved:\n")
cat("  Data/phase9_dds_adjusted.rds\n")
cat("  Data/phase9_resLFC_adjusted.rds\n")
```

---

## Decision Checkpoint 5

> **Simple vs adjusted model:**
>
> 1. Correlation between fold changes: r = ___
> 2. Gender confounding assessment: ___
> 3. Final decision: ___
>
> (Fill in after rendering.)

---

## Session info

```{r session-info}
sessionInfo()
```
