---
title: "Phase 7 – GSE152418 Bulk RNA-seq Data Loading & EDA"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

## Goal

This document implements **Phase 7** — loading and exploring the bulk RNA-seq
dataset GSE152418 (Arunachalam et al., Science 2020). This dataset contains
whole-blood RNA-seq from COVID-19 patients of varying severity plus healthy
controls. We will:

1. Load the raw count matrix and sample metadata
2. Clean and define severity groups
3. Inspect group sizes, gender balance, and missing data
4. Filter low-count genes
5. Align and verify count matrix with metadata

---

## 1. Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(DESeq2)
library(pheatmap)
library(GEOquery)
library(org.Hs.eg.db)

knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6
)

dir.create("plots/phase7", showWarnings = FALSE, recursive = TRUE)
```

---

## 2. Load the raw count matrix

```{r load-counts}
counts_raw <- read.table("Data/GSE152418/GSE152418_p20047_Study1_RawCounts.txt.gz",
                         header = TRUE, sep = "\t", row.names = 1,
                         check.names = FALSE)

cat("Raw count matrix dimensions:\n")
cat("  Genes:", nrow(counts_raw), "\n")
cat("  Samples:", ncol(counts_raw), "\n")
cat("\nFirst few gene names:\n")
head(rownames(counts_raw))
cat("\nFirst few sample names:\n")
head(colnames(counts_raw))
cat("\nCorner of the matrix:\n")
print(counts_raw[1:5, 1:5])
```

---

## 3. Load metadata from GEO series matrix

```{r load-metadata}
# Try loading from local series matrix file
geo <- getGEO(filename = "Data/GSE152418/GSE152418_series_matrix.txt.gz",
              getGPL = FALSE)

metadata <- pData(geo)

cat("Metadata dimensions:", nrow(metadata), "samples x", ncol(metadata), "columns\n")
cat("\nColumn names:\n")
print(colnames(metadata))
```

---

## 4. Clean and inspect metadata

```{r clean-metadata}
# Inspect key columns
cat("Available characteristics columns:\n")
char_cols <- grep("characteristics", colnames(metadata), value = TRUE)
print(char_cols)

# Print first few values of each characteristics column
for (col in char_cols) {
  cat("\n", col, ":\n")
  print(head(unique(metadata[[col]]), 20))
}

# Extract relevant fields from characteristics columns
# (Adjust column names based on what the series matrix contains)
metadata$title_clean <- metadata$title

# Try to extract disease status and severity
# These are typically in characteristics_ch1, characteristics_ch1.1, etc.
cat("\nSample titles:\n")
print(metadata$title)
```

---

## 5. Parse severity and demographic information

```{r parse-severity}
# Extract fields from characteristics columns
# GEO typically stores: disease state, severity, gender, age, etc.

# Parse each characteristics column
for (col in char_cols) {
  vals <- as.character(metadata[[col]])
  # Extract field name and value
  if (any(grepl(":", vals))) {
    field_name <- sub(":.*", "", vals[1])
    field_name <- trimws(tolower(gsub(" ", "_", field_name)))
    metadata[[field_name]] <- trimws(sub(".*:", "", vals))
    cat("Parsed:", field_name, "→", length(unique(metadata[[field_name]])), "unique values\n")
    print(table(metadata[[field_name]]))
    cat("\n")
  }
}
```

---

## 6. Define severity contrast groups

```{r define-groups}
# The severity column was parsed from characteristics_ch1.3
# Values: Convalescent, Moderate, Severe, ICU, Healthy
cat("Severity values parsed from GEO:\n")
print(table(metadata$severity))

# Create binary severity group:
# Severe_ICU = Severe + ICU patients
# Healthy_Mild = Healthy + Moderate patients
# Exclude Convalescent
metadata$severity_group <- case_when(
  metadata$severity %in% c("Healthy")              ~ "Healthy_Mild",
  metadata$severity %in% c("Moderate")             ~ "Healthy_Mild",
  metadata$severity %in% c("Severe", "ICU")        ~ "Severe_ICU",
  metadata$severity %in% c("Convalescent")         ~ NA_character_,
  TRUE                                              ~ NA_character_
)

cat("\nSeverity group distribution:\n")
print(table(metadata$severity_group, useNA = "ifany"))

# Also show original severity within each group
cat("\nOriginal severity within groups:\n")
print(table(metadata$severity, metadata$severity_group, useNA = "ifany"))

# Remove convalescent / unclassified samples
metadata_clean <- metadata %>% filter(!is.na(severity_group))
cat("\nSamples after excluding convalescent:", nrow(metadata_clean), "\n")

# Set as factor with reference level
metadata_clean$severity_group <- relevel(factor(metadata_clean$severity_group),
                                         ref = "Healthy_Mild")
```

---

## 7. Inspect group sizes and gender balance

```{r group-sizes, fig.width=10, fig.height=5}
cat("Group sizes:\n")
print(table(metadata_clean$severity_group))

# Check for gender column
gender_col <- grep("gender|sex", colnames(metadata_clean), value = TRUE, ignore.case = TRUE)
if (length(gender_col) > 0) {
  cat("\nGender distribution:\n")
  print(table(metadata_clean[[gender_col[1]]]))

  cat("\nGender × Severity:\n")
  gender_sev_table <- table(metadata_clean[[gender_col[1]]], metadata_clean$severity_group)
  print(gender_sev_table)

  # Visualise
  p <- ggplot(metadata_clean, aes(x = severity_group, fill = .data[[gender_col[1]]])) +
    geom_bar(position = "dodge") +
    labs(title = "Gender balance across severity groups",
         x = "Severity group", y = "Count", fill = "Gender") +
    theme_classic(base_size = 14)
  print(p)
  ggsave("plots/phase7/phase7_gender_balance.png", p, width = 8, height = 5, dpi = 300)
} else {
  cat("\nNo gender column found in metadata.\n")
}
```

---

## 8. Check missing data

```{r missing-data}
cat("Percentage of non-missing values per metadata column:\n")
completeness <- colMeans(!is.na(metadata_clean)) * 100
print(round(completeness[completeness < 100], 1))
if (all(completeness == 100)) cat("All columns are 100% complete.\n")
```

---

## 9. Severity distribution bar plot

```{r severity-barplot, fig.width=8, fig.height=5}
p <- ggplot(metadata_clean, aes(x = severity_group, fill = severity_group)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
  scale_fill_manual(values = c("Healthy_Mild" = "steelblue",
                                "Severe_ICU" = "firebrick")) +
  labs(title = "Sample distribution by severity group (GSE152418)",
       x = "Severity group", y = "Number of samples") +
  theme_classic(base_size = 14) +
  theme(legend.position = "none")

print(p)
ggsave("plots/phase7/phase7_severity_distribution.png", p, width = 8, height = 5, dpi = 300)
```

---

## 10. Align count matrix with metadata

```{r align-counts}
# Find matching sample identifiers between counts and metadata
cat("Count matrix sample names (first 10):\n")
print(head(colnames(counts_raw), 10))
cat("\nMetadata sample identifiers (first 10):\n")
print(head(metadata_clean$title, 10))
cat("\nMetadata geo_accession (first 10):\n")
print(head(metadata_clean$geo_accession, 10))

# Try to match by title or geo_accession
if (any(colnames(counts_raw) %in% metadata_clean$title)) {
  match_col <- "title"
} else if (any(colnames(counts_raw) %in% metadata_clean$geo_accession)) {
  match_col <- "geo_accession"
} else {
  # Try substring matching
  cat("\nNo direct match found. Attempting partial matching...\n")
  match_col <- NULL
}

if (!is.null(match_col)) {
  # Subset metadata to samples present in count matrix
  shared <- intersect(colnames(counts_raw), metadata_clean[[match_col]])
  cat("\nShared samples between counts and metadata:", length(shared), "\n")

  # Align
  metadata_clean <- metadata_clean[metadata_clean[[match_col]] %in% shared, ]
  rownames(metadata_clean) <- metadata_clean[[match_col]]
  counts_aligned <- counts_raw[, metadata_clean[[match_col]]]

  # Verify
  cat("Alignment check:", all(colnames(counts_aligned) == rownames(metadata_clean)), "\n")
} else {
  cat("Manual alignment needed — inspect sample names above.\n")
  counts_aligned <- counts_raw
}

cat("\nAligned matrix: ", nrow(counts_aligned), "genes x", ncol(counts_aligned), "samples\n")
```

---

## 11. Filter low-count genes

```{r filter-genes}
cat("Genes before filtering:", nrow(counts_aligned), "\n")

# Strategy: keep genes with at least 10 counts in at least 3 samples
keep <- rowSums(counts_aligned >= 10) >= 3
counts_filtered <- counts_aligned[keep, ]

cat("Genes after filtering:", nrow(counts_filtered), "\n")
cat("Genes removed:", sum(!keep), "\n")
cat("Percentage retained:", round(100 * sum(keep) / length(keep), 1), "%\n")
```

---

## 12. Library size overview

```{r library-sizes, fig.width=10, fig.height=5}
lib_sizes <- colSums(counts_filtered)

p <- data.frame(
  sample = names(lib_sizes),
  lib_size = lib_sizes,
  severity = metadata_clean$severity_group
) %>%
  ggplot(aes(x = reorder(sample, lib_size), y = lib_size / 1e6,
             fill = severity)) +
  geom_col() +
  scale_fill_manual(values = c("Healthy_Mild" = "steelblue",
                                "Severe_ICU" = "firebrick")) +
  labs(title = "Library sizes (GSE152418)",
       x = "Sample", y = "Library size (millions of reads)",
       fill = "Severity") +
  theme_classic(base_size = 12) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 7))

print(p)
ggsave("plots/phase7/phase7_library_sizes.png", p, width = 10, height = 5, dpi = 300)
```

---

## 13. Save cleaned data for Phase 8

```{r save-data}
saveRDS(counts_filtered, file = "Data/phase7_bulk_counts_filtered.rds")
saveRDS(metadata_clean, file = "Data/phase7_bulk_metadata.rds")

cat("Saved:\n")
cat("  Data/phase7_bulk_counts_filtered.rds\n")
cat("  Data/phase7_bulk_metadata.rds\n")
```

---

## Decision Checkpoint 3

> **Data quality and group definition:**
>
> 1. Why group Severe + ICU as one class?
> 2. Why exclude Convalescent samples?
> 3. Are the group sizes sufficient for DESeq2?
>
> (Fill in after rendering.)

---

## Session info

```{r session-info}
sessionInfo()
```
