\documentclass[11pt, a4paper]{article}
\usepackage[margin=2.5cm]{geometry}
\usepackage{parskip}
\usepackage{titlesec}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{array}

\titleformat{\section}{\large\bfseries}{}{0em}{}[\titlerule]
\titleformat{\subsection}{\normalsize\bfseries}{}{0em}{}
\titleformat{\subsubsection}{\normalsize\itshape}{}{0em}{}

\title{\textbf{scRNA-seq Project Overview}\\
\large COVID-19 PBMC — Monocyte Dysregulation}
\author{}
\date{}

\begin{document}
\maketitle

\noindent
This document summarizes the analysis pipeline across all five phases.
The dataset is single-cell RNA-seq from peripheral blood mononuclear cells (PBMCs)
of COVID-19 patients (GSE149689, Lee et al.\ 2020).
The central question is whether CD14\textsuperscript{+} monocytes in severe COVID-19
show simultaneous activation of interferon-stimulated genes (ISGs) and
pro-inflammatory genes (TNF/IL-1\(\beta\)), a state called \textit{co-activation}.

% ─────────────────────────────────────────────
\section{Background}
% ─────────────────────────────────────────────

\subsection{The project question in simple words}

The project asks one central question:

\begin{quote}
\textit{In severe COVID-19, do blood monocytes enter a special overactive state
where they are both fighting the virus and causing strong inflammation at the same time?}
\end{quote}

These two programs are:
\begin{itemize}[nosep]
  \item \textbf{Antiviral program} — interferon-stimulated genes (ISGs) that fight viruses.
  \item \textbf{Inflammatory program} — TNF and IL-1\(\beta\) genes that drive aggressive inflammation.
\end{itemize}

If a monocyte has \textit{both programs active at the same time}, it is called \textbf{co-activated}.
The project tests whether this co-activated state is real, repeatable, and linked to disease severity.

Why it matters: if these cells really enter this mixed overactive state, it helps explain
why severe COVID causes such strong immune damage.

The project checks this in two ways:
\begin{itemize}[nosep]
  \item \textbf{Single-cell RNA-seq} — look at cells one by one and see which monocytes have both programs on.
  \item \textbf{Bulk RNA-seq} — check whether the same overall pattern appears in whole-blood samples.
\end{itemize}

% ────────────────────────────────────────────────
\subsection{The research question — unpacked word by word}

The full question is:

\begin{quote}
\small
\textit{``In peripheral blood mononuclear cells (PBMCs), is severe COVID-19 associated with
a classical-monocyte state that shows both Type~I interferon (ISG) activation and
TNF/IL-1 inflammatory activation, reproducible across independent scRNA-seq cohorts
and detectable in bulk PBMC/blood RNA-seq?''}
\end{quote}

\begin{description}[leftmargin=0pt, itemsep=4pt]
  \item[\normalfont\textbf{``In peripheral blood mononuclear cells (PBMCs)''}]
    We are working with \textbf{blood immune cells}.
    PBMCs include monocytes, T cells, B cells, and NK cells.

  \item[\normalfont\textbf{``is severe COVID-19 associated with''}]
    Does severe COVID \textbf{come together with} this pattern?
    Not necessarily causing it — just linked to it.

  \item[\normalfont\textbf{``a classical-monocyte state''}]
    A \textbf{monocyte} is a specific type of immune cell.
    A \textbf{state} means the cell's current behavior or activity profile.

  \item[\normalfont\textbf{``that shows both''}]
    Two things must be on \textbf{at the same time}. That is the key claim.

  \item[\normalfont\textbf{``Type~I interferon (ISG) activation''}]
    The cell's \textbf{antiviral program} is turned on.
    ISGs (interferon-stimulated genes) are genes the cell activates to fight a virus.

  \item[\normalfont\textbf{``and TNF/IL-1 inflammatory activation''}]
    The cell's \textbf{inflammation program} is also turned on.
    TNF and IL-1 are signals that make the immune response more aggressive.

  \item[\normalfont\textbf{``reproducible across independent scRNA-seq cohorts''}]
    The pattern appears in \textbf{more than one dataset}, not just by chance.

  \item[\normalfont\textbf{``and detectable in bulk PBMC/blood RNA-seq''}]
    The same signal is also visible when looking at all cells mixed together,
    not just cell by cell.
\end{description}

% ────────────────────────────────────────────────────────
\subsection{Biology basics — cells, genes, and RNA}

\subsubsection{The cell and its instruction book}

Your body is made of \textbf{cells}. Every cell contains the same \textbf{DNA},
which is like a full instruction book for the body. A \textbf{gene} is one instruction
in that book.

But a cell does \textit{not} use all genes all the time. When it wants to use a gene,
it makes a temporary copy called \textbf{RNA}. So:

\begin{itemize}[nosep]
  \item \textbf{DNA} = full instruction book (permanent)
  \item \textbf{gene} = one instruction in the book
  \item \textbf{RNA} = the working copy of a gene that is being used right now
\end{itemize}

This means RNA tells us \textbf{what the cell is doing right now}.
That is why we measure RNA instead of DNA to understand cell behavior.

\subsubsection{What scRNA-seq measures}

Single-cell RNA sequencing (\textbf{scRNA-seq}) measures, for each individual cell,
which genes are active and how strongly. The result is a large table:

\begin{itemize}[nosep]
  \item rows = genes
  \item columns = cells
  \item each number = how many RNA molecules from that gene were detected in that cell
\end{itemize}

These counts are usually based on \textbf{UMIs} (unique molecular identifiers),
a technology that counts each RNA molecule once to avoid overcounting.

\subsubsection{What is a monocyte?}

A \textbf{monocyte} is a type of immune cell found in blood. Its job is to detect
threats (bacteria, viruses, damaged tissue) and respond to them.
\textbf{Classical monocytes} (CD14\textsuperscript{+}) are the most common subtype
and the ones studied in this project. In severe COVID-19, these cells may become
abnormally overactivated, which is the central hypothesis of the project.

% ─────────────────────────────────────────────────────────
\subsection{QC metrics — nFeature\_RNA, nCount\_RNA, percent.mt}

Not every captured droplet is a real healthy cell. These three numbers help identify
and remove low-quality entries.

\subsubsection{nFeature\_RNA}

\textbf{How many different genes were detected in this cell.}

Example: if a cell has RNA detected from 800 different genes, then \texttt{nFeature\_RNA = 800}.
This measures \textit{variety}.

\subsubsection{nCount\_RNA}

\textbf{How many total RNA molecules were detected in this cell.}

Example: gene A gives 10 counts, gene B gives 4, gene C gives 1 → \texttt{nCount\_RNA = 15}.
This measures \textit{amount}.

\subsubsection{How they are related}

Think of a shopping bag:
\begin{itemize}[nosep]
  \item \texttt{nFeature\_RNA} = how many \textit{different kinds of items} are in the bag
  \item \texttt{nCount\_RNA} = how many \textit{total items} are in the bag
\end{itemize}
A cell with more total RNA tends to also have more detected genes, but not always.
That is why both are kept as separate signals.

\subsubsection{percent.mt}

\textbf{What fraction of the cell's RNA comes from mitochondrial genes.}
Mitochondria are the energy structures inside a cell.
A cell with unusually high mitochondrial RNA has likely been damaged or is dying,
because when a cell breaks, regular RNA leaks out but mitochondrial RNA is retained.

\subsubsection{Why all three are used together}

Each metric catches something different:

\medskip
\noindent
\begin{tabular}{@{}p{3.2cm}p{3.5cm}p{5.5cm}@{}}
\toprule
\textbf{Metric} & \textbf{Suspicious value} & \textbf{Likely problem} \\
\midrule
\texttt{nFeature\_RNA} & very low & empty droplet / almost no cell \\
\texttt{nFeature\_RNA} & very high & doublet (two cells captured together) \\
\texttt{nCount\_RNA} & very low & empty droplet / weak capture \\
\texttt{nCount\_RNA} & very high & doublet \\
\texttt{percent.mt} & very high & damaged or dying cell \\
\bottomrule
\end{tabular}

\medskip
A \textbf{doublet} is when two cells are accidentally captured together, making the
machine think they are one cell. They appear unusually rich in genes and counts
because they contain RNA from two real cells.

Using all three metrics together is like checking multiple vital signs before trusting
a measurement. Phase 1 of this project uses these numbers to clean the dataset so
that all downstream biology is based on real, single, healthy cells.

% ─────────────────────────────────────────────
\section{Phase 1 — QC and Preprocessing}
% ─────────────────────────────────────────────

\subsection{Load data and inspect \small\normalfont\textit{(Steps 3–4)}}
The raw count matrix from the 10X Genomics pipeline is loaded from the
GSE149689 directory (three files: \texttt{matrix.mtx}, \texttt{barcodes.tsv},
\texttt{features.tsv}).
We then report the number of genes, number of cells, and the sparsity of the
matrix to confirm the data loaded correctly and to get a first sense of scale.

\textbf{Results (GSE149689):}
\begin{itemize}[nosep]
  \item Genes: 33,538
  \item Cells (before QC): 85,144
  \item Total matrix entries: 2,855,559,472
  \item Non-zero entries: 127,119,627
  \item \textbf{Sparsity: 95.55\%}
  \item Average genes per cell: 1,493
\end{itemize}

\textbf{What this means:}
\begin{itemize}[nosep]
  \item \textbf{33,538 genes} is the standard coverage for a human genome — this confirms the data is complete and not truncated.
  \item \textbf{85,144 cells} is a large dataset. This number will shrink after QC filtering removes low-quality droplets.
  \item \textbf{95.55\% sparsity} means that in 95.5\% of all gene-cell combinations, no RNA was detected. This is completely normal and expected in scRNA-seq — a cell only actively uses a small fraction of all genes at any given moment, and low-expression genes are often not captured at all. It does \textit{not} mean the data is bad.
  \item \textbf{1,493 average genes per cell} is a reasonable starting point. Cells with far fewer genes (e.g.\ below 500) are likely empty droplets, and cells with far more (e.g.\ above 6,000) may be doublets. This average gives a reference to anchor the QC thresholds chosen in the next steps.
\end{itemize}

\textit{Note: the sparsity calculation required casting matrix dimensions to
\texttt{numeric} before multiplying, because the product (2.86 billion) exceeds
R's maximum integer value and would otherwise produce \texttt{NA}.}

\subsection{Create Seurat object and compute QC metrics \small\normalfont\textit{(Steps 5–8)}}
A Seurat object is created from the count matrix, keeping only genes expressed
in at least 3 cells.
For each cell we compute three quality metrics:
\begin{itemize}[nosep]
  \item \textbf{nFeature\_RNA} — number of distinct genes detected.
  \item \textbf{nCount\_RNA} — total UMI (read) count.
  \item \textbf{percent.mt} — fraction of reads mapping to mitochondrial genes.
\end{itemize}
High mitochondrial content indicates a damaged or dying cell.
These metrics are visualised with violin plots and sorted scatter plots
so we can choose sensible cutoff values.

\subsection{Filter low-quality cells \small\normalfont\textit{(Steps 9–12)}}
Cells are removed if they fall outside the chosen thresholds:
too few genes (likely empty droplets), too many genes or UMIs (likely doublets),
or too high mitochondrial content (likely dying cells).
The filtered Seurat object is saved for Phase 2.

% ─────────────────────────────────────────────
\section{Phase 2 — Normalization and Dimensionality Reduction}
% ─────────────────────────────────────────────

\subsection{Normalize and identify variable features \small\normalfont\textit{(Step 4)}}
Raw counts are log-normalized (divided by total cell counts, scaled to 10{,}000,
then log-transformed) so that cells of different sequencing depths are comparable.
The 3{,}000 most variable genes are selected to focus on biologically informative
signal.
Finally, the data are scaled and the mitochondrial percentage is regressed out
to remove its confounding effect.

\textit{Note: SCTransform was attempted but fails with an out-of-memory error
on this dataset size. LogNormalize is used instead and is well-established
in the literature.}

\subsection{PCA and choosing the number of components \small\normalfont\textit{(Steps 5–6)}}
PCA is run on the variable genes.
We inspect the elbow plot (variance explained per PC) and choose a cutoff
at the point where adding more PCs yields diminishing returns.
This cutoff value (default 20 PCs) is reused in all downstream steps.

\subsection{UMAP and t-SNE embeddings \small\normalfont\textit{(Steps 7–9)}}
UMAP and t-SNE are computed from the chosen PCA dimensions to produce a
2-D layout for visualisation.
Both embeddings are compared side-by-side; UMAP is used as the primary
visualisation going forward.

% ─────────────────────────────────────────────
\section{Phase 3 — Clustering and Cell-Type Annotation}
% ─────────────────────────────────────────────

\subsection{Graph-based clustering \small\normalfont\textit{(Steps 4–7)}}
A shared-nearest-neighbour (SNN) graph is built from the PCA embedding.
Clusters are detected with the Louvain algorithm at five different resolution
values (0.1 to 1.0) and displayed on the UMAP grid to find the resolution
that gives biologically meaningful, non-over-split clusters.
The chosen resolution is then fixed and used for all downstream steps.

\subsection{Cell-type annotation \small\normalfont\textit{(Steps 8–9)}}
Two complementary approaches are used together:
\begin{itemize}[nosep]
  \item \textbf{Manual markers} — canonical marker genes for monocytes
        (\textit{CD14, LYZ, FCGR3A}), T cells (\textit{CD3D, CD4, CD8A}),
        B cells (\textit{MS4A1, CD79A}), NK cells (\textit{NKG7, GNLY}),
        and dendritic cells (\textit{FCER1A, CST3}) are visualised
        as feature plots and a dot plot.
  \item \textbf{Automated (SingleR)} — each cell is labelled by comparing its
        expression profile against the BlueprintEncode reference using
        the SingleR package. The resulting labels are overlaid on the UMAP and
        cross-tabulated against graph-based clusters.
\end{itemize}

\subsection{Find cluster-defining markers \small\normalfont\textit{(Step 10)}}
\texttt{FindAllMarkers()} runs a differential expression test comparing each
cluster against all other cells, returning the genes that best define each
cluster. The top 5 markers per cluster are shown in a heatmap and a table.

% ─────────────────────────────────────────────
\section{Phase 4 — CD14\textsuperscript{+} Monocyte Deep-Dive}
% ─────────────────────────────────────────────

\subsection{Add severity metadata and subset monocytes \small\normalfont\textit{(Steps 3–4)}}
Each cell's barcode suffix encodes its original sample (donor) identity.
This is mapped to condition (COVID / Flu / Healthy) and severity
(Severe / Mild / Healthy / Flu) based on the sample table from Lee et al.\ 2020.
Cells annotated as monocytes by SingleR are then extracted into a separate
object for focused analysis.

\subsection{Module scoring — ISG and TNF/IL-1\(\beta\) \small\normalfont\textit{(Steps 5–6)}}
Two gene-set scores are computed with \texttt{AddModuleScore()}:
\begin{itemize}[nosep]
  \item \textbf{ISG score} — interferon-stimulated genes:
        \textit{ISG15, IFIT1, MX1, OAS1, IFI44L}.
  \item \textbf{TNF/IL-1\(\beta\) score} — pro-inflammatory genes:
        \textit{TNF, IL1B, CCL3, CCL4, CXCL8}.
\end{itemize}
Each score measures how strongly a cell expresses that programme relative to
background.

\subsection{Co-activation analysis and severity comparison \small\normalfont\textit{(Steps 7–12)}}
A cell is labelled \textit{co-activated} if both its ISG score and
TNF/IL-1\(\beta\) score are above zero.
A scatter plot of the two scores (coloured by severity) visualises the
co-activated quadrant.
Violin plots compare the scores across severity groups at the single-cell level.
Scores are then averaged per donor and compared between Mild and Severe COVID
patients with Wilcoxon rank-sum tests; a Kruskal-Wallis test covers all groups.

\subsection{Differential expression and volcano plot \small\normalfont\textit{(Steps 13–14)}}
\texttt{FindMarkers()} identifies genes differentially expressed between
severe COVID monocytes and healthy monocytes.
Results are shown as a volcano plot (log2 fold-change vs.\ adjusted p-value),
highlighting the top upregulated and downregulated genes.

% ─────────────────────────────────────────────
\section{Phase 5 — Replication on GSE150861}
% ─────────────────────────────────────────────

\subsection{Overview \small\normalfont\textit{(Steps 1–3)}}
GSE150861 (Guo et al.\ 2020) contains PBMCs from two severe COVID-19 patients
treated with Tocilizumab (anti-IL-6R), sampled at days 1, 5, and 7.
This phase checks whether the co-activated monocyte state found in Phase 4
appears in a completely independent dataset, and whether it changes over time
under treatment.

\subsection{QC, normalization, and clustering \small\normalfont\textit{(Steps 4–7)}}
Exactly the same pipeline as Phases 1--3 is applied: QC filtering,
LogNormalize, PCA (20 PCs), UMAP, and Louvain clustering at resolution 0.5.
Cells are annotated with the same manual markers and SingleR reference.

\subsection{Module scoring and co-activation replication \small\normalfont\textit{(Steps 8–10)}}
The same ISG and TNF/IL-1\(\beta\) gene lists from Phase 4 are used unchanged.
A scatter plot of the two scores, now faceted by patient and coloured by
time point, shows whether the co-activated quadrant is populated on day 1.

\subsection{Longitudinal dynamics and Tocilizumab effect \small\normalfont\textit{(Steps 11–12)}}
The percentage of co-activated monocytes and the mean scores are computed
per patient per day.
Line plots track how these values change from day 1 to day 7.
Violin plots and Wilcoxon tests compare day 1 against later time points,
probing whether Tocilizumab reduces the TNF/IL-1\(\beta\) component
while the ISG component persists.

\end{document}
